# Variables and Distributions

Time to play in data land! We began our journey with a promise to help you understand and quantify your uncertainty. Now, we shift to the heart and soul of data - the distribution. Every variable has a distribution, and understanding that distribution is key to understanding the data. We will explore the most common distributions and how to use them to make sense of your data. Let's get started.

## Learning Objectives

1.  Understand the concept of a variable and how they may be represented in data.
2.  Learn about the most common distributions and how to use them.
3.  Understand the concept of a probability density function and how to use it.
4.  Learn how to use distributions to make predictions and quantify uncertainty.

## Variables

Everything we wish to understand about people or things comes in the form of a variable. If I want to know why people like chocolate ice cream, I must assign some form of numerical value to "like" such that I can learn from my observations. Why?  Learning via these methods require data.  Each variable is a set of observations. We assign numbers to observations and these observations serve as our data. Each variable is unique. Some variables contain data that represent a singular characteristic of a person or thing; other variables contain data that represent multiple characteristics. These characteristics we often call constructs in psychological science. From this point forward, we shall refer to the underlying characteristic that we measure as the construct.

Let's examine this a bit more closely. Consider two variables height and weight. Height is a variable that contains data representing a singular characteristic of a person - the length of the person's body when standing. Changes in height measures a singular characteristic of a person that can be altered only by skeletal growth/decay. Weight, on the other hand, contains data representing a single characteristic (mass) but determined by many contributing factors (e.g., water, fat, muscle, bone, etc.). Combined, height and weight make a composite variable called Body Mass Index (BMI) - [an often used but derided](https://www.health.harvard.edu/blog/how-useful-is-the-body-mass-index-bmi-201603309339) predictor of health and well-being.


|                    | Singular Construct | Multiple Constructs |
|--------------------|--------------------|---------------------|
| **Singular Cause** | IDEAL (Height)     | Less IDEAL (Weight) |
| **Multiple Cause** | Most Common (g)    | REALLY BAD (BMI)    |

In the table above, we see that the ideal variable is one that contains data representing a singular characteristic of a person and is determined by a singular cause. Similarly, we see BMI labeled as "REALLY BAD."  These are abstractions right now but they become important as we delve into psychological science and the uncertainty around our constructs.

### Types of Variables

We have many types of variables to choose from when we begin our journey into data land. Each type of variable has its own unique characteristics and uses. Let's explore the most common types of variables. Decades ago, psychological scientists were hamstrung by W.W. Stevens levels of measurement. Stevens argued that there were four levels of measurement: nominal, ordinal, interval, and ratio. These levels of measurement were used to determine the permissible analyses based solely on the logic of arithmetic. Times have changed but arithmetic remains the same. The argument, however, that the level of measurement dictates the analysis is simply untrue. We may do whatever analysis we choose. Which ones make sense depends upon a whole host of factors that we address in this book. So, keep reading. Below is a much easier taxonomy for understanding variables.

#### Continuous Variables

Continuous variables are those that can take on any value within a range. These variables are often measured and can be divided into smaller and smaller units. Examples of continuous variables include height, weight, and temperature. We care not only about the integer values (e.g., 5 feet) but also the decimal values as well (e.g., 5.5 feet). Continuous variables are the most commonly assumed type of variable in psychological science. We emphasize assumed here because we see many instances where continuity, that is, the continuous nature of a variable, makes no sense and yet, in those instances, is treated as such. This is a problem we will return to. For now, a continuous variable is a variable containing any value between a range of possible values including negative and positive infinity.

#### Discrete Variables

Discrete variables are the countable cousins of continuous variables. They take on separated, distinct values - usually whole numbers - with nothing in between. The number of children in a family, the number of times you checked your phone today, the number of correct answers on a quiz: all discrete. You can have 2 children or 3 children, but never 2.4 of them. The gap matters. A Likert scale is the classic troublemaker: a 0-to-5 item offers only the integers 0, 1, 2, 3, 4, 5, so it is *technically* discrete, yet we routinely treat it as continuous - a convenience we return to later. Much of the mischief in applied statistics comes from treating a discrete variable as if it were continuous (or the reverse), so always ask yourself: *can the thing between two values actually exist?*

#### Categorical Variables

Categorical variables sort observations into buckets that have **no inherent numerical order** - think of `pet` (dog, cat, mouse, fish), eye color, political party, or ice-cream flavor. Because the values are non-numeric, they must be converted into numbers at some point before we can compute with them; we can *code* them (1 = chocolate, 2 = vanilla) for convenience, but those numbers are labels, not quantities - vanilla is not "twice" chocolate. A categorical variable with exactly two buckets (yes/no, treatment/control, alive/dead) is called **dichotomous** or **binary**, and it turns out to be one of the most useful variables in all of statistics, as we will see when point-biserial correlation and ANOVA arrive.

#### Ordinal Variables

Ordinal variables are categories with a **rank order** but without guaranteed equal spacing between ranks. Think of a race (1st, 2nd, 3rd) or a survey item (strongly disagree → strongly agree). We know 1st beat 2nd, but the ordering does not tell us *by how much* - the gap between 1st and 2nd may be a mile and the gap between 2nd and 3rd a whisker. Ordinal data live in the uneasy space between categorical and continuous, and reasonable people argue about how to analyze them. We will side-step the dogma: what you can do depends on the question, not on a label Stevens handed down.

### Representing Variables Numerically

To compute anything, we must turn observations into numbers. For continuous and discrete variables that is natural - the number *is* the measurement. For categorical variables we assign codes, but we must remember that the software does not know those codes are arbitrary; it will happily average "flavor" and hand you 1.7 if you let it. In R, we protect ourselves by storing categories as **factors**, which keeps their labels attached and their order (if any) explicit:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch01-heights`, `ch01-skewed`, `ch01-shapes`. For example, `book_data("ch01-heights")` in R, Julia, or Python, or `!bookdata name = "ch01-heights".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

flavor <- fct(c("choc", "van", "choc", "straw", "van"),
              levels = c("choc", "van", "straw"))
as.integer(flavor)                 # the codes R uses under the hood
tibble(flavor) |> count(flavor)    # what we actually care about: the counts
```
### SPSS
```
* Categories live as string values with VALUE LABELS attached; the labels
* travel with the data the way an R factor's levels do.
DATA LIST LIST /flavor (A6).
BEGIN DATA
choc
van
choc
straw
van
END DATA.

* AUTORECODE exposes the integer codes sitting under the labels.
AUTORECODE VARIABLES=flavor /INTO flavor_code.
* What we actually care about: the counts.
FREQUENCIES VARIABLES=flavor.
```
### Julia
```julia
using CategoricalArrays, DataFrames
flavor = categorical(["choc", "van", "choc", "straw", "van"],
                     levels = ["choc", "van", "straw"])
levelcode.(flavor)                  # the codes Julia uses under the hood
combine(groupby(DataFrame(flavor = flavor), :flavor), nrow => :n)
```
### Python
```python
import pandas as pd
flavor = pd.Categorical(["choc", "van", "choc", "straw", "van"],
                        categories=["choc", "van", "straw"])
flavor.codes                        # the codes pandas uses under the hood
pd.Series(flavor).value_counts()    # what we actually care about: the counts
```
:::

### Plotting Variables

Before you compute a single statistic, **look at your data**. A picture reveals shape, spread, gaps, and outliers that a lone number will hide from you. The workhorse for a single variable is the histogram.

#### Histograms

A histogram slices the range of a variable into bins and counts how many observations fall in each. The result is a picture of the variable's **distribution** - the pattern of how often each value shows up.

::: {.panel-tabset}
### R
```{r}
set.seed(20)
heights <- tibble(height = rnorm(500, mean = 68, sd = 4))  # adult heights (inches)
ggplot(heights, aes(x = height)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(title = "The distribution of 500 simulated heights",
       x = "height (inches)", y = "count") +
  theme_minimal()
```
### SPSS
```
* Build 500 simulated adult heights, then draw the histogram.
INPUT PROGRAM.
LOOP #i = 1 TO 500.
  COMPUTE height = RV.NORMAL(68, 4).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

GRAPH /HISTOGRAM = height
  /TITLE = 'The distribution of 500 simulated heights'.
```
### Julia
```julia
using Random, Distributions, Plots
Random.seed!(20)
height = rand(Normal(68, 4), 500)          # adult heights (inches)

histogram(height, bins = 30, legend = false,
          title = "The distribution of 500 simulated heights",
          xlabel = "height (inches)", ylabel = "count")
```
### Python
```python
import numpy as np
import matplotlib.pyplot as plt

rng = np.random.default_rng(20)
height = rng.normal(68, 4, 500)            # adult heights (inches)

plt.hist(height, bins=30, color="steelblue", edgecolor="white")
plt.title("The distribution of 500 simulated heights")
plt.xlabel("height (inches)"); plt.ylabel("count")
plt.show()
```
:::

That mound-shaped pile is the thing we spend the rest of the book reasoning about.

## Shapes

Distributions have shapes, and the shape tells a story. Three features do most of the work:

- **Central tendency** - where the pile sits (the subject of the [next chapter](03-CTmeasures.qmd)).
- **Spread** - how wide the pile is (the chapter after that).
- **Symmetry** - whether the pile leans. A distribution with a long right tail is **right- (positively) skewed**; a long left tail is **left- (negatively) skewed**. Income is famously right-skewed: most people cluster low, a few soar high, and the mean gets dragged toward the millionaires.

::: {.panel-tabset}
### R
```{r}
set.seed(21)
skewed <- tibble(value = rexp(1000, rate = 1))   # a right-skewed variable
ggplot(skewed, aes(x = value)) +
  geom_histogram(bins = 40, fill = "darkorange", colour = "white") +
  labs(title = "A right-skewed distribution (long tail to the right)",
       x = "value", y = "count") +
  theme_minimal()
```
### SPSS
```
* A right-skewed variable: exponential with rate 1.
INPUT PROGRAM.
LOOP #i = 1 TO 1000.
  COMPUTE value = RV.EXP(1).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

GRAPH /HISTOGRAM = value
  /TITLE = 'A right-skewed distribution (long tail to the right)'.
```
### Julia
```julia
using Random, Distributions, Plots
Random.seed!(21)
value = rand(Exponential(1), 1000)         # a right-skewed variable

histogram(value, bins = 40, legend = false,
          title = "A right-skewed distribution (long tail to the right)",
          xlabel = "value", ylabel = "count")
```
### Python
```python
import numpy as np
import matplotlib.pyplot as plt

rng = np.random.default_rng(21)
value = rng.exponential(1.0, 1000)         # a right-skewed variable

plt.hist(value, bins=40, color="darkorange", edgecolor="white")
plt.title("A right-skewed distribution (long tail to the right)")
plt.xlabel("value"); plt.ylabel("count")
plt.show()
```
:::

## Distributions

A **distribution** is the full accounting of how often each value (or range of values) occurs. Some distributions are so common - they arise from the same kinds of processes again and again - that we have given them names, formulas, and R functions. Knowing a handful of them is like knowing a handful of chords: most songs are built from them.

### Names for Shapes

- **Normal (Gaussian)** - the bell curve. Arises whenever many small, independent influences add up (height, measurement error). `rnorm()`.
- **Uniform** - every value in a range equally likely. `runif()`.
- **Binomial** - the count of "successes" in a fixed number of yes/no trials (heads in 10 coin flips). `rbinom()`.
- **Poisson** - the count of rare events in a fixed window (typos per page, emails per hour). `rpois()`.

::: {.panel-tabset}
### R
```{r}
set.seed(22)
# draw 1000 values from each of the four named shapes, then stack them
# into one long tibble so ggplot can facet them side by side
tibble(
  Normal   = rnorm(1000),
  Uniform  = runif(1000),
  Binomial = rbinom(1000, 10, .5),
  Poisson  = rpois(1000, 3)
) |>
  pivot_longer(everything(), names_to = "shape", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "grey85", colour = "white") +
  facet_wrap(~ shape, scales = "free") +
  labs(x = NULL, y = "count") +
  theme_minimal()
```
### SPSS
```
* Draw 1000 values from each of the four named shapes.
INPUT PROGRAM.
LOOP #i = 1 TO 1000.
  COMPUTE Normal   = RV.NORMAL(0, 1).
  COMPUTE Uniform  = RV.UNIFORM(0, 1).
  COMPUTE Binomial = RV.BINOM(10, 0.5).
  COMPUTE Poisson  = RV.POISSON(3).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

* SPSS has no faceting, so each shape gets its own GRAPH command.
GRAPH /HISTOGRAM = Normal   /TITLE = 'Normal'.
GRAPH /HISTOGRAM = Uniform  /TITLE = 'Uniform'.
GRAPH /HISTOGRAM = Binomial /TITLE = 'Binomial'.
GRAPH /HISTOGRAM = Poisson  /TITLE = 'Poisson'.
```
### Julia
```julia
using Random, Distributions, Plots
Random.seed!(22)

shapes = ["Normal"   => rand(Normal(0, 1), 1000),
          "Uniform"  => rand(Uniform(0, 1), 1000),
          "Binomial" => rand(Binomial(10, 0.5), 1000),
          "Poisson"  => rand(Poisson(3), 1000)]

# one panel per shape, laid out 2 x 2
plot([histogram(v, bins = 30, title = k, legend = false) for (k, v) in shapes]...,
     layout = (2, 2))
```
### Python
```python
import numpy as np
import matplotlib.pyplot as plt

rng = np.random.default_rng(22)
shapes = {"Normal":   rng.normal(0, 1, 1000),
          "Uniform":  rng.uniform(0, 1, 1000),
          "Binomial": rng.binomial(10, 0.5, 1000),
          "Poisson":  rng.poisson(3, 1000)}

fig, axes = plt.subplots(2, 2)             # one panel per shape
for ax, (name, v) in zip(axes.flat, shapes.items()):
    ax.hist(v, bins=30, color="lightgrey", edgecolor="white")
    ax.set_title(name)
fig.tight_layout(); plt.show()
```
:::

### Probability Density Functions

For a continuous variable, asking "what is the probability of *exactly* 68.0000 inches?" is a trap - the answer is zero, because there are infinitely many possible values. Instead we describe a continuous distribution with a **probability density function (PDF)**: a curve whose *area* between two points gives the probability of landing in that range. The whole area under the curve is exactly 1 (something must happen). The normal PDF is the famous bell:

::: {.panel-tabset}
### R
```{r}
tibble(z = seq(-4, 4, length.out = 400)) |>
  mutate(density = dnorm(z)) |>
  ggplot(aes(x = z, y = density)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(title = "The standard normal PDF", x = "z", y = "density") +
  theme_minimal()
```
### SPSS
```
* The standard normal density across a grid of z values from -4 to 4.
INPUT PROGRAM.
LOOP #i = -400 TO 400.
  COMPUTE z = #i / 100.
  COMPUTE density = PDF.NORMAL(z, 0, 1).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

GRAPH /LINE(SIMPLE) = VALUE(density) BY z
  /TITLE = 'The standard normal PDF'.
```
### Julia
```julia
using Distributions, Plots
z = range(-4, 4, length = 400)

plot(z, pdf.(Normal(0, 1), z), linewidth = 2, legend = false,
     title = "The standard normal PDF", xlabel = "z", ylabel = "density")
vline!([0], linestyle = :dash, color = :grey)
```
### Python
```python
import numpy as np
import matplotlib.pyplot as plt
from scipy.stats import norm

z = np.linspace(-4, 4, 400)

plt.plot(z, norm.pdf(z), color="steelblue", linewidth=2)
plt.axvline(0, linestyle="--", color="grey")
plt.title("The standard normal PDF")
plt.xlabel("z"); plt.ylabel("density")
plt.show()
```
:::

The height of the curve is *density*, not probability; probability is the shaded area you sweep out between two z-values. That single idea, probability as area under a density, is the engine of everything inferential to come, starting with the [z-distribution](06-zdistribution.qmd). But first we need to pin down the two summaries every distribution demands: its center and its spread.
