# Expectation and Understanding: The Z-Distribution

Every distribution we have met so far has its own units - inches, dollars, IQ points, reaction times. That makes them impossible to compare directly. The **z-distribution** is our universal translator: a way to put *any* value, from *any* normal distribution, onto one common ruler. Master this one distribution and the entire inferential half of the book opens up, because every test we run is, underneath, a z-in-disguise.

## Learning Objectives

1.  Understand the z-score as a value's distance from the mean, in standard-deviation units.
2.  Understand the standard normal distribution and read probabilities as areas.
3.  Use R (`pnorm`, `qnorm`) instead of a dusty z-table.
4.  See why this distribution is the foundation for the inferential chapters ahead.

## The Z-Score: One Ruler to Rule Them All

A **z-score** answers a single question: *how many standard deviations is this value from its mean?*

$$z = \frac{x - \bar{x}}{s}$$

That is the whole formula. Subtract the mean (now centered at 0), divide by the standard deviation (now measured in SD-units), and every variable on Earth speaks the same language.

```{r}
sat  <- 1350;  sat_mean  <- 1050; sat_sd  <- 200
act  <- 30;    act_mean  <- 21;   act_sd  <- 5

z_sat <- (sat - sat_mean) / sat_sd
z_act <- (act - act_mean) / act_sd
c(z_SAT = z_sat, z_ACT = z_act)
```

Is a 1350 SAT better than a 30 ACT? On their own scales, who could say. In z-units the question is trivial: the ACT score is further above its mean, so it is the more impressive result. That is the magic of standardizing - **incomparable things become comparable.**

## The Standard Normal Distribution

Standardize a normal variable and you get *the* **standard normal distribution**: a bell curve with mean 0 and standard deviation 1. It never changes, which is exactly why it is useful - we can chart it once and reuse it forever.

```{r}
curve(dnorm(x), from = -4, to = 4, lwd = 2, col = "steelblue",
      main = "The standard normal: mean 0, sd 1", xlab = "z", ylab = "density")
abline(v = c(-2, -1, 0, 1, 2), lty = 3, col = "grey60")
```

The famous **68–95–99.7 rule** lives here: about 68% of values fall between $z=\pm1$, about 95% between $z=\pm2$, and about 99.7% between $z=\pm3$. Anything past $z = \pm2$ is already unusual; past $\pm3$, rare.

## Probability Is Area (and R Is Your Z-Table)

Older books make you hunt through a printed **z-table** to convert a z-score into a probability. We have R, so we will never do that. Recall from the [distributions chapter](02-distributions.qmd) that probability is *area under the density curve*. Two functions give it to you instantly:

- `pnorm(z)` - the area to the **left** of `z` (the cumulative probability).
- `qnorm(p)` - the inverse: the z-score with area `p` to its left (a **quantile**).

```{r}
pnorm(1)             # P(Z < 1): area to the left of z = 1
1 - pnorm(2)         # P(Z > 2): the upper tail beyond z = 2
pnorm(2) - pnorm(-2) # P(-2 < Z < 2): the middle ~95%
qnorm(0.975)         # which z cuts off the top 2.5%? (the famous 1.96)
```

That last line is worth a star: `qnorm(0.975)` returns **1.96**, the number that shows up in every 95% confidence interval and two-tailed test you will ever run. It is not magic - it is just the z-score with 2.5% of the area beyond it in each tail.

```{r}
# picture the "unusual" tails beyond +/- 1.96
z <- seq(-4, 4, length.out = 400)
plot(z, dnorm(z), type = "l", lwd = 2, col = "steelblue",
     main = "The middle 95% and the 5% that would surprise us",
     xlab = "z", ylab = "density")
abline(v = c(-1.96, 1.96), col = "red", lwd = 2, lty = 2)
```

## Percentiles: Where Do You Stand?

Turn a z-score into a **percentile** with `pnorm()` and you can say exactly where an observation stands in its distribution:

```{r}
z_height <- (74 - 68) / 4              # a 74-inch person, mean 68, sd 4
percentile <- pnorm(z_height) * 100
c(z = z_height, percentile = percentile)
```

A z of 1.5 places this person around the 93rd percentile - taller than about 93% of the population. Percentile ranks, growth charts, and standardized-test scores are all `pnorm()` in another form.

## Another Ruler: Percent of Maximum Possible (POMP)

The z-score compares a value to *its own sample* - to the mean and spread of the data you happened to collect. That is exactly right for asking "how unusual is this person?" But it carries a limitation: because z depends on the sample, the same raw score turns into a different z in a different sample, and z-scores from two different measures are not truly on the same footing.

When the goal is to compare *measures* - say, is work engagement higher than well-being in this study? - a different ruler is more honest: **POMP**, the **Percent Of Maximum Possible**. POMP rescales a score to the range the instrument *could* produce, from its lowest possible value to its highest:

$$\text{POMP} = 100 \times \frac{X - \text{min}}{\text{max} - \text{min}}$$

A POMP of 0 is the lowest possible score, 100 the highest, and 50 the exact middle of the scale. Crucially, that meaning does not depend on the sample. Two measures with completely different response formats - a 1–5 scale and a 1–7 scale - both land on a common 0–100 scale you can read and compare at a glance:

::: {.panel-tabset}
### R
```{r}
mood     <- 4    # answered on a 1-5 scale
vitality <- 5    # answered on a 1-7 scale

pomp <- function(x, lo, hi) 100 * (x - lo) / (hi - lo)
c(mood_POMP     = pomp(mood,     lo = 1, hi = 5),
  vitality_POMP = pomp(vitality, lo = 1, hi = 7))
```

### SPSS
```
* POMP: 100 * (score - min) / (max - min), using each scale's theoretical min and max.
COMPUTE mood_POMP     = 100 * (mood - 1) / (5 - 1).
COMPUTE vitality_POMP = 100 * (vitality - 1) / (7 - 1).
EXECUTE.
```

### Julia
```julia
mood_POMP     = 100 .* (df.mood     .- 1) ./ (5 - 1)
vitality_POMP = 100 .* (df.vitality .- 1) ./ (7 - 1)
```
:::

On their raw scales, 4 and 5 are hard to compare. In POMP units, the mood score sits at 75% of its possible range and vitality at about 67%, so this person is, if anything, in a slightly better mood than their vitality would suggest. This is the standardization we now use in most of our own papers (PEM): it lets a reader compare outcomes across measures directly, and it has a useful side effect - if a scale's response format changes partway through a study, the shift in POMP makes the problem hard to miss.

::: {.callout-important}
## z or POMP: two questions, two rulers
Reach for a **z-score** to ask *how unusual is this value within its distribution?* - it locates a person relative to their sample. Reach for **POMP** to ask *how high is this score on what the instrument can measure?* - it places a score relative to the possible range, so different measures become comparable. Neither is better; they answer different questions.
:::

::: {.callout-important}
## Why This Chapter Is the Hinge of the Book

Everything inferential rests on one move: **standardize, then read the area.** When we test a hypothesis, we convert our result to a z-like statistic and ask "how far into the tail is this, if nothing were going on?" When we build a confidence interval, we walk out $\pm 1.96$ standard errors. The z-distribution is not one more topic - it is the **shared machinery** of the entire inferential section that follows.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  IQ has mean 100 and sd 15. Using `pnorm`/`qnorm`, find (a) the percentile of an IQ of 130, (b) the probability of an IQ above 145, and (c) the IQ that marks the top 1%.
2.  Standardize a variable of your own with `scale()` and confirm its mean is 0 and sd is 1 (to rounding).
3.  Recreate the shaded-tails plot, but mark $\pm 1.65$ instead of $\pm 1.96$. What test decision does 1.65 correspond to (hint: revisit the one-tailed critical value)?
:::

## Where We Go Next

We can now describe where a single value stands. Real science, though, asks how two variables move *together*. The tool that measures that shared movement - and the true engine under correlation and regression - is [covariance](07-covariance.qmd).
