# Covariance and Association

We want to emphasize that most of our work relies heavily on covariance structures. Everything from correlation to regression to ANOVA to structural equation models is, at bottom, an argument about how variables **vary together**. So before we build models, we need to understand the raw material they are built from. Variance measured how a single variable spreads around its own mean; **covariance** measures how *two* variables spread around their means *at the same time*.

## Learning Objectives

1.  See covariance as the natural two-variable extension of variance.
2.  Compute and interpret a covariance, and understand its units problem.
3.  Standardize covariance into correlation, and know what correlation can and cannot tell you.
4.  Recognize covariance as the engine beneath the General Linear Model.

## Why Start With Causes?

Science usually begins with a hunch about a **cause**: sleep affects mood, study time affects grades, a drug affects recovery. We cannot see causes directly; we can only watch variables move. If X truly drives Y, then when X is high we expect Y to be high (or low) *systematically* - the two should **co-vary**. Covariance is our first, humblest measurement of that co-movement. It cannot prove the cause (that famous warning is coming), but a cause that leaves *no* covariance behind is a cause with no evidence.

## From Variance to Covariance

Recall variance: the average product of a deviation *with itself*, $\frac{\sum(x-\bar x)(x-\bar x)}{n-1}$. Covariance makes one small change - it multiplies each variable's deviation by *the other's*:

$$\text{cov}(x, y) = \frac{\sum (x - \bar{x})(y - \bar{y})}{n - 1}$$

Think about the sign of that product. When a point is above the mean on *both* variables (or below on both), the two deviations share a sign and their product is **positive**. When it is high on one and low on the other, the product is **negative**. Sum them up: mostly-positive products mean the variables rise together; mostly-negative means one rises as the other falls.

```{r}
set.seed(5)
study  <- rnorm(100, 5, 2)
grade  <- 60 + 4 * study + rnorm(100, 0, 5)    # grade rises with study

# by hand:
cov_hand <- sum((study - mean(study)) * (grade - mean(grade))) / (length(study) - 1)
c(by_hand = cov_hand, r_cov = cov(study, grade))   # match
```

Positive, as expected: more study, more grade. Let's *see* it, colouring each point by the sign of its deviation product:

```{r}
sign_prod <- (study - mean(study)) * (grade - mean(grade))
plot(study, grade, pch = 19, col = ifelse(sign_prod > 0, "steelblue", "red"),
     main = "Blue points push covariance up; red points push it down")
abline(v = mean(study), h = mean(grade), lty = 2, col = "grey50")
```

The blue points (agreeing deviations) dominate, so covariance is positive. That picture *is* covariance.

## The Units Problem, and the Fix: Correlation

Covariance has a serious drawback for interpretation: its size depends on the **units**. Measure grade in points versus proportions and the covariance changes a lot, even though the relationship is identical. This is the same units problem the standard deviation had, and we fix it the same way, by **standardizing**. Divide covariance by both standard deviations and you get the **correlation coefficient**, $r$:

$$r = \frac{\text{cov}(x, y)}{s_x \, s_y}$$

```{r}
r_hand <- cov(study, grade) / (sd(study) * sd(grade))
c(by_hand = r_hand, r_cor = cor(study, grade))     # match
```

Correlation is covariance stripped of its units, squeezed onto a fixed scale from $-1$ (perfect negative) through $0$ (no linear relationship) to $+1$ (perfect positive). Now the number means the same thing whatever the variables - which is why $r$, not raw covariance, is what we report.

## What Correlation Will Not Tell You

Two lasting warnings. First, the familiar one: **correlation is not causation.** Ice-cream sales and drownings correlate (both driven by summer heat), but ice cream does not cause drowning. Second, and less appreciated: $r$ measures only the **linear** part of a relationship. A strong curved relationship can produce $r \approx 0$. This is why we told you to *always plot your data* - the single number can hide the whole story:

```{r}
# a perfect U-shape: strong relationship, ~zero correlation
xx <- seq(-3, 3, length.out = 100)
yy <- xx^2
c(correlation = round(cor(xx, yy), 3))    # ~0, yet Y is completely determined by X!
plot(xx, yy, pch = 19, col = "darkorange",
     main = "r is near zero, but the relationship is perfect (just not linear)")
```

## Anscombe's Quartet: Same Numbers, Four Different Stories

The U-shape above was one warning. Here is the most famous one, and the reason "always plot your data" belongs on the first page of every statistics course. In 1973, Francis Anscombe built four small datasets that share almost every summary statistic you would normally report - the same mean of $x$, the same mean of $y$, the same variances, the same correlation, and the same regression line - yet describe four completely different worlds. R ships them in the built-in `anscombe` data:

```{r}
a <- anscombe
sapply(1:4, function(i) round(cor(a[[paste0("x", i)]], a[[paste0("y", i)]]), 3))
```

Four identical correlations, $r = 0.816$. Every set also has the same fitted line, $\hat{y} = 3 + 0.5\,x$. If you reported only the numbers, you would call these datasets interchangeable. Now plot them:

```{r}
#| fig-cap: "Anscombe's quartet: identical means, variances, correlation, and regression line — but four different stories."
#| fig-height: 6
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (i in 1:4) {
  x <- a[[paste0("x", i)]]; y <- a[[paste0("y", i)]]
  plot(x, y, pch = 19, col = "steelblue", xlim = c(3, 20), ylim = c(2, 14),
       main = paste("Dataset", i), xlab = "x", ylab = "y")
  abline(lm(y ~ x), col = "firebrick", lwd = 2)   # the SAME line every time
}
par(op)
```

The same red line runs through all four, but only the first tells the story the line claims:

- **Dataset 1** is a genuine, if noisy, straight-line relationship. The line fits.
- **Dataset 2** is a smooth *curve*. A straight line is the wrong model here; the correlation is real, but the shape is not linear.
- **Dataset 3** is a tight straight line spoiled by a single **outlier** that drags the fitted line off course.
- **Dataset 4** has no relationship among most of the points (they all share one $x$ value); a single far-off point, on its own, creates the entire correlation. That lone point has enormous **leverage**.

Identical statistics, four different truths. The picture told you in a moment what the correlation coefficient never could. This is why the [graphical display of data](21-graphics.qmd) is not decoration; it is often the fastest way to catch a relationship that is curved, contaminated, or driven by a single point.

::: {.callout-important}
## From the picture to the residuals

Look again at what the scatter plots did: they showed you *where the straight line was wrong*. That is exactly the job of a **residual**. Recall the chain that runs through this book. We began with [dispersion](04-Dmeasures.qmd), how a variable spreads, and dispersion is the raw material of **uncertainty**. Once we start making **predictions**, that uncertainty splits into two pieces: the **predicted value** $\hat{y}$, what the model expects, and the **residual** $y - \hat{y}$, what the model missed. The residuals are where the model is wrong, and it is the residuals, not the fit, that tell you whether the model is any good and whether your assumption of a straight-line relationship holds.

Fit a line to the curved Dataset 2 and plot the residuals, and the curve the correlation hid comes right back:

```{r}
#| fig-height: 3.4
m2 <- lm(y2 ~ x2, data = anscombe)
plot(fitted(m2), resid(m2), pch = 19, col = "steelblue",
     xlab = "fitted (predicted) value", ylab = "residual (y minus predicted)",
     main = "Dataset 2 residuals arch — the misfit the scatter revealed")
abline(h = 0, lty = 2, col = "grey50")
```

The residuals bow upward at the ends and dip in the middle, the signature of a curved relationship forced through a straight line. The scatter plot and the residual plot told the same story.
:::

Here is why that matters far beyond four small datasets. With two variables you can *see* a bad linear fit directly in the scatter plot. But real models carry many predictors, and you cannot plot a relationship in five or ten dimensions. Anscombe's quartet stops being usable the moment you reach [multiple regression](15-MRC.qmd), because there is no two-dimensional picture left to draw. What survives is the **residual**. Residual analysis carries the "look at where the line is wrong" habit from the simple scatter plot into models too complex to plot. So treat Anscombe's quartet as a bridge, not a destination: it teaches, in pictures anyone can read, the very thing you will later have to check with [residuals](14-bivariateregression.qmd) once the pictures run out.

::: {.callout-important}
## Covariance Is the Engine of the GLM

Here is the payoff that carries into the rest of the book. The slope of a simple regression is *literally* covariance divided by variance:

```{r}
b_from_cov <- cov(study, grade) / var(study)
c(cov_over_var = b_from_cov, lm_slope = coef(lm(grade ~ study))[["study"]])
```

They match exactly. Regression, correlation, and everything downstream are built from covariance. When we say the General Linear Model "models covariance structure," this is what we mean, and it is why this chapter is the real beginning of the modeling half of the book.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Simulate two variables with a *negative* relationship. Compute their covariance and correlation by hand and confirm against `cov()` and `cor()`. What is the sign, and why?
2.  Show numerically that `cov(x, y) / var(x)` equals the slope of `lm(y ~ x)` for your data.
3.  Construct a variable pair with an obvious relationship but a correlation near zero (hint: make it nonlinear). Plot it, and write one sentence on why reporting only $r$ would mislead a reader.
:::

## Where We Go Next

We now hold the engine (covariance) but have not yet built the machine. Before we assemble full models, a short detour to make sure our sensors are trustworthy: are our measurements [reliable](10-reliability.qmd) and [valid](11-validity.qmd)? A model built on a broken ruler predicts nothing, no matter how strong the covariance looks.
