# Covariance and Association

We want to emphasize that most of our work relies heavily on covariance structures. Everything from correlation to regression to ANOVA to structural equation models is, at bottom, an argument about how variables **vary together**. So before we build models, we need to understand the raw material they are built from. Variance measured how a single variable spreads around its own mean; **covariance** measures how *two* variables spread around their means *at the same time*.

## Learning Objectives

1.  See covariance as the natural two-variable extension of variance.
2.  Compute and interpret a covariance, and understand its units problem.
3.  Standardize covariance into correlation, and know what correlation can and cannot tell you.
4.  Recognize covariance as the engine beneath the General Linear Model.

## Why Start With Causes?

Science usually begins with a hunch about a **cause**: sleep affects mood, study time affects grades, a drug affects recovery. We cannot see causes directly; we can only watch variables move. If X truly drives Y, then when X is high we expect Y to be high (or low) *systematically* - the two should **co-vary**. Covariance is our first, humblest measurement of that co-movement. It cannot prove the cause (that famous warning is coming), but a cause that leaves *no* covariance behind is a cause with no evidence.

## From Variance to Covariance

Recall variance: the average product of a deviation *with itself*, $\frac{\sum(x-\bar x)(x-\bar x)}{n-1}$. Covariance makes one small change - it multiplies each variable's deviation by *the other's*:

$$\text{cov}(x, y) = \frac{\sum (x - \bar{x})(y - \bar{y})}{n - 1}$$

Think about the sign of that product. When a point is above the mean on *both* variables (or below on both), the two deviations share a sign and their product is **positive**. When it is high on one and low on the other, the product is **negative**. Sum them up: mostly-positive products mean the variables rise together; mostly-negative means one rises as the other falls.

```{r}
set.seed(5)
study  <- rnorm(100, 5, 2)
grade  <- 60 + 4 * study + rnorm(100, 0, 5)    # grade rises with study

# by hand:
cov_hand <- sum((study - mean(study)) * (grade - mean(grade))) / (length(study) - 1)
c(by_hand = cov_hand, r_cov = cov(study, grade))   # match
```

Positive, as expected: more study, more grade. Let's *see* it, colouring each point by the sign of its deviation product:

```{r}
sign_prod <- (study - mean(study)) * (grade - mean(grade))
plot(study, grade, pch = 19, col = ifelse(sign_prod > 0, "steelblue", "red"),
     main = "Blue points push covariance up; red points push it down")
abline(v = mean(study), h = mean(grade), lty = 2, col = "grey50")
```

The blue points (agreeing deviations) dominate, so covariance is positive. That picture *is* covariance.

## The Units Problem, and the Fix: Correlation

Covariance has a serious drawback for interpretation: its size depends on the **units**. Measure grade in points versus proportions and the covariance changes a lot, even though the relationship is identical. This is the same units problem the standard deviation had, and we fix it the same way, by **standardizing**. Divide covariance by both standard deviations and you get the **correlation coefficient**, $r$:

$$r = \frac{\text{cov}(x, y)}{s_x \, s_y}$$

```{r}
r_hand <- cov(study, grade) / (sd(study) * sd(grade))
c(by_hand = r_hand, r_cor = cor(study, grade))     # match
```

Correlation is covariance stripped of its units, squeezed onto a fixed scale from $-1$ (perfect negative) through $0$ (no linear relationship) to $+1$ (perfect positive). Now the number means the same thing whatever the variables - which is why $r$, not raw covariance, is what we report.

## What Correlation Will Not Tell You

Two lasting warnings. First, the familiar one: **correlation is not causation.** Ice-cream sales and drownings correlate (both driven by summer heat), but ice cream does not cause drowning. Second, and less appreciated: $r$ measures only the **linear** part of a relationship. A strong curved relationship can produce $r \approx 0$. This is why we told you to *always plot your data* - the single number can hide the whole story:

```{r}
# a perfect U-shape: strong relationship, ~zero correlation
xx <- seq(-3, 3, length.out = 100)
yy <- xx^2
c(correlation = round(cor(xx, yy), 3))    # ~0, yet Y is completely determined by X!
plot(xx, yy, pch = 19, col = "darkorange",
     main = "r is near zero, but the relationship is perfect (just not linear)")
```

::: {.callout-important}
## Covariance Is the Engine of the GLM

Here is the payoff that carries into the rest of the book. The slope of a simple regression is *literally* covariance divided by variance:

```{r}
b_from_cov <- cov(study, grade) / var(study)
c(cov_over_var = b_from_cov, lm_slope = coef(lm(grade ~ study))[["study"]])
```

They match exactly. Regression, correlation, and everything downstream are built from covariance. When we say the General Linear Model "models covariance structure," this is what we mean, and it is why this chapter is the real beginning of the modeling half of the book.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Simulate two variables with a *negative* relationship. Compute their covariance and correlation by hand and confirm against `cov()` and `cor()`. What is the sign, and why?
2.  Show numerically that `cov(x, y) / var(x)` equals the slope of `lm(y ~ x)` for your data.
3.  Construct a variable pair with an obvious relationship but a correlation near zero (hint: make it nonlinear). Plot it, and write one sentence on why reporting only $r$ would mislead a reader.
:::

## Where We Go Next

We now hold the engine (covariance) but have not yet built the machine. Before we assemble full models, a short detour to make sure our sensors are trustworthy: are our measurements [reliable](10-reliability.qmd) and [valid](11-validity.qmd)? A model built on a broken ruler predicts nothing, no matter how strong the covariance looks.
