# Statistical Power: How to Find Things That May Not Be There

Simply put, statistical power is the probability of success - where "success" means **detecting an effect that is really there**. In the language of the last chapter, power is the one cell of the error table you can actually *plan for* before you collect a single observation. Everything else about a study - your hypotheses, your $\alpha$, the vagaries of sampling - is set once the data arrive. Power is where you take back control.

## Learning Objectives

1.  Define power as $1 - \beta$ and locate it in the error table.
2.  Identify the four levers that raise or lower power.
3.  Use R to compute power and to plan a sample size.
4.  Understand the honest trade-off between power and Type I error.

## Power Is One Cell of a Table You Already Know

Recall the decision table from [Hypothesis Testing](08-hypothesistesting.qmd):

|                          | $H_0$ is really **true** | $H_0$ is really **false** |
|--------------------------|--------------------------|---------------------------|
| **We reject $H_0$**      | Type I error ($\alpha$)  | **Correct ✓ (power, $1-\beta$)** |
| **We fail to reject $H_0$** | Correct ✓ ($1-\alpha$) | Type II error ($\beta$)   |

**Power** is the highlighted cell: the probability we correctly reject the null *given that the null is false* - that is, given there is a real effect to find. Its complement is $\beta$, the Type II error rate (the miss). By long convention we aim for power of **0.80**: an 80% chance of catching a true effect, and thus a 20% chance of missing it. That 20% should bother you; it is why underpowered research is such a quiet disaster. When power is low, you not only miss real effects - the significant results you *do* find are more likely to be flukes.

## The Four Levers

Power is a function of four things, and only four. Three you can influence; one you mostly cannot.

1.  **Effect size** - how big the true effect is. Bigger effects are easier to detect. We usually express it in standardized units, e.g., Cohen's $d = \frac{\text{mean difference}}{\text{standard deviation}}$. You do not *control* the true effect, but you can sometimes measure it more cleanly to keep it from being buried in noise.
2.  **Sample size ($n$)** - more data, more power. This is the lever you actually turn. Nearly every underpowered study is really a too-small study.
3.  **Significance level ($\alpha$)** - a more *lenient* $\alpha$ (say .10) makes it easier to reject, so power goes **up**; a *stricter* $\alpha$ (say .01) makes it harder to reject, so power goes **down**. (More on this trade-off below - it is exactly backwards in a lot of lecture notes, including an old version of mine.)
4.  **Directionality** - a one-tailed test concentrates all of $\alpha$ in the predicted direction, so it has more power *there* - at the cost of being blind to an effect in the other direction.

::: {.callout-warning}
## The $\alpha$-Power Trade-off (Getting It Right)

It is tempting - and wrong - to say "lower $\alpha$ gives higher power." Lowering $\alpha$ shrinks the rejection region, which makes it *harder* to reject the null, which **lowers** power. You cannot buy fewer false alarms (small $\alpha$) and more hits (high power) for free by tightening $\alpha$ alone; those two pull in opposite directions. The only way to improve *both* at once is to change something outside this trade-off - almost always, **collect a larger sample**.
:::

## Seeing Power in R

Base R ships with `power.t.test()`, which ties together effect size, $n$, $\alpha$, and power - give it any three and it solves for the fourth. First, the trade-offs, made concrete:

```{r}
# Hold the effect (d = 0.5) and n = 30 per group; vary alpha
sapply(c(0.01, 0.05, 0.10), function(a)
  power.t.test(n = 30, delta = 0.5, sd = 1, sig.level = a)$power)
```

Read left to right: as $\alpha$ goes from strict (.01) to lenient (.10), power **rises**. Stricter $\alpha$ costs you power - the trade-off, in numbers.

Now the lever you actually control - sample size:

```{r}
# Hold effect (d = 0.5) and alpha (.05); vary n per group
sapply(c(20, 50, 100), function(nn)
  power.t.test(n = nn, delta = 0.5, sd = 1, sig.level = 0.05)$power)
```

More data, more power - and unlike the $\alpha$ trade-off, this buys power *without* raising your false-alarm rate.

## Planning a Study: Solve for $n$

The real payoff is planning. If you want 80% power to detect a medium effect ($d = 0.5$) at $\alpha = .05$, how many people do you need? Leave `n` out and let R solve for it:

```{r}
power.t.test(delta = 0.5, sd = 1, sig.level = 0.05, power = 0.80)
```

The answer (about 64 per group) is a number you want *before* you run the study, not after. Running first and computing power later - so-called "post-hoc power" - tells you almost nothing useful; power is a design tool, not a consolation prize.

## Power Is Really Just Simulation

If the formulas ever feel like a black box, remember they are only shortcuts for something you can *watch* directly. Simulate a world where the effect is real, run the test many times, and count how often you catch it - that fraction **is** the power:

```{r}
set.seed(1988)
d <- 0.5; n <- 64
hits <- replicate(2000, {
  group1 <- rnorm(n, mean = 0,  sd = 1)
  group2 <- rnorm(n, mean = d,  sd = 1)   # a real difference of d
  t.test(group1, group2)$p.value < 0.05   # did we detect it?
})
mean(hits)   # simulated power ~ 0.80
```

The simulation lands right where `power.t.test()` said it would. No magic - just a claim about how often a true effect trips the wire.

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Use `power.t.test()` to find the sample size needed for 80% power to detect a *small* effect ($d = 0.2$) at $\alpha = .05$. Compare it to the $n$ for a *large* effect ($d = 0.8$). Why the enormous difference?
2.  Fix $n = 40$ per group and $d = 0.4$. Compute power at $\alpha = .01$, $.05$, and $.10$. State in one sentence what you are giving up as you raise $\alpha$ to gain power.
3.  Write the simulation from the last section yourself, but for a one-sample test against a null mean. Confirm your simulated power matches `power.t.test(..., type = "one.sample")`.

If part 2 makes you a little uncomfortable, good - that discomfort *is* the trade-off, and pretending it away is how bad studies get funded.
:::

## Where We Go Next

Power closes out our tour of inference: we now know how to test, how we can be wrong, and how to design a study that can actually find what it is looking for. Next we turn from *testing* our conclusions to *trusting our measurements* - the twin questions of [reliability](10-reliability.qmd) and [validity](11-validity.qmd). After all, the most powerful test in the world is worthless if the ruler is rubber.
