# Statistical Power: How to Find Things That May Not Be There

Simply put, statistical power is the probability of success - where "success" means **detecting an effect that is really there**. In the language of the last chapter, power is the one cell of the error table you can actually *plan for* before you collect a single observation. Everything else about a study - your hypotheses, your $\alpha$, the vagaries of sampling - is set once the data arrive. Power is where you take back control.

::: {.callout-note}
## Power is a SWAG - and that's the honest part
Here is the framing (PEM) I use with my own students. A power analysis is really an estimate of

$$P(\text{Party at the Pub!} \mid \text{your research plans}),$$

the probability you'll be celebrating a detected effect *given* the study you're about to run - which is to say, a **S**cientific **W**ild **A**ss **G**uess. That is not an insult to the method; it is honesty about it, because power depends on things you must *guess* before any data exist (chiefly the true effect size). Why bother guessing at all? Because power protects the things you care about (**funding**, your time, scarce participants), and it is the one part of a study you can shape in advance. The goal is not false precision; it is **self-reliance**. Power estimation is daunting, important, and easy to fool yourself with, and it is *far* more within your reach than most people realize, because when the formulas intimidate you, you can always **simulate** it.
:::

## Learning Objectives

1.  Define power as $1 - \beta$ and locate it in the error table.
2.  Identify the four levers that raise or lower power.
3.  Use R to compute power and to plan a sample size.
4.  Understand the honest trade-off between power and Type I error.

## Power Is One Cell of a Table You Already Know

Recall the decision table from [Hypothesis Testing](08-hypothesistesting.qmd):

|                          | $H_0$ is really **true** | $H_0$ is really **false** |
|--------------------------|--------------------------|---------------------------|
| **We reject $H_0$**      | Type I error ($\alpha$)  | **Correct ✓ (power, $1-\beta$)** |
| **We fail to reject $H_0$** | Correct ✓ ($1-\alpha$) | Type II error ($\beta$)   |

**Power** is the highlighted cell: the probability we correctly reject the null *given that the null is false* - that is, given there is a real effect to find. Its complement is $\beta$, the Type II error rate (the miss). By long convention we aim for power of **0.80**: an 80% chance of catching a true effect, and thus a 20% chance of missing it. That 20% should bother you; it is why underpowered research is such a serious problem. When power is low, you not only miss real effects, but the significant results you *do* find are more likely to be flukes.

## The Four Levers

Power is a function of four things, and only four. Three you can influence; one you mostly cannot.

1.  **Effect size** - how big the true effect is. Bigger effects are easier to detect. We usually express it in standardized units, e.g., Cohen's $d = \frac{\text{mean difference}}{\text{standard deviation}}$. You do not *control* the true effect, but you can sometimes measure it more cleanly to keep it from being buried in noise.
2.  **Sample size ($n$)** - more data, more power. This is the lever you actually turn. Nearly every underpowered study is really a too-small study.
3.  **Significance level ($\alpha$)** - a more *lenient* $\alpha$ (say .10) makes it easier to reject, so power goes **up**; a *stricter* $\alpha$ (say .01) makes it harder to reject, so power goes **down**. (More on this trade-off below; it is exactly backwards in a lot of lecture notes, including an old version of mine.)
4.  **Directionality** - a one-tailed test concentrates all of $\alpha$ in the predicted direction, so it has more power *there* - at the cost of being blind to an effect in the other direction.

::: {.callout-warning}
## The $\alpha$-Power Trade-off (Getting It Right)

It is tempting, and wrong, to say "lower $\alpha$ gives higher power." Lowering $\alpha$ shrinks the rejection region, which makes it *harder* to reject the null, which **lowers** power. You cannot buy fewer false alarms (small $\alpha$) and more hits (high power) for free by tightening $\alpha$ alone; those two pull in opposite directions. The only way to improve *both* at once is to change something outside this trade-off - almost always, **collect a larger sample**.
:::

::: {.callout-note}
## Power is a story about variance
Every lever above is really about the [signal-to-noise](04-Dmeasures.qmd) ratio we learned to decompose as $S^2_{obs} = S^2_{true} + S^2_{error}$. Detecting an effect gets easier as the **between-group** variance (the signal - your effect) grows relative to the **within-group** variance (the noise). That is exactly why Cohen's $d$ puts the mean difference *over* the standard deviation: power lives in the **ratio**, not the numerator alone. So two moves you can sometimes control - a **cleaner measure** (less $S^2_{error}$) and a **stronger manipulation** (more signal) - buy you power just as surely as more subjects do.
:::

## Rules of Thumb (A Rough First Pass)

Before any software, a few back-of-the-envelope numbers get you into the right ballpark for the conventional 80% power:

- **Comparing groups** (mean differences): about **$n = 30$ per group** is the classic target - and never go below ~7 per cell if you can possibly avoid it.
- **Associations** (correlations, $r$): you want **$N > 50$** or so.
- **Measurement models** (EFA/CFA): **$N > 300$** is usually safe.

These are deliberately crude (they hide the effect size you're implicitly assuming), but they are enough to sanity-check a plan, or to notice a design is hopeless *before* you spend anything. Then you refine with a real calculation, which comes in two flavors: a **point estimate** from a formula (next), and a **simulation** you can watch (further down). The right habit is to reach for an existing tool first, make sure you understand the model you're planning (a $t$-test, a regression, a mixed model), and then simulate to check your guesses.

## Seeing Power in R

Base R ships with `power.t.test()`, which ties together effect size, $n$, $\alpha$, and power - give it any three and it solves for the fourth. First, the trade-offs, made concrete:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
This chapter computes everything from first principles, so there is no dataset to load. See [Getting the Book's Data](setup.qmd) if you want the setup files for your language.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
library(broom)        # tidy(), glance(), augment() for model output
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

# Hold the effect (d = 0.5) and n = 30 per group; vary alpha
tibble(alpha = c(0.01, 0.05, 0.10)) |>
  mutate(power = map_dbl(alpha, \(a)
    power.t.test(n = 30, delta = 0.5, sd = 1, sig.level = a)$power)) |>
  round2()
```
### SPSS
```
* Base SPSS has no power analysis. Power and sample-size planning live in
* the separate, separately licensed SamplePower module (Analyze > Power
* Analysis in recent versions), or in an external tool such as G*Power.
*
* * The question here - how power moves as alpha goes .01 -> .05 -> .10 -
* is answered in the R, Julia, and Python tabs.
```
### Julia
```julia
using Distributions
# Power of a two-sample t test: the probability a noncentral t exceeds
# the critical value. delta = 0.5, n = 30 per group.
function power_t(n, delta, alpha)
    df = 2n - 2
    ncp = delta * sqrt(n / 2)
    crit = quantile(TDist(df), 1 - alpha / 2)
    ccdf(NoncentralT(df, ncp), crit) + cdf(NoncentralT(df, ncp), -crit)
end

[power_t(30, 0.5, a) for a in (0.01, 0.05, 0.10)]
```
### Python
```python
from statsmodels.stats.power import TTestIndPower

# Hold the effect (d = 0.5) and n = 30 per group; vary alpha
analysis = TTestIndPower()
[analysis.power(effect_size=0.5, nobs1=30, alpha=a) for a in (0.01, 0.05, 0.10)]
```
:::

Read left to right: as $\alpha$ goes from strict (.01) to lenient (.10), power **rises**. Stricter $\alpha$ costs you power - the trade-off, in numbers.

Now the lever you actually control - sample size:

::: {.panel-tabset}
### R
```{r}
# Hold effect (d = 0.5) and alpha (.05); vary n per group
tibble(n_per_group = c(20, 50, 100)) |>
  mutate(power = map_dbl(n_per_group, \(nn)
    power.t.test(n = nn, delta = 0.5, sd = 1, sig.level = 0.05)$power)) |>
  round2()
```
### SPSS
```
* Base SPSS has no power analysis. Power and sample-size planning live in
* the separate, separately licensed SamplePower module (Analyze > Power
* Analysis in recent versions), or in an external tool such as G*Power.
*
* * The question here - how power grows with n = 20, 50, 100 per group -
* is answered in the R, Julia, and Python tabs.
```
### Julia
```julia
# Hold effect (d = 0.5) and alpha (.05); vary n per group
[power_t(n, 0.5, 0.05) for n in (20, 50, 100)]
```
### Python
```python
# Hold effect (d = 0.5) and alpha (.05); vary n per group
[analysis.power(effect_size=0.5, nobs1=n, alpha=0.05) for n in (20, 50, 100)]
```
:::

More data, more power - and unlike the $\alpha$ trade-off, this buys power *without* raising your false-alarm rate.

## Planning a Study: Solve for $n$

The real payoff is planning. If you want 80% power to detect a medium effect ($d = 0.5$) at $\alpha = .05$, how many people do you need? Leave `n` out and let R solve for it:

::: {.panel-tabset}
### R
```{r}
power.t.test(delta = 0.5, sd = 1, sig.level = 0.05, power = 0.80) |>
  tidy() |>          # broom turns the printed result into a one-row table
  round2()
```
### SPSS
```
* Base SPSS has no power analysis. Power and sample-size planning live in
* the separate, separately licensed SamplePower module (Analyze > Power
* Analysis in recent versions), or in an external tool such as G*Power.
*
* * The question here - the n needed for 80% power at d = 0.5 -
* is answered in the R, Julia, and Python tabs.
```
### Julia
```julia
using Roots
# Solve for the n that delivers 80% power at d = 0.5, alpha = .05
find_zero(n -> power_t(n, 0.5, 0.05) - 0.80, (5.0, 500.0))
```
### Python
```python
# Leave nobs1 out and let statsmodels solve for it
analysis.solve_power(effect_size=0.5, alpha=0.05, power=0.80)
```
:::

The answer (about 64 per group) is a number you want *before* you run the study, not after. Running first and computing power later, so-called "post-hoc power," tells you almost nothing useful; power is a design tool, not a consolation prize.

## Power Is Really Just Simulation

If the formulas ever feel like a black box, remember they are only shortcuts for something you can *watch* directly. The recipe is four steps: **generate** data from a world where the effect is real, **run** your model, **collect** whether it hit ($p < \alpha$), and **repeat** a few thousand times. The fraction of hits **is** the power, and unlike a formula, this recipe works for *any* design you can simulate (a mixed model, a mediation, a custom test), long after the closed-form equations give up:

::: {.panel-tabset}
### R
```{r}
set.seed(1988)
effect <- 0.5; n <- 64

hits <- map_lgl(1:2000, \(i) {
  group1 <- rnorm(n, mean = 0,      sd = 1)
  group2 <- rnorm(n, mean = effect, sd = 1)   # a real difference of d
  t.test(group1, group2)$p.value < 0.05       # did we detect it?
})

tibble(simulated_power = mean(hits)) |> round2()   # ~ 0.80
```
### SPSS
```
* The simulation approach DOES work in base SPSS, because it needs only
* data generation and a t test - no power module required. Generate two
* groups with a real difference, test, repeat, and count the hits.
INPUT PROGRAM.
LOOP #rep = 1 TO 2000.
  LOOP #i = 1 TO 64.
    COMPUTE rep = #rep.
    COMPUTE grp = 1.
    COMPUTE y = RV.NORMAL(0, 1).
    END CASE.
    COMPUTE grp = 2.
    COMPUTE y = RV.NORMAL(0.5, 1).
    END CASE.
  END LOOP.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

* Run the t test within each replication, then the fraction with p < .05
* is the simulated power.
SORT CASES BY rep.
SPLIT FILE BY rep.
T-TEST GROUPS=grp(1 2) /VARIABLES=y.
SPLIT FILE OFF.
```
### Julia
```julia
using Random, Distributions, HypothesisTests, Statistics
Random.seed!(1988)
effect, n = 0.5, 64

hits = [pvalue(UnequalVarianceTTest(rand(Normal(0, 1), n),
                                    rand(Normal(effect, 1), n))) < 0.05
        for _ in 1:2000]

mean(hits)      # simulated power ~ 0.80
```
### Python
```python
import numpy as np
from scipy import stats

rng = np.random.default_rng(1988)
effect, n = 0.5, 64

hits = [stats.ttest_ind(rng.normal(0, 1, n),
                        rng.normal(effect, 1, n)).pvalue < 0.05
        for _ in range(2000)]

np.mean(hits)   # simulated power ~ 0.80
```
:::

The simulation lands right where `power.t.test()` said it would. It is simply a count of how often a true effect trips the wire.

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Use `power.t.test()` to find the sample size needed for 80% power to detect a *small* effect ($d = 0.2$) at $\alpha = .05$. Compare it to the $n$ for a *large* effect ($d = 0.8$). Why the enormous difference?
2.  Fix $n = 40$ per group and $d = 0.4$. Compute power at $\alpha = .01$, $.05$, and $.10$. State in one sentence what you are giving up as you raise $\alpha$ to gain power.
3.  Write the simulation from the last section yourself, but for a one-sample test against a null mean. Confirm your simulated power matches `power.t.test(..., type = "one.sample")`.

If part 2 makes you a little uncomfortable, good: that discomfort *is* the trade-off, and ignoring it is how underpowered studies get funded.
:::

## Where We Go Next

Power closes out our tour of inference: we now know how to test, how we can be wrong, and how to design a study that can actually find what it is looking for. Next we turn from *testing* our conclusions to *trusting our measurements* - the twin questions of [reliability](10-reliability.qmd) and [validity](11-validity.qmd). After all, the most powerful test in the world is worthless if the ruler is rubber.
