# Reliability (not Validity) of Data

Let's introduce an idea that reframes the whole enterprise: your **measures are sensors**. A questionnaire, a stopwatch, a bathroom scale - each is a device that reads out a number *about* something, and like any sensor it is imperfect. The number is not the truth; it is the truth plus some noise. Before we trust a measurement to carry an argument, we must ask two separate questions: is it **consistent** (reliability, this chapter), and is it **on target** (validity, the next)? These are not the same question, and confusing them wrecks research.

## Learning Objectives

1.  Understand the classical test theory model: observed = true + error.
2.  Define reliability as a ratio of variances.
3.  Know the main ways to estimate reliability and when each applies.
4.  See how unreliability *attenuates* the relationships you care about.

## Observed = True + Error

Classical test theory rests on one deceptively simple equation:

$$O = T + E$$

Every **observed** score ($O$) is a person's **true** score ($T$) - what we actually want - plus **error** ($E$), the sensor's noise. Because true score and random error are independent, the same decomposition holds for their *variances*:

$$\text{var}(O) = \text{var}(T) + \text{var}(E)$$

Observed-score variance is always larger than true-score variance, because error only ever *adds* spread. Our whole goal in measurement is to make the error term small.

## Reliability Is a Ratio of Variances

**Reliability** ($\rho$) is the fraction of the observed variance that is *real* - the signal share:

$$\rho = \frac{\text{var}(T)}{\text{var}(O)} = \frac{\text{var}(T)}{\text{var}(T) + \text{var}(E)}$$

It runs from 0 (all noise) to 1 (all signal).

::: {.callout-note}
## From the Field
Reliability is necessary but not sufficient: a consistent measure can still be *meaningless* until it is anchored to something real. Sechrest, McKnight, and McKnight [@SechrestMcKnight1996] made exactly this argument for psychotherapy-outcome measures - **calibration**, not just consistency, is what lets a score be interpreted. It is the same warning the dartboard makes later in this chapter, made in print.
:::

Let's build a world where we *know* the true scores and confirm the definition:

```{r}
set.seed(8)
truth <- rnorm(1000, 100, 15)      # the true scores we wish we could see
error <- rnorm(1000, 0, 8)          # sensor noise
observed <- truth + error           # what we actually record

reliability <- var(truth) / var(observed)
c(reliability = round(reliability, 3))
```

About 0.78 of the variance in our observed scores is real signal; the rest is sensor noise. In practice we can never see `truth`, so we *estimate* reliability from clever comparisons.

## Four Ways to Estimate Reliability

1.  **Test–retest** - measure the same people twice and correlate. High correlation means the sensor gives a stable reading over time. (Careful: real change looks like unreliability.)
2.  **Internal consistency** - do the items of a scale agree with each other? Summarized by **Cronbach's alpha**.
3.  **Split-half** - split the items in two, score each half, correlate. A quick internal-consistency check.
4.  **Inter-rater** - do two human coders agree? Essential whenever judgment is the sensor.

### Cronbach's Alpha, By Hand

Alpha asks: relative to the total score's variance, how small are the individual item variances? If items measure the same thing, the whole is much more than the sum of its jittery parts.

$$\alpha = \frac{k}{k-1}\left(1 - \frac{\sum \text{var(item)}}{\text{var(total)}}\right)$$

```{r}
set.seed(9)
# simulate 6 items that share a common "true" factor plus item-specific noise
common <- rnorm(500)
items  <- sapply(1:6, function(i) common + rnorm(500))   # 500 people x 6 items
items  <- as.data.frame(items)

k         <- ncol(items)
var_items <- sum(apply(items, 2, var))
var_total <- var(rowSums(items))
alpha <- (k / (k - 1)) * (1 - var_items / var_total)
c(cronbach_alpha = round(alpha, 3))
```

An alpha around 0.85 says the six items hang together well; they are largely reading the same underlying construct. By rough convention, 0.70 is "acceptable," 0.80+ "good," but do not lean on the threshold too hard: alpha climbs with more items whether or not they are any good, so a high alpha is necessary, not sufficient.

## Why You Should Care: Attenuation

Now the practical consequence. Unreliable measures **attenuate** (shrink) the correlations you are trying to detect. A real relationship between two constructs looks weaker than it is, purely because your sensors are noisy:

```{r}
set.seed(10)
Tx <- rnorm(1000); Ty <- 0.6 * Tx + rnorm(1000, 0, 0.8)  # true correlation ~ .6
# now observe each through a noisy sensor:
Ox <- Tx + rnorm(1000, 0, 1)
Oy <- Ty + rnorm(1000, 0, 1)
c(true_r = round(cor(Tx, Ty), 2), observed_r = round(cor(Ox, Oy), 2))
```

The observed correlation is noticeably smaller than the true one. A real effect can hide behind a bad ruler, which means a null result might be a measurement failure rather than the absence of an effect. Reliability is more than bookkeeping; it can be the difference between finding your effect and missing it.

## A Real Scale, End to End

Enough simulation. Let's run the whole pipeline on real data: a 10-item scale (`x1`–`x10`) from the graduate GLM course (`data/moddat2.csv`, $n = 30$). Three moves: **key** the items, **estimate** $\alpha$, and check **which items earn their place**.

Some items are worded in reverse, so a high true standing produces a *low* score - flip them first or they fight the rest of the scale and sink $\alpha$. We spot a reverse item by its *negative* item-rest correlation (it correlates the wrong way with the sum of the others):

```{r}
d <- read.csv("data/moddat2.csv")
X <- as.matrix(d[, paste0("x", 1:10)])

alpha_keyed <- function(X) {
  r     <- sapply(seq_len(ncol(X)),                         # item-rest correlations
                  function(j) cor(X[, j], rowSums(X[, -j, drop = FALSE])))
  keys  <- ifelse(r < 0, -1, 1)                             # flip the reverse-worded items
  Xk    <- sweep(X, 2, keys, `*`)
  k     <- ncol(Xk)
  alpha <- (k / (k - 1)) * (1 - sum(apply(Xk, 2, var)) / var(rowSums(Xk)))
  list(alpha = alpha, keys = keys, scored = Xk)
}
a <- alpha_keyed(X)
a$keys              # -1 marks a reverse-scored item
round(a$alpha, 3)   # Cronbach's alpha for the keyed scale
```

A healthy $\alpha$. Now the classic "should I cut this item?" diagnostic - recompute $\alpha$ with each item *deleted*:

```{r}
drop <- sapply(1:10, function(j) alpha_keyed(X[, -j, drop = FALSE])$alpha)
names(drop) <- colnames(X)
round(drop, 3)
```

Every "alpha-if-deleted" value sits *below* the full-scale $\alpha$, so no item is dead weight - each one, even the weakest, adds more signal than noise. (Had any value come out *above* the full $\alpha$, that item would be a candidate for the chopping block: removing it would make the scale *more* reliable.)

## Why Reliability Buys You Power

Now the payoff, and the link back to [statistical power](09-statisticalpower.qmd). Detection is the between-group signal over the within-group noise, and unreliability *is* that noise ($S^2_{error}$). Build the keyed scale score, then ask a real question of it - do women and men differ? - sizing the effect with Cohen's $d$:

```{r}
score <- rowSums(a$scored)                 # the keyed total score

EScalc <- function(dv, cond) {             # McKnight's teaching function for Cohen's d
  lv <- levels(factor(cond))
  x1 <- dv[cond == lv[1]]; x2 <- dv[cond == lv[2]]
  (mean(x1) - mean(x2)) / sqrt((var(x1) + var(x2)) / 2)
}
round(EScalc(score, d$sex), 2)
```

A large $d$. But the point is *why* it was detectable: because the scale was reliable. Had $\alpha$ been 0.5 instead, half the total-score variance would be pure error, the groups would blur together, and the same real difference could sink below significance. **A reliable measure is the cheapest power you will ever buy**: you pay once, in the design, and it pays back in every analysis that follows. (In practice you'd reach for `psych::alpha(X, check.keys = TRUE)`, which does the keying, the drop statistics, and much more in one call - but now you know exactly what it is doing under the hood.)

::: {.callout-important}
## The Dartboard: Consistency ≠ Accuracy

Picture throwing darts. **Reliability** is how *tightly grouped* your throws are - consistency. A tight cluster is reliable *even if it is nowhere near the bullseye*. That "near the bullseye" part - accuracy, being on target - is **validity**, and it is a completely separate question. A bathroom scale that always reads five pounds heavy is perfectly reliable and perfectly wrong. Reliability is *necessary* for validity (you cannot hit the bullseye if your throws scatter randomly) but nowhere near *sufficient*. That is the whole reason the next chapter exists.
:::

::: {.callout-note}
## Beyond Classical Test Theory

Everything in this chapter lives inside **classical test theory** - the $O = T + E$ world, where "reliability" is a single number for a whole test. It is the right place to start, but it has a blind spot: it treats every item as interchangeable and pins a person's score to the particular test they happened to take. Modern measurement - **item response theory** and its **Rasch** model - drops that assumption, estimating each *item's* difficulty and each *person's* ability on a common scale, so that reliability can vary across the range of the trait. We flag it here as the road ahead; the intuitions you build with $O = T + E$ are exactly the ones that make item response theory click later.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Simulate true scores and add three different amounts of error (small, medium, large). Compute reliability each time. Plot reliability against error variance - what is the shape?
2.  Build an 8-item scale where four items share a common factor and four are pure noise. Compute Cronbach's alpha for all 8 items, then for just the good 4. What happens, and what does it warn you about trusting alpha alone?
3.  Re-run the attenuation demo with *more* reliable sensors (smaller measurement error). Confirm the observed correlation creeps back toward the true value.
:::

## Where We Go Next

A reliable sensor is consistent, but consistency is worthless if it is consistently measuring the *wrong thing*. That is the bullseye question - are we measuring what we intend? - and it is the subject of [validity](11-validity.qmd).
