# Introducing the General Linear Model (GLM)

The General Linear Model (GLM) is a statistical model that is used to model the relationship between a single dependent variable and one or more independent variables. Multiple regression and ANOVA are specific applications of the GLM. In fact, the t-test, correlation, regression, ANOVA, and ANCOVA are really one technique seen from different angles. Learn the one, and the rest come with it.

## Learning Objectives

1.  State the General Linear Model in one equation.
2.  See that covariance is what drives it.
3.  Watch a t-test, an ANOVA, and a regression return the *same* answer.
4.  Understand what "general" does and does not mean.

## One Equation to Rule Them All

Strip away the vocabulary and every model in this section is:

$$Y = b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_k X_k + e$$

An outcome $Y$ is modeled as a weighted sum of predictors plus error $e$. That is it. Whether the $X$'s are continuous (regression), categorical (ANOVA), or a mix (ANCOVA) changes only how we *code* them - never the underlying machine. And, as we saw in the [covariance chapter](07-covariance.qmd), the weights themselves are built from covariances and variances. Covariance is the fuel; the GLM is the engine.

## The Same Answer, Three Ways

Let's take one simple question, do two groups differ on an outcome, and answer it with three "different" methods. Watch the p-value:

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
library(broom)        # tidy(), glance(), augment() for model output
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(12)
d <- tibble(group = factor(rep(c("control", "treatment"), each = 40))) |>
  mutate(y = rnorm(80) + if_else(group == "treatment", 0.6, 0))

p_ttest <- t.test(y ~ group, data = d, var.equal = TRUE) |> tidy() |> pull(p.value)
p_anova <- aov(y ~ group, data = d) |> tidy() |> slice(1) |> pull(p.value)
p_regr  <- lm(y ~ group, data = d)  |> tidy() |> slice(2) |> pull(p.value)

tibble(method = c("t-test", "ANOVA", "regression"),
       p      = c(p_ttest, p_anova, p_regr)) |>
  mutate(p = fmt_p(p))

# not merely close when rounded - identical to full machine precision:
all.equal(p_ttest, p_anova) && all.equal(p_ttest, p_regr)
```
### SPSS
```
* Same question, three ways: t-test, ANOVA, regression.
T-TEST GROUPS=group('control' 'treatment') /VARIABLES=y.
ONEWAY y BY group /STATISTICS DESCRIPTIVES.
REGRESSION /STATISTICS COEFF R ANOVA /DEPENDENT y /METHOD=ENTER group.
```
### Julia
```julia
using HypothesisTests, GLM
EqualVarianceTTest(df.y[df.group .== "control"], df.y[df.group .== "treatment"])
ftest(lm(@formula(y ~ group), df).model)
coeftable(lm(@formula(y ~ group), df))
```
### Python
```python
from scipy import stats
import statsmodels.formula.api as smf
from statsmodels.stats.anova import anova_lm

stats.ttest_ind(df.y[df.group == "control"], df.y[df.group == "treatment"])
fit = smf.ols("y ~ group", data=df).fit()
anova_lm(fit)                  # the ANOVA
fit.summary()                  # the regression
```
:::

Three identical p-values. The t-test, the ANOVA, and the regression are literally the same model here - the t-test is a two-group ANOVA, the two-group ANOVA is a regression with one dummy-coded predictor, and (as a bonus) $F = t^2$:

```{r}
t_stat <- t.test(y ~ group, data = d, var.equal = TRUE) |> tidy() |> pull(statistic)
F_stat <- aov(y ~ group, data = d) |> tidy() |> slice(1) |> pull(statistic)

tibble(t_squared = t_stat^2, F = F_stat) |> round2()   # F = t^2
```

Once you see this, a huge amount of statistics stops being a pile of disconnected recipes and becomes a single, learnable idea.

## What "General" Means (and What It Doesn't)

The GLM is **general** because it swallows so many named procedures. But it does carry assumptions - chiefly that the errors are roughly normal and that the relationship is *linear in the parameters*. When the outcome is not continuous-and-normal (a yes/no, a count), we need the **generalized** linear model, logistic and Poisson regression, which we met briefly at the end of the [ANOVA & GLM chapter](16-ANOVAandGLM.qmd). Do not confuse the two: the *general* linear model is the unifying frame for this whole section; the *generalized* linear model extends it to other kinds of outcomes.

::: {.callout-important}
## The Roadmap From Here

Everything remaining in this book is a special case of the equation above:

- **[Point-biserial correlation](13-pointbiserialcorrelation.qmd)** - the bridge: one continuous, one dichotomous variable.
- **[Bivariate regression](14-bivariateregression.qmd)** - one continuous predictor.
- **[Multiple regression](15-MRC.qmd)** - many continuous predictors.
- **[ANOVA and the GLM](16-ANOVAandGLM.qmd)** - categorical predictors via coding.
- **[Basic ANOVA](17-basicANOVA.qmd)** - partitioning variance, the classical view.

Same model, different predictors. Keep the one equation in your head and you will not get lost.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Generate your own two-group data. Confirm that `t.test(..., var.equal = TRUE)`, `aov()`, and `lm()` give the identical p-value, and that $F = t^2$.
2.  Add a *continuous* covariate to the outcome and fit `lm(y ~ group + covariate)`. You have just run an ANCOVA without learning a new technique - explain to a friend why that is not cheating.
3.  In one sentence, state the difference between the *general* linear model and the *generalized* linear model.
:::

## Where We Go Next

Let's build the bridge that makes the "categorical predictors are just regression" claim concrete, using the simplest possible categorical predictor - a single yes/no. That is the [point-biserial correlation](13-pointbiserialcorrelation.qmd).
