# Multiple Regression

One predictor is rarely enough. The world is multivariate: college GPA depends on high-school GPA *and* study time *and* sleep *and* a dozen other things. Multiple regression (often called **MRC**, for Multiple Regression/Correlation) is simply the line from the last chapter grown a dimension:

$$y = b_1 x_1 + b_2 x_2 + \cdots + b_k x_k + a$$

The logic does not change one bit. We still minimize the sum of squared residuals, and we still quantify our uncertainty about every coefficient. What *does* change, and what trips up almost everyone, is what a coefficient now **means** when the predictors are tangled up with each other. That tangle is the story of this chapter.

## Learning Objectives

1.  Understand what a regression coefficient means when other predictors are present (partialling).
2.  Distinguish the semipartial correlation, the partial correlation, and the standardized coefficient.
3.  Interpret $R$, $R^2$, and adjusted $R^2$ as ratios of variance.
4.  Diagnose multicollinearity with the VIF and know your options for dealing with it.

## Some Deliberately Tangled Data

To see the interesting behavior, we need predictors that are correlated with each other. The cleanest way to *manufacture* that is to build them from a shared source, then add noise - so we know the truth going in.

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch15-tangled`, `ch15-typess`. For example, `book_data("ch15-tangled")` in R, Julia, or Python, or `!bookdata name = "ch15-tangled".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.

The real GPA data is `data/gpa.csv`.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
library(broom)        # tidy(), glance(), augment() for model output
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(1936)  # Fisher's "The Design of Experiments"
base <- rnorm(100)
dat <- tibble(
  y  = base + rnorm(100),
  x1 = base + rnorm(100),
  x2 = base + rnorm(100)
)

# x1 and x2 are correlated because they share 'base'
dat |> cor() |> as_tibble(rownames = "variable") |> round2()
```
### SPSS
```{pspp}
INSERT FILE='data/sim/ch15-tangled.sps'.
CORRELATIONS /VARIABLES=y x1 x2.
```
### Julia
```{julia}
using CSV, DataFrames
df = CSV.read("data/sim/ch15-tangled.csv", DataFrame)

using Statistics
cor(Matrix(df))
```
### Python
```{python}
import pandas as pd
df = pd.read_csv("data/sim/ch15-tangled.csv")

df.corr().round(2)
```
:::

Now fit the two-predictor model:

::: {.panel-tabset}
### R
```{r}
lm3 <- lm(y ~ x1 + x2, data = dat)
tidy2(lm3)
```
### SPSS
```{pspp}
INSERT FILE='data/sim/ch15-tangled.sps'.
REGRESSION /STATISTICS COEFF R ANOVA /DEPENDENT y /METHOD=ENTER x1 x2.
```
### Julia
```{julia}
using CSV, DataFrames
df = CSV.read("data/sim/ch15-tangled.csv", DataFrame)

using GLM
coeftable(lm(@formula(y ~ x1 + x2), df))
```
### Python
```{python}
import pandas as pd
df = pd.read_csv("data/sim/ch15-tangled.csv")

import statsmodels.formula.api as smf
smf.ols("y ~ x1 + x2", data=df).fit().params
```
:::

Here is the puzzle that drives the chapter. If you regress `y` on `x1` *alone*, you get one slope. In the joint model above you get a **different** slope. Why? Because in the joint model, `b1` is the effect of `x1` *with `x2` held constant* - it has been **partialled**. R did not lie to you in either model; the two slopes answer two different questions.

## Three Things People Confuse

When predictors overlap, there are three distinct quantities, and sloppy language treats them as one. Picture the classic Venn diagram from Cohen, Cohen, West, and Aiken [@aiken_applied_2002]: circle $Y$ overlaps circles $X_1$ and $X_2$, and $X_1$ and $X_2$ overlap each other. Label the piece of $Y$ that *only* $X_1$ explains as area $a$, and the piece $Y$ has left over (unexplained) as area $e$.

- **Semipartial correlation** $sr_1$: the unique slice, $sr_1^2 = a$. "How much of *all* of Y's variance does $x_1$ uniquely explain?"
- **Partial correlation** $pr_1$: the unique slice relative to what is *left to explain*, $pr_1^2 = \frac{a}{a+e}$. "Of the Y variance not already explained by $x_2$, how much does $x_1$ explain?"
- **Standardized regression coefficient** $\beta_1$: the slope R reports when everything is standardized. This is what `lm()` gives you.

All three are built from the same three correlations, but they are **not** the same number. Here they are, by formula:

```{r}
r <- dat |>
  summarise(r_y1 = cor(x1, y), r_y2 = cor(x2, y), r_12 = cor(x1, x2))

r |>
  summarise(
    # semipartial correlation of x1 (unique slice = area a)
    semipartial = (r_y1 - r_y2 * r_12) / sqrt(1 - r_12^2),
    # partial correlation of x1 (a / (a + e))
    partial     = (r_y1 - r_y2 * r_12) / sqrt((1 - r_y2^2) * (1 - r_12^2)),
    # standardized regression coefficient (beta) for x1
    beta        = (r_y1 - r_y2 * r_12) / (1 - r_12^2)
  ) |>
  round2()
```

Notice they share the same numerator and differ only in the denominator - which is exactly why they get muddled. Let us confirm that $\beta_1$ is really the standardized slope R reports:

::: {.panel-tabset}
### R
```{r}
datz <- dat |> mutate(across(everything(), \(v) (v - mean(v)) / sd(v)))

lm3z <- lm(y ~ x1 + x2, data = datz)
tidy2(lm3z) |> filter(term == "x1")   # compare to beta above
```
### SPSS
```{pspp}
INSERT FILE='data/sim/ch15-tangled.sps'.
* REGRESSION prints the standardized Beta by default.
REGRESSION /STATISTICS COEFF R ANOVA /DEPENDENT y /METHOD=ENTER x1 x2.
```
### Julia
```{julia}
using CSV, DataFrames
df = CSV.read("data/sim/ch15-tangled.csv", DataFrame)

using GLM, Statistics
df.zy  = (df.y  .- mean(df.y))  ./ std(df.y)
df.zx1 = (df.x1 .- mean(df.x1)) ./ std(df.x1)
df.zx2 = (df.x2 .- mean(df.x2)) ./ std(df.x2)
coeftable(lm(@formula(zy ~ zx1 + zx2), df))
```
### Python
```{python}
import pandas as pd
df = pd.read_csv("data/sim/ch15-tangled.csv")

import statsmodels.formula.api as smf
zd = df.apply(lambda v: (v - v.mean()) / v.std())   # z-score every column
smf.ols("y ~ x1 + x2", data=zd).fit().params
```
:::

Same number. So **the coefficient `lm()` reports is a *partial* effect**: it controls for the other predictors. That is the single most important sentence in multiple regression. When someone says "controlling for X," this is the machinery they are invoking.

::: {.callout-important}
## $sr$ vs. $pr$ vs. $\beta$ - Say It Out Loud

- The **semipartial** ($sr$) is $x_1$'s unique contribution measured against *all* of Y. Square it and you get the increase in $R^2$ from adding $x_1$ last. This is usually the one you want to report.
- The **partial** ($pr$) measures the same unique contribution but against only the *unexplained* part of Y, so it is always at least as large in magnitude as the semipartial.
- The **standardized coefficient** ($\beta$) is the slope, in standard-deviation units, and is what the model uses to predict.

Three questions, three numbers. Do not let a textbook (or a reviewer) tell you they are interchangeable.
:::

## $R$, $R^2$, and the "Shrunken" $R^2$

Statistics is about variance, and $R^2$ is the cleanest variance story we have. **$R$** is just the correlation between what we observed and what the model predicts; **$R^2$** is the ratio of predicted-to-total variance:

```{r}
R2 <- augment(lm3) |> summarise(R2 = var(.fitted) / var(y)) |> pull(R2)

augment(lm3) |>
  summarise(R = cor(y, .fitted), R2 = var(.fitted) / var(y)) |>
  mutate(R2_from_glance = glance(lm3)$r.squared) |>
  round2()
```

That is it: $R^2 = \frac{\text{var}(\hat{y})}{\text{var}(y)}$, the fraction of Y's variance the model reproduces. But $R^2$ has a flaw: add any predictor, even pure noise, and it never goes down. So we penalize it for complexity to get the **adjusted (or "shrunken") $R^2$**:

$$R^2_{adj} = 1 - (1 - R^2)\,\frac{n - 1}{n - k - 1}$$

```{r}
n <- nrow(dat); k <- 2

tibble(
  R2_adj          = 1 - (1 - R2) * (n - 1) / (n - k - 1),
  adj_from_glance = glance(lm3)$adj.r.squared
) |>
  round2()
```

The adjustment is why you should *never* judge a model by raw $R^2$ alone. Throw in enough predictors and you can make $R^2$ look impressive while the model is worthless.

## Testing the Coefficients (Rebuilding the Table Again)

As always, we do not trust the machine until we can reproduce it. The standard error of a coefficient is:

$$SE_{b_1} = \frac{s_y}{s_{x_1}} \cdot \sqrt{\text{VIF}_1} \cdot \sqrt{\frac{1 - R^2}{n - k - 1}}$$

where the **variance inflation factor** $\text{VIF}_1 = \frac{1}{1 - R^2_1}$, and $R^2_1$ comes from regressing $x_1$ on all the *other* predictors. Watch:

```{r}
R2_1 <- glance(lm(x1 ~ x2, data = dat))$r.squared   # x1 explained by x2
VIF1 <- 1 / (1 - R2_1)

b1   <- tidy(lm3) |> filter(term == "x1") |> pull(estimate)
SEb1 <- (sd(dat$y) / sd(dat$x1)) * sqrt(VIF1) * sqrt((1 - R2) / (n - k - 1))

tibble(SE = SEb1, t = b1 / SEb1) |>
  mutate(p = 2 * pt(-abs(t), df = n - k - 1),   # df = n - k - 1, NOT n - 1
         p = fmt_p(p)) |>
  round2()

tidy2(lm3) |> filter(term == "x1")              # compare
```

The by-hand values match R's table. Two things to burn in:

1.  The degrees of freedom for a coefficient's $t$ test are $n - k - 1$ - we spend one df per predictor **and** one for the intercept. (An older version of this lecture used $n - 1$ here; that is wrong, and with several predictors it matters.)
2.  The **VIF sits right inside the standard error**. When predictors are collinear, VIF climbs, the SE inflates, $t$ shrinks, and your once-significant predictor goes quiet, even though nothing about the *relationship* changed. Collinearity does not bias your slopes; it makes them **imprecise**.

For the whole model, the omnibus test is an $F$ built from $R^2$:

$$F = \frac{R^2 / k}{(1 - R^2)/(n - k - 1)} \quad \text{with } df_1 = k,\ df_2 = n - k - 1$$

```{r}
Fmodel <- (R2 / k) / ((1 - R2) / (n - k - 1))

tibble(F = Fmodel,
       p = pf(Fmodel, df1 = k, df2 = n - k - 1, lower.tail = FALSE)) |>
  mutate(p = fmt_p(p)) |>
  round2()

# compare: glance() reports the same F, its df, and its p
glance(lm3) |>
  select(statistic, df, df.residual, p.value) |>
  mutate(p.value = fmt_p(p.value)) |>
  round2()
```

## Multicollinearity: Seeing the Damage

Let us watch the VIF at work directly. As two predictors become more correlated, the VIF explodes:

```{r}
#| label: fig-vif
#| fig-cap: "The variance inflation factor as two predictors grow more correlated. Past about .95 it climbs steeply; the dashed line marks the rough VIF = 10 rule of thumb."
tibble(r = seq(0, 0.99, length.out = 500)) |>
  mutate(VIF = 1 / (1 - r^2)) |>
  ggplot(aes(x = r, y = VIF)) +
  # a common (rough) rule-of-thumb threshold
  geom_hline(yintercept = 10, colour = "red", linetype = "dashed") +
  geom_line(linewidth = 1) +
  labs(x = "correlation between X1 and X2", y = "VIF") +
  theme_book()
```

Past a correlation of about 0.95 the curve rises very steeply. A predictor that is 95% predictable from the others carries almost no *unique* information, so the model cannot pin down its slope. The red line marks a rough $\text{VIF} = 10$ rule of thumb, but treat it as a warning sign, not a hard rule.

## Partitioning Variance: Types of Sums of Squares

The VIF told us that collinearity inflates our uncertainty. Here is the deeper reason, and it is a matter of *bookkeeping*: when predictors overlap, the variance they jointly explain cannot be cleanly assigned to one or the other. How you divide it up is a choice, and there are three main conventions, the **types of sums of squares**.

**Type I (sequential).** Each predictor is credited with the variance it explains *after* the predictors entered before it. It is order-dependent, and the pieces add up to the model total. Watch the credit shift when we swap the order:

::: {.panel-tabset}
### R
```{r}
set.seed(42)
d2 <- tibble(x1 = rnorm(200)) |>
  mutate(
    x2 = 0.7 * x1 + rnorm(200, sd = sqrt(1 - 0.49)),  # x1 and x2 correlate ~ .7
    y  = 0.5 * x1 + 0.4 * x2 + rnorm(200)
  )

# x1 first, then x2
anova(lm(y ~ x1 + x2, data = d2)) |>
  tidy() |> slice(1:2) |> select(term, sumsq) |> round2()

# x2 first, then x1 - same data, different split
anova(lm(y ~ x2 + x1, data = d2)) |>
  tidy() |> slice(1:2) |> select(term, sumsq) |> round2()
```
### SPSS
```{pspp}
INSERT FILE='data/sim/ch15-typess.sps'.
* Type I (sequential) sums of squares; DESIGN order sets the entry order.
UNIANOVA y WITH x1 x2 /METHOD=SSTYPE(1) /DESIGN=x1 x2.
UNIANOVA y WITH x1 x2 /METHOD=SSTYPE(1) /DESIGN=x2 x1.
```
### Julia
```{julia}
using CSV, DataFrames
df = CSV.read("data/sim/ch15-typess.csv", DataFrame)

using GLM
# Julia's GLM has no anova() table. Its equivalent is ftest() over NESTED
# models, which is what a Type I split really is: add one term at a time and
# ask what each addition bought. Swap the order and the credit moves.
ftest(lm(@formula(y ~ 1), df).model,
      lm(@formula(y ~ x1), df).model,
      lm(@formula(y ~ x1 + x2), df).model)

ftest(lm(@formula(y ~ 1), df).model,
      lm(@formula(y ~ x2), df).model,
      lm(@formula(y ~ x2 + x1), df).model)
```
### Python
```{python}
import pandas as pd
df = pd.read_csv("data/sim/ch15-typess.csv")

import statsmodels.formula.api as smf
from statsmodels.stats.anova import anova_lm

# typ=1 is the sequential (Type I) split, so the credit depends on term
# order; swap the formula terms and the sums of squares move.
anova_lm(smf.ols("y ~ x1 + x2", data=df).fit(), typ=1)
anova_lm(smf.ols("y ~ x2 + x1", data=df).fit(), typ=1)
```
:::

Same data, same model, but x1 is credited with far more variance when it enters first. Whoever goes first gets the shared variance. Sequential sums of squares reward order, which is appropriate when theory dictates an order and misleading when it does not.

**Type III (unique, or "each last").** Each predictor is credited only with the variance that is *uniquely* its own, the part no other predictor can account for. This is order-free: every predictor is evaluated as if it had entered last. It is exactly the **squared semipartial correlation** from earlier in this chapter, and you can read it off as the drop in fit when you remove that one predictor from the full model:

::: {.panel-tabset}
### R
```{r}
full <- lm(y ~ x1 + x2, data = d2)

# the drop in fit when each predictor is removed from the full model
x1_unique <- anova(lm(y ~ x2, data = d2), full) |> tidy() |> pull(sumsq) |> last()
x2_unique <- anova(lm(y ~ x1, data = d2), full) |> tidy() |> pull(sumsq) |> last()

tibble(x1_unique = x1_unique,   # what x1 adds, controlling x2
       x2_unique = x2_unique) |> # what x2 adds, controlling x1
  round2()
```
### SPSS
```{pspp}
INSERT FILE='data/sim/ch15-typess.sps'.
* Type III (unique / "each last") sums of squares; SPSS's default.
UNIANOVA y WITH x1 x2 /METHOD=SSTYPE(3) /DESIGN=x1 x2.
```
### Julia
```{julia}
using CSV, DataFrames
df = CSV.read("data/sim/ch15-typess.csv", DataFrame)

using GLM
# Type III (unique): compare the FULL model against the model with one term
# dropped, which is exactly "what does this predictor add, controlling for
# everything else".
full = lm(@formula(y ~ x1 + x2), df)

ftest(lm(@formula(y ~ x2), df).model, full.model)   # what x1 adds
ftest(lm(@formula(y ~ x1), df).model, full.model)   # what x2 adds
```
### Python
```{python}
import pandas as pd
df = pd.read_csv("data/sim/ch15-typess.csv")

import statsmodels.formula.api as smf
from statsmodels.stats.anova import anova_lm

# Type III (unique) sums of squares: each term controlling for all others.
anova_lm(smf.ols("y ~ x1 + x2", data=df).fit(), typ=3)
```
:::

These do not depend on order, but they do **not** add up to the model total:

```{r}
tibble(
  unique_sum = x1_unique + x2_unique,
  model_SS   = anova(full) |> tidy() |> slice(1:2) |> pull(sumsq) |> sum()
) |>
  round2()
```

The gap is the **shared variance** - the part both predictors could explain, which Type III refuses to award to either. With collinear predictors that shared piece can be large (here it is most of the model), and that is the collinearity problem stated exactly: much of the explained variance cannot be attributed to one predictor over the other.

**Type II** sits between the two: like Type III it credits each term controlling for the others, but it respects *marginality* (a main effect is tested controlling for other main effects, but not for the interactions that contain it). For a model without interactions, Type II and Type III agree.

::: {.callout-important}
## Which type to report is an interpretive choice

There is no universally correct type; there is the type that answers your question.

- Use **Type I** when the predictors have a genuine, theory-given order: enter the control variable first and the variable of interest last, then read its sequential contribution.
- Use **Type III** (the default in much of psychology, and what standardized coefficients and semipartial correlations report) when you want each predictor's *unique* contribution, controlling for all the others.

The point to carry away is this: whenever predictors are correlated, some of the explained variance is **shared and unattributable**, and different conventions handle it differently. Reporting a single "variance explained by X" without saying how you partitioned it is incomplete. Collinearity is not only a nuisance to diagnose with a VIF; it is a real limit on what the data can say about any one predictor, and *how you partition the variance* is part of the answer.
:::

## Six Ways to Deal With Collinearity

There is no single fix; there are trade-offs. In rough order of how often they are the *right* call:

1.  **Delete a redundant predictor.** If two variables measure nearly the same thing, you probably do not need both. Cheapest fix, but you lose whatever unique signal was there.
2.  **Combine them.** Average the standardized predictors into a single composite (the code is just below). Often the most defensible move when the predictors are indicators of one underlying construct.
3.  **Use a shrinkage method** (e.g., ridge regression), which deliberately trades a little bias for a lot of stability.
4.  **Reduce to principal components** and regress on those - orthogonal by construction, so VIF becomes 1, at the cost of interpretability.
5.  **Center the predictors** (subtract the mean). This does *nothing* for correlational collinearity, but it removes the *nonessential* collinearity created by interaction and polynomial terms - which is why we center before building those.
6.  **Impose logical order.** Sometimes theory tells you which predictor comes "first," and you enter them hierarchically rather than pretending they compete on equal footing.

::: {.callout-note}
## Combining Predictors, In Code

```{r}
d2 |>
  mutate(
    z1        = (x1 - mean(x1)) / sd(x1),
    z2        = (x2 - mean(x2)) / sd(x2),
    composite = (z1 + z2) / 2
  ) |>
  lm(y ~ composite, data = _) |>
  tidy2()
```

One predictor, no collinearity, and often a cleaner story than two fighting over the same variance.
:::

## A Real Example: Predicting College GPA

Let's leave the simulated world and put all of this to work on the real GPA dataset from the [last chapter](14-bivariateregression.qmd). There we predicted college GPA from high-school GPA alone. Now we add the other things an admissions office actually has: SAT total and the quality of the applicant's letters of recommendation.

::: {.panel-tabset}
### R
```{r}
gpa <- read_csv("data/gpa.csv", show_col_types = FALSE)
fit <- lm(CollegeGPA ~ HSGPA + SATtotal + QLOR, data = gpa)

tidy2(fit)
glance(fit) |> select(r.squared, adj.r.squared) |> round2()
```
### SPSS
```{pspp}
GET DATA /TYPE=TXT /FILE='data/gpa.csv'
  /DELIMITERS=',' /QUALIFIER='"' /FIRSTCASE=2
  /VARIABLES=CollegeGPA F16.6 HSGPA F16.6 SATtotal F16.6 QLOR F16.6.
REGRESSION /STATISTICS COEFF R ANOVA
  /DEPENDENT CollegeGPA /METHOD=ENTER HSGPA SATtotal QLOR.
```
### Julia
```{julia}
using CSV, DataFrames, GLM
gpa = CSV.read("data/gpa.csv", DataFrame)

coeftable(lm(@formula(CollegeGPA ~ HSGPA + SATtotal + QLOR), gpa))
```
### Python
```{python}
import pandas as pd
import statsmodels.formula.api as smf
gpa = pd.read_csv("data/gpa.csv")
fit = smf.ols("CollegeGPA ~ HSGPA + SATtotal + QLOR", data=gpa).fit()
fit.summary()                  # coefficients, R-squared, adjusted R-squared
```
:::

Every coefficient here is a **partial** effect - the pull of that predictor with the other two held constant. Notice HSGPA's slope *shrank* from its bivariate value: some of what looked like "high-school GPA" was really SAT and letter quality traveling alongside it. That is partialling doing its job.

Now check for collinearity by hand, exactly as we defined it - each VIF is $1/(1-R^2_j)$ where $R^2_j$ regresses that predictor on the others:

```{r}
preds <- c("HSGPA", "SATtotal", "QLOR")

tibble(predictor = preds) |>
  mutate(VIF = map_dbl(predictor, \(p) {
    others <- setdiff(preds, p)                       # regress p on the rest
    r2 <- glance(lm(reformulate(others, response = p), data = gpa))$r.squared
    1 / (1 - r2)
  })) |>
  round2()
```

All three VIFs sit near 1 - our predictors carry mostly *distinct* information, so the model can pin down each slope cleanly. (Recall the HSGPA–QLOR correlation was the highest at 0.63; that is real but nowhere near the danger zone.) This is the whole workflow in miniature: fit, read the partial effects, and check that collinearity has not quietly wrecked your standard errors.

::: {.callout-tip}
## Do One Yourself

Simulate three predictors where `x3` is nearly a copy of `x1` (e.g., `x3 <- x1 + rnorm(100, 0, 0.2)`), plus an outcome `y` that truly depends on `x1` and `x2`.

1.  Fit `y ~ x1 + x2 + x3` and report the coefficients, their standard errors, and each VIF (compute the VIFs by hand with `1/(1 - R2_j)`).
2.  Now drop `x3` and refit. What happened to the standard error and $t$ for `x1`? Explain *why* in terms of the VIF sitting inside the SE.
3.  Compute the semipartial correlation for `x2` in the full model, and confirm its square equals the drop in $R^2$ when you remove `x2`.

If you can explain part 2 to a skeptical colleague, you understand multicollinearity better than the software does.
:::

## Where We Go Next

We have been treating every predictor as continuous. But some of the most important predictors are **categorical** - treatment vs. control, cat vs. dog, before vs. after. It turns out ANOVA is not a separate technique at all; it is just multiple regression with cleverly coded categorical predictors. That unification - the General Linear Model - is the subject of the next chapter, [ANOVA and the GLM](16-ANOVAandGLM.qmd).
