# Multiple Regression

One predictor is rarely enough. The world is multivariate: college GPA depends on high-school GPA *and* study time *and* sleep *and* a dozen other things. Multiple regression (often called **MRC**, for Multiple Regression/Correlation) is simply the line from the last chapter grown a dimension:

$$y = b_1 x_1 + b_2 x_2 + \cdots + b_k x_k + a$$

The logic does not change one bit. We still minimize the sum of squared residuals, and we still quantify our uncertainty about every coefficient. What *does* change, and what trips up almost everyone, is what a coefficient now **means** when the predictors are tangled up with each other. That tangle is the story of this chapter.

## Learning Objectives

1.  Understand what a regression coefficient means when other predictors are present (partialling).
2.  Distinguish the semipartial correlation, the partial correlation, and the standardized coefficient.
3.  Interpret $R$, $R^2$, and adjusted $R^2$ as ratios of variance.
4.  Diagnose multicollinearity with the VIF and know your options for dealing with it.

## Some Deliberately Tangled Data

To see the interesting behavior, we need predictors that are correlated with each other. The cleanest way to *manufacture* that is to build them from a shared source, then add noise - so we know the truth going in.

```{r}
set.seed(1936)  # Fisher's "The Design of Experiments"
base <- rnorm(100)
y  <- base + rnorm(100)
x1 <- base + rnorm(100)
x2 <- base + rnorm(100)
dat <- data.frame(y, x1, x2)

# x1 and x2 are correlated because they share 'base'
round(cor(dat), 2)
```

Now fit the two-predictor model:

```{r}
lm3 <- lm(y ~ x1 + x2, data = dat)
coef(lm3)
```

Here is the puzzle that drives the chapter. If you regress `y` on `x1` *alone*, you get one slope. In the joint model above you get a **different** slope. Why? Because in the joint model, `b1` is the effect of `x1` *with `x2` held constant* - it has been **partialled**. R did not lie to you in either model; the two slopes answer two different questions.

## Three Things People Confuse

When predictors overlap, there are three distinct quantities, and sloppy language treats them as one. Picture the classic Venn diagram from Cohen, Cohen, West, and Aiken [@aiken_applied_2002]: circle $Y$ overlaps circles $X_1$ and $X_2$, and $X_1$ and $X_2$ overlap each other. Label the piece of $Y$ that *only* $X_1$ explains as area $a$, and the piece $Y$ has left over (unexplained) as area $e$.

- **Semipartial correlation** $sr_1$: the unique slice, $sr_1^2 = a$. "How much of *all* of Y's variance does $x_1$ uniquely explain?"
- **Partial correlation** $pr_1$: the unique slice relative to what is *left to explain*, $pr_1^2 = \frac{a}{a+e}$. "Of the Y variance not already explained by $x_2$, how much does $x_1$ explain?"
- **Standardized regression coefficient** $\beta_1$: the slope R reports when everything is standardized. This is what `lm()` gives you.

All three are built from the same three correlations, but they are **not** the same number. Here they are, by formula:

```{r}
r_y1 <- cor(x1, y); r_y2 <- cor(x2, y); r_12 <- cor(x1, x2)

# semipartial correlation of x1 (unique slice = area a)
sr1 <- (r_y1 - r_y2 * r_12) / sqrt(1 - r_12^2)

# partial correlation of x1 (a / (a + e))
pr1 <- (r_y1 - r_y2 * r_12) / sqrt((1 - r_y2^2) * (1 - r_12^2))

# standardized regression coefficient (beta) for x1
beta1 <- (r_y1 - r_y2 * r_12) / (1 - r_12^2)

round(c(semipartial = sr1, partial = pr1, beta = beta1), 3)
```

Notice they share the same numerator and differ only in the denominator - which is exactly why they get muddled. Let us confirm that $\beta_1$ is really the standardized slope R reports:

```{r}
lm3z <- lm(scale(y) ~ scale(x1) + scale(x2), data = dat)
round(coef(lm3z)[["scale(x1)"]], 3)   # compare to beta1 above
```

Same number. So **the coefficient `lm()` reports is a *partial* effect**: it controls for the other predictors. That is the single most important sentence in multiple regression. When someone says "controlling for X," this is the machinery they are invoking.

::: {.callout-important}
## $sr$ vs. $pr$ vs. $\beta$ - Say It Out Loud

- The **semipartial** ($sr$) is $x_1$'s unique contribution measured against *all* of Y. Square it and you get the increase in $R^2$ from adding $x_1$ last. This is usually the one you want to report.
- The **partial** ($pr$) measures the same unique contribution but against only the *unexplained* part of Y, so it is always at least as large in magnitude as the semipartial.
- The **standardized coefficient** ($\beta$) is the slope, in standard-deviation units, and is what the model uses to predict.

Three questions, three numbers. Do not let a textbook (or a reviewer) tell you they are interchangeable.
:::

## $R$, $R^2$, and the "Shrunken" $R^2$

Statistics is about variance, and $R^2$ is the cleanest variance story we have. **$R$** is just the correlation between what we observed and what the model predicts; **$R^2$** is the ratio of predicted-to-total variance:

```{r}
R  <- cor(y, predict(lm3))
R2 <- var(predict(lm3)) / var(y)
round(c(R = R, R2 = R2, R2_from_summary = summary(lm3)$r.squared), 3)
```

That is it: $R^2 = \frac{\text{var}(\hat{y})}{\text{var}(y)}$, the fraction of Y's variance the model reproduces. But $R^2$ has a flaw: add any predictor, even pure noise, and it never goes down. So we penalize it for complexity to get the **adjusted (or "shrunken") $R^2$**:

$$R^2_{adj} = 1 - (1 - R^2)\,\frac{n - 1}{n - k - 1}$$

```{r}
n <- length(y); k <- 2
R2_adj <- 1 - (1 - R2) * (n - 1) / (n - k - 1)
round(c(R2_adj = R2_adj, adj_from_summary = summary(lm3)$adj.r.squared), 3)
```

The adjustment is why you should *never* judge a model by raw $R^2$ alone. Throw in enough predictors and you can make $R^2$ look impressive while the model is worthless.

## Testing the Coefficients (Rebuilding the Table Again)

As always, we do not trust the machine until we can reproduce it. The standard error of a coefficient is:

$$SE_{b_1} = \frac{s_y}{s_{x_1}} \cdot \sqrt{\text{VIF}_1} \cdot \sqrt{\frac{1 - R^2}{n - k - 1}}$$

where the **variance inflation factor** $\text{VIF}_1 = \frac{1}{1 - R^2_1}$, and $R^2_1$ comes from regressing $x_1$ on all the *other* predictors. Watch:

```{r}
R2_1  <- summary(lm(x1 ~ x2, data = dat))$r.squared   # x1 explained by x2
VIF1  <- 1 / (1 - R2_1)

SEb1 <- (sd(y) / sd(x1)) * sqrt(VIF1) * sqrt((1 - R2) / (n - k - 1))
b1   <- coef(lm3)[["x1"]]
t1   <- b1 / SEb1
p1   <- 2 * pt(-abs(t1), df = n - k - 1)      # df = n - k - 1, NOT n - 1

round(c(SE = SEb1, t = t1, p = p1), 4)
summary(lm3)$coefficients["x1", ]              # compare
```

The by-hand values match R's table. Two things to burn in:

1.  The degrees of freedom for a coefficient's $t$ test are $n - k - 1$ - we spend one df per predictor **and** one for the intercept. (An older version of this lecture used $n - 1$ here; that is wrong, and with several predictors it matters.)
2.  The **VIF sits right inside the standard error**. When predictors are collinear, VIF climbs, the SE inflates, $t$ shrinks, and your once-significant predictor goes quiet, even though nothing about the *relationship* changed. Collinearity does not bias your slopes; it makes them **imprecise**.

For the whole model, the omnibus test is an $F$ built from $R^2$:

$$F = \frac{R^2 / k}{(1 - R^2)/(n - k - 1)} \quad \text{with } df_1 = k,\ df_2 = n - k - 1$$

```{r}
Fmodel <- (R2 / k) / ((1 - R2) / (n - k - 1))
pF     <- pf(Fmodel, df1 = k, df2 = n - k - 1, lower.tail = FALSE)
round(c(F = Fmodel, p = pF), 5)
summary(lm3)$fstatistic          # compare F and its df
```

## Multicollinearity: Seeing the Damage

Let us watch the VIF at work directly. As two predictors become more correlated, the VIF explodes:

```{r}
curve(1 / (1 - x^2), from = 0, to = 0.99, n = 500,
      xlab = "correlation between X1 and X2", ylab = "VIF",
      main = "Collinearity inflates the variance of your estimates")
abline(h = 10, col = "red", lty = 2)   # a common (rough) rule-of-thumb threshold
```

Past a correlation of about 0.95 the curve rises very steeply. A predictor that is 95% predictable from the others carries almost no *unique* information, so the model cannot pin down its slope. The red line marks a rough $\text{VIF} = 10$ rule of thumb, but treat it as a warning sign, not a hard rule.

## Six Ways to Deal With Collinearity

There is no single fix; there are trade-offs. In rough order of how often they are the *right* call:

1.  **Delete a redundant predictor.** If two variables measure nearly the same thing, you probably do not need both. Cheapest fix, but you lose whatever unique signal was there.
2.  **Combine them.** Average the standardized predictors into a single composite: `rowMeans(scale(cbind(x1, x2)))`. Often the most defensible move when the predictors are indicators of one underlying construct.
3.  **Use a shrinkage method** (e.g., ridge regression), which deliberately trades a little bias for a lot of stability.
4.  **Reduce to principal components** and regress on those - orthogonal by construction, so VIF becomes 1, at the cost of interpretability.
5.  **Center the predictors** (subtract the mean). This does *nothing* for correlational collinearity, but it removes the *nonessential* collinearity created by interaction and polynomial terms - which is why we center before building those.
6.  **Impose logical order.** Sometimes theory tells you which predictor comes "first," and you enter them hierarchically rather than pretending they compete on equal footing.

::: {.callout-note}
## Combining Predictors, In Code

```{r}
composite <- rowMeans(data.frame(x1z = scale(x1), x2z = scale(x2)))
round(summary(lm(y ~ composite))$coefficients, 3)
```

One predictor, no collinearity, and often a cleaner story than two fighting over the same variance.
:::

## A Real Example: Predicting College GPA

Let's leave the simulated world and put all of this to work on the real GPA dataset from the [last chapter](14-bivariateregression.qmd). There we predicted college GPA from high-school GPA alone. Now we add the other things an admissions office actually has: SAT total and the quality of the applicant's letters of recommendation.

```{r}
gpa <- read.csv("data/gpa.csv")
fit <- lm(CollegeGPA ~ HSGPA + SATtotal + QLOR, data = gpa)
round(summary(fit)$coefficients, 4)
c(R2 = summary(fit)$r.squared, adj_R2 = summary(fit)$adj.r.squared)
```

Every coefficient here is a **partial** effect - the pull of that predictor with the other two held constant. Notice HSGPA's slope *shrank* from its bivariate value: some of what looked like "high-school GPA" was really SAT and letter quality traveling alongside it. That is partialling doing its job.

Now check for collinearity by hand, exactly as we defined it - each VIF is $1/(1-R^2_j)$ where $R^2_j$ regresses that predictor on the others:

```{r}
preds <- c("HSGPA", "SATtotal", "QLOR")
vif <- sapply(preds, function(p) {
  others <- setdiff(preds, p)
  r2 <- summary(lm(as.formula(paste(p, "~", paste(others, collapse = "+"))), data = gpa))$r.squared
  1 / (1 - r2)
})
round(vif, 2)
```

All three VIFs sit near 1 - our predictors carry mostly *distinct* information, so the model can pin down each slope cleanly. (Recall the HSGPA–QLOR correlation was the highest at 0.63; that is real but nowhere near the danger zone.) This is the whole workflow in miniature: fit, read the partial effects, and check that collinearity has not quietly wrecked your standard errors.

::: {.callout-tip}
## Do One Yourself

Simulate three predictors where `x3` is nearly a copy of `x1` (e.g., `x3 <- x1 + rnorm(100, 0, 0.2)`), plus an outcome `y` that truly depends on `x1` and `x2`.

1.  Fit `y ~ x1 + x2 + x3` and report the coefficients, their standard errors, and each VIF (compute the VIFs by hand with `1/(1 - R2_j)`).
2.  Now drop `x3` and refit. What happened to the standard error and $t$ for `x1`? Explain *why* in terms of the VIF sitting inside the SE.
3.  Compute the semipartial correlation for `x2` in the full model, and confirm its square equals the drop in $R^2$ when you remove `x2`.

If you can explain part 2 to a skeptical colleague, you understand multicollinearity better than the software does.
:::

## Where We Go Next

We have been treating every predictor as continuous. But some of the most important predictors are **categorical** - treatment vs. control, cat vs. dog, before vs. after. It turns out ANOVA is not a separate technique at all; it is just multiple regression with cleverly coded categorical predictors. That unification - the General Linear Model - is the subject of the next chapter, [ANOVA and the GLM](16-ANOVAandGLM.qmd).
