# Basic ANOVA

We closed the last chapter with a slogan: **ANOVA is regression with categorical predictors.** That is true, and it is the deep view. But ANOVA also has a classical view, one built entirely out of the variance ideas from early in the book, and seeing that view makes the machinery click. Analysis of variance earns its name honestly: it works by *partitioning variance* into the part explained by group membership and the part left over. Let's take it apart.

## Learning Objectives

1.  Partition the total sum of squares into between-group and within-group pieces.
2.  Build the ANOVA $F$ ratio from those pieces and understand what it compares.
3.  Confirm that the classical ANOVA and the regression GLM give the identical test.
4.  Follow a significant omnibus test with appropriate post-hoc comparisons.

## The Question and the Data

One-way ANOVA asks: **do three or more groups differ on a continuous outcome?** Say we test a plant fertilizer at three doses:

```{r}
set.seed(17)
dose  <- factor(rep(c("none", "low", "high"), each = 30),
                levels = c("none", "low", "high"))
growth <- rnorm(90, mean = rep(c(10, 12, 15), each = 30), sd = 3)
tapply(growth, dose, mean)
```

The group means differ in our sample - but is that real, or just sampling noise? ANOVA answers by comparing two kinds of variation.

## Partitioning the Variance

Here is the whole idea in one line: the **total** spread of every observation around the grand mean can be split cleanly into spread *between* groups and spread *within* groups.

$$SS_{total} = SS_{between} + SS_{within}$$

- $SS_{total}$: how far each observation is from the overall (grand) mean.
- $SS_{between}$: how far each *group mean* is from the grand mean - the variation our grouping *explains* (the signal).
- $SS_{within}$: how far each observation is from *its own group's* mean - the leftover variation (the noise).

Let's build all three by hand and confirm they add up:

```{r}
grand <- mean(growth)
group_means <- tapply(growth, dose, mean)

SS_total   <- sum((growth - grand)^2)
SS_between <- sum(table(dose) * (group_means - grand)^2)
SS_within  <- sum((growth - group_means[dose])^2)

c(SS_total = SS_total,
  between_plus_within = SS_between + SS_within)   # they match
```

The total variance really does split exactly into "explained by group" plus "unexplained." That partition *is* analysis of variance.

## The F Ratio

We cannot compare sums of squares directly - they depend on how many things went into each. We convert each to a **mean square** by dividing by its degrees of freedom, then take their ratio:

$$F = \frac{MS_{between}}{MS_{within}} = \frac{SS_{between}/(k-1)}{SS_{within}/(N-k)}$$

with $k$ groups and $N$ observations. $F$ asks a simple question: **is the variation between groups bigger than the variation within them?** If groups matter, the between-group spread is large relative to the within-group noise and $F$ climbs well above 1.

```{r}
k <- 3; N <- 90
MS_between <- SS_between / (k - 1)
MS_within  <- SS_within  / (N - k)
F_stat <- MS_between / MS_within
p_val  <- pf(F_stat, df1 = k - 1, df2 = N - k, lower.tail = FALSE)
round(c(F = F_stat, p = p_val), 5)
```

And of course, R will do the bookkeeping for us with `aov()` - confirm it matches our by-hand work:

::: {.panel-tabset}
### R
```{r}
summary(aov(growth ~ dose))
```
### SPSS
```
ONEWAY growth BY dose /STATISTICS DESCRIPTIVES /POSTHOC=TUKEY.
```
### Julia
```julia
using GLM
ftest(lm(@formula(growth ~ dose), df).model)
```
:::

Same $F$, same p, exactly as we computed by hand.

## It Really Is the Same as Regression

To honor the slogan from the last chapter, let's prove the classical ANOVA and the regression GLM are one and the same by pulling the identical $F$ out of `lm()`:

::: {.panel-tabset}
### R
```{r}
fit <- lm(growth ~ dose)
anova(fit)              # the regression's F test == the ANOVA
c(regression_F = summary(fit)$fstatistic[1],
  aov_F        = summary(aov(growth ~ dose))[[1]][["F value"]][1])
```
### SPSS
```
REGRESSION /STATISTICS COEFF R ANOVA /DEPENDENT growth /METHOD=ENTER dose.
```
### Julia
```julia
using GLM
coeftable(lm(@formula(growth ~ dose), df))
```
:::

Identical. The "two techniques" were always one. ANOVA is just the variance-partitioning *story* told about a regression on categorical predictors.

## After a Significant F: Post-Hoc Comparisons

The omnibus $F$ only says "*somewhere* among these groups there is a difference." It does not say *which* groups differ. For that we run **post-hoc** comparisons - and because we are now making several comparisons at once, we must control the inflated false-alarm rate (recall Type I error from the [hypothesis-testing chapter](08-hypothesistesting.qmd)). Tukey's Honest Significant Difference does exactly that:

```{r}
TukeyHSD(aov(growth ~ dose))
```

Each row is a pairwise comparison with a confidence interval already adjusted for the fact that we peeked at all of them. Any interval that excludes zero is a real difference; any that includes zero is not.

::: {.callout-important}
## Do Not Skip Straight to the Pairwise Tests

The two-step procedure, omnibus $F$ first and post-hoc comparisons only if it is significant, is not just a formality. Running every pairwise t-test without the omnibus gate (and without correction) is how you manufacture false positives, one comparison at a time. The $F$ is the gatekeeper, and Tukey (or a similar correction) keeps the gate honest.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Simulate four groups on a continuous outcome. Compute $SS_{total}$, $SS_{between}$, and $SS_{within}$ by hand, confirm they sum, and build $F$ from the mean squares. Check against `summary(aov(...))`.
2.  Confirm the same $F$ falls out of `anova(lm(y ~ group))`.
3.  Make the group means nearly equal, then very different, keeping the within-group SD fixed. Watch what happens to $F$ - and explain it in terms of the between-versus-within comparison.
:::

## Where We Go Next

You now hold the thread that runs through all of it: **quantify your uncertainty, and remember that variance is the currency you quantify it in.** From the first histogram to this $F$ ratio, it was one idea the whole way down. But there is a loose end worth pulling. Every ANOVA you just ran handed R a factor and trusted it to build the right columns of numbers, and *how* those columns are built quietly decides what each coefficient means. That is the subject of the [next chapter](19-coding.qmd), where we stop trusting the defaults and take the coding into our own hands.
