# Measuring Like Ben Wright: Rasch in R, Python, and Julia

In the [reliability chapter](10-reliability.qmd) we flagged a road past classical test theory. Here we walk it, in the company of the person who paved it. **Benjamin D. Wright** (1926–2015) of the University of Chicago spent a career insisting that psychology could have *real measurement* - the kind physicists take for granted, where the ruler does not depend on what you happen to measure with it, and the measurement does not depend on which ruler you happened to grab. His software **BIGSTEPS** (and its descendant WINSTEPS) put the Rasch model in the hands of a generation.

Wright liked to say that if you understood the arithmetic, you could compute a Rasch analysis on the back of an envelope. Let's honor that: we will build a working Rasch calibration - the same computation BIGSTEPS performs - *from scratch*, in about thirty lines, and then show it is the same thirty lines in **R**, **Python**, and **Julia**. Ben would have wanted his legacy to run in whatever language you love.

## Learning Objectives

1.  State the dichotomous Rasch model and why its estimates are "objective."
2.  Understand joint maximum likelihood estimation (JMLE) - the BIGSTEPS engine.
3.  Calibrate items and measure persons from scratch in R.
4.  Recognize that the method is the mathematics, not the software - the same code runs in three languages.

## The Rasch Model

The dichotomous Rasch model says the probability that person $v$ (with ability $\theta_v$) answers item $i$ (with difficulty $\delta_i$) correctly depends on **one thing**: the difference between ability and difficulty.

$$P(x_{vi} = 1) = \frac{e^{\theta_v - \delta_i}}{1 + e^{\theta_v - \delta_i}}$$

That is the logistic curve from the [z-distribution](06-zdistribution.qmd) and [GLM](12-GLMcovariance.qmd) chapters, now measuring *people* against *items* on a single shared scale (the **logit**). Two consequences make Wright's eyes light up:

- **Sufficiency.** A person's raw score (number correct) carries *all* the information about their ability; an item's raw score (number of people who passed) carries all the information about its difficulty. Nothing else matters - not *which* items, not *which* people.
- **Objectivity.** Because ability and difficulty live on the same subtractive scale, item calibrations come out (in principle) the same whatever sample you use, and person measures come out the same whatever items you use. Wright called this **sample-free** item calibration and **test-free** person measurement [@wright_best_1979]. It is the closest psychology gets to a physicist's ruler.

## Wright's Data: The Knox Cube Test

Wright's running example in *Best Test Design* [@wright_best_1979] is the **Knox Cube Test (KCT)**: a child watches the examiner tap a sequence on four cubes, then tries to repeat it. Each of the 18 items is a tapping pattern, and the items are named by the taps - `1-4`, `2-3`, `1-3-2-4`, and so on. Longer sequences are harder, so the KCT is very nearly a perfect ladder of difficulty - which is exactly why Wright loved it for teaching.

We use Wright's **actual data**: 35 children, 18 items, the very responses from *Best Test Design*, stored as a plain text table (`data/knox.dat`). The children even keep their names - Richard, Tracie, Walter, and the rest.

::: {.callout-note}
## Working in SPSS, Julia, or Python?
This chapter uses Ben Wright's real Knox Cube Test responses, `data/knox.dat`, which every tab reads directly. See [Getting the Book's Data](setup.qmd) if you want the setup files for your language.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

## Wright's own Knox Cube Test responses (Best Test Design, 1979):
## 35 children (rows) by 18 tapping items (columns), 1 = repeated correctly.
kct <- read_table("data/knox.dat", show_col_types = FALSE)

taps <- c("1-4","2-3","1-2-4","1-3-4","2-1-4","3-4-1","1-4-3-2","1-4-2-3",
          "1-3-2-4","2-4-3-1","1-3-1-2-4","1-3-2-4-3","1-4-3-2-4","1-4-2-3-4-1",
          "1-3-2-4-1-3","1-4-2-3-1-4","1-4-3-1-2-4","4-1-3-4-2-1-4")

# the estimator below is matrix arithmetic, so we hand it a matrix
X <- kct |> select(num_range("i", 1:18)) |> as.matrix()
rownames(X) <- kct$Name
colnames(X) <- taps
X[1:6, 1:9]   # a corner: the first children, the easiest items
```
### SPSS
```
* Reading Wright's Knox Cube Test responses: 35 children (rows) by 18
* tapping items (columns), 1 = repeated the pattern correctly.
GET DATA /TYPE=TXT /FILE='data/knox.dat'
  /DELIMITERS=' \t' /FIRSTCASE=2
  /VARIABLES=Name A10 sex A1 i1 TO i18 F1.0.

LIST VARIABLES=Name i1 TO i9 /CASES=6.
```
### Julia
```julia
using DelimitedFiles, DataFrames
raw = readdlm("data/knox.dat", header = true)
kct = DataFrame(raw[1], vec(raw[2]))

taps = ["1-4","2-3","1-2-4","1-3-4","2-1-4","3-4-1","1-4-3-2","1-4-2-3",
        "1-3-2-4","2-4-3-1","1-3-1-2-4","1-3-2-4-3","1-4-3-2-4","1-4-2-3-4-1",
        "1-3-2-4-1-3","1-4-2-3-1-4","1-4-3-1-2-4","4-1-3-4-2-1-4"]

X = Matrix(select(kct, r"^i\d+$"))
X[1:6, 1:9]        # a corner: the first children, the easiest items
```
### Python
```python
import numpy as np

# skip the header, drop the Name and sex columns
rows = [ln.split() for ln in open("data/knox.dat")][1:]
X = np.array([[int(v) for v in r[2:]] for r in rows], float)

taps = ["1-4","2-3","1-2-4","1-3-4","2-1-4","3-4-1","1-4-3-2","1-4-2-3",
        "1-3-2-4","2-4-3-1","1-3-1-2-4","1-3-2-4-3","1-4-3-2-4","1-4-2-3-4-1",
        "1-3-2-4-1-3","1-4-2-3-1-4","1-4-3-1-2-4","4-1-3-4-2-1-4"]

X[:6, :9]          # a corner: the first children, the easiest items
```
:::

Before we estimate anything, look at the item margins - how many of the 35 children passed each item:

::: {.panel-tabset}
### R
```{r}
# passes per item, easy (left) to hard (right)
tibble(item = colnames(X), passes = colSums(X))
```
### SPSS
```
* Passes per item, easy (left) to hard (right). Everyone passes the first
* three and no one passes the last - those carry no information about
* DIFFERENCES, and cannot get a finite logit.
DESCRIPTIVES VARIABLES=i1 TO i18 /STATISTICS=SUM.
```
### Julia
```julia
[(taps[j], sum(X[:, j])) for j in 1:size(X, 2)]
```
### Python
```python
dict(zip(taps, X.sum(axis=0).astype(int)))
```
:::

Everyone passes the first three items and **no one** passes the last. Those items carry no information about *differences* - and you cannot compute a finite logit for a perfect or zero score (it is $\log$ of $\infty$). So the calibration engine will set them aside automatically. Watch for that in a moment: 18 items go in, but only the estimable ones come out.

## The Engine: Joint Maximum Likelihood (What BIGSTEPS Does)

BIGSTEPS estimates every ability and every difficulty at once by **joint maximum likelihood** (Wright's "UCON"). The recipe is a beautifully simple loop of Newton steps:

1.  Start each person's ability and each item's difficulty from its raw score (a log-odds).
2.  **Nudge each ability** so the person's *expected* score matches their *observed* score.
3.  **Nudge each difficulty** the same way, using the item's score.
4.  Re-center the item difficulties (the scale needs an origin; convention puts it at the mean item).
5.  Repeat until nothing moves.

Every "nudge" is a one-step Newton update: (observed − expected) divided by the information (the sum of $P(1-P)$). Here it is, whole:

::: {.panel-tabset}
### R
```{r}
rasch_jmle <- function(X, tol = 1e-6, maxit = 300) {
  # BIGSTEPS drops persons/items with perfect or zero scores (no finite estimate)
  repeat {
    ci <- colSums(X); ri <- rowSums(X)
    keepI <- ci > 0 & ci < nrow(X); keepP <- ri > 0 & ri < ncol(X)
    if (all(keepI) && all(keepP)) break
    X <- X[keepP, keepI, drop = FALSE]
  }
  N <- nrow(X); L <- ncol(X)
  r <- rowSums(X); s <- colSums(X)
  b <- log(r / (L - r))                     # person abilities, from raw scores
  d <- log((N - s) / s); d <- d - mean(d)   # item difficulties, centered

  # This outer loop must stay a loop: each sweep starts from the estimates the
  # previous sweep produced, so the steps genuinely depend on one another.
  # The two *inner* sweeps do not - within a sweep, every person's update uses
  # only their own ability and the current difficulties - so those map cleanly.
  for (it in 1:maxit) {
    d0 <- d
    b <- map2_dbl(b, r, \(bv, rv) {          # nudge each ability
      Pv <- plogis(bv - d)
      bv + (rv - sum(Pv)) / sum(Pv * (1 - Pv))
    })
    d <- map2_dbl(d, s, \(di, si) {          # nudge each difficulty
      Pi <- plogis(b - di)
      di - (si - sum(Pi)) / sum(Pi * (1 - Pi))
    })
    d <- d - mean(d)                         # fix the origin at the mean item
    if (max(abs(d - d0)) < tol) break
  }
  list(difficulty = d, ability = b, items = colnames(X), iterations = it)
}

fit <- rasch_jmle(X)
tibble(item = fit$items, difficulty = fit$difficulty) |> round2()
```
### SPSS
```
* Base SPSS has no Rasch or IRT procedure. Item calibration needs a
* dedicated engine - Winsteps/BIGSTEPS (Wright's own), ConQuest, or R's
* eRm/TAM - or the SPSS Python/R integration plug-in calling one of those.
*
* * The joint-maximum-likelihood calibration below
* is shown in the R, Julia, and Python tabs.
```
### Julia
```julia
using Statistics
σ(z) = 1 / (1 + exp(-z))

# Wright's Knox Cube Test data: skip the header, drop the Name & sex columns.
rows = split.(readlines("data/knox.dat")[2:end])
X = [parse(Int, rows[i][j]) for i in 1:length(rows), j in 3:20]

function rasch_jmle(X; tol=1e-6, maxit=300)
    X = float.(X)
    while true                                    # drop perfect/zero rows & cols
        ci = vec(sum(X, dims=1)); ri = vec(sum(X, dims=2))
        keepI = (ci .> 0) .& (ci .< size(X, 1))
        keepP = (ri .> 0) .& (ri .< size(X, 2))
        (all(keepI) && all(keepP)) && break
        X = X[keepP, keepI]
    end
    N, L = size(X)
    r = vec(sum(X, dims=2)); s = vec(sum(X, dims=1))
    b = log.(r ./ (L .- r))                       # abilities from raw scores
    d = log.((N .- s) ./ s); d .-= mean(d)        # difficulties, centered
    for _ in 1:maxit
        d0 = copy(d)
        for v in 1:N
            Pv = σ.(b[v] .- d)
            b[v] += (r[v] - sum(Pv)) / sum(Pv .* (1 .- Pv))
        end
        for i in 1:L
            Pi = σ.(b .- d[i])
            d[i] -= (s[i] - sum(Pi)) / sum(Pi .* (1 .- Pi))
        end
        d .-= mean(d)                             # fix the origin
        maximum(abs.(d .- d0)) < tol && break
    end
    return d, b
end

difficulty, ability = rasch_jmle(X)
```
### Python
```python
import numpy as np

# Wright's Knox Cube Test data: skip the header, drop the Name & sex columns.
rows = [ln.split() for ln in open("data/knox.dat")][1:]
X = np.array([[int(v) for v in r[2:]] for r in rows], float)

def rasch_jmle(X, tol=1e-6, maxit=300):
    X = np.asarray(X, float)
    while True:                                   # drop perfect/zero rows & cols
        ci, ri = X.sum(0), X.sum(1)
        keepI = (ci > 0) & (ci < X.shape[0])
        keepP = (ri > 0) & (ri < X.shape[1])
        if keepI.all() and keepP.all():
            break
        X = X[keepP][:, keepI]
    N, L = X.shape
    r, s = X.sum(1), X.sum(0)
    b = np.log(r / (L - r))                       # abilities from raw scores
    d = np.log((N - s) / s); d -= d.mean()        # difficulties, centered
    for _ in range(maxit):
        d0 = d.copy()
        for v in range(N):
            Pv = 1 / (1 + np.exp(-(b[v] - d)))
            b[v] += (r[v] - Pv.sum()) / (Pv * (1 - Pv)).sum()
        for i in range(L):
            Pi = 1 / (1 + np.exp(-(b - d[i])))
            d[i] -= (s[i] - Pi.sum()) / (Pi * (1 - Pi)).sum()
        d -= d.mean()                             # fix the origin
        if np.max(np.abs(d - d0)) < tol:
            break
    return d, b

difficulty, ability = rasch_jmle(X)
```
:::

Read that difficulty column top to bottom: the short tapping sequences sit low (easy), the long ones sit high (hard), climbing steadily just as Wright's KCT does. **We just calibrated a test the way BIGSTEPS calibrates a test** - in thirty lines, and every number rebuildable by hand. The persons are measured on the very same logit scale, so a child's ability and an item's difficulty are directly comparable: if $\theta_v > \delta_i$, the odds favor a correct tap.

::: {.panel-tabset}
### R
```{r}
cat("Converged in", fit$iterations, "iterations.\n")

# the 35 children, measured in logits on the item scale
tibble(ability = fit$ability) |>
  summarise(
    n      = n(),
    min    = min(ability),
    median = median(ability),
    mean   = mean(ability),
    max    = max(ability)
  ) |>
  round2()
```
### SPSS
```
* Base SPSS has no Rasch or IRT procedure. Item calibration needs a
* dedicated engine - Winsteps/BIGSTEPS (Wright's own), ConQuest, or R's
* eRm/TAM - or the SPSS Python/R integration plug-in calling one of those.
*
* * The person measures (abilities) summarised below
* is shown in the R, Julia, and Python tabs.
```
### Julia
```julia
using Statistics
println("Converged in ", fit.iterations, " iterations.")

# the 35 children, measured in logits on the item scale
length(fit.ability), minimum(fit.ability), median(fit.ability),
mean(fit.ability), maximum(fit.ability)
```
### Python
```python
print(f"Converged in {iterations} iterations.")

# the 35 children, measured in logits on the item scale
import numpy as np
len(ability), ability.min(), np.median(ability), ability.mean(), ability.max()
```
:::

Click through those tabs and read them side by side. Same loop, same updates, same convergence, same estimates - only the punctuation changes. That is the whole point Wright kept making: **the measurement is in the mathematics, not the software.** BIGSTEPS was never magic; it was this loop, run carefully.

::: {.callout-important}
## What the Real Tools Add

Our thirty lines calibrate items and measure persons - the heart of BIGSTEPS. Production tools add three things worth knowing about: (1) **fit statistics** (Wright's *infit* and *outfit* mean-squares) that flag items and people who do not behave as the model predicts; (2) better small-sample bias corrections than raw JMLE; and (3) standard errors for every estimate. When you do this for real, reach for a maintained package - **`eRm`** or **`TAM`** in R, **`py-irt`** in Python, or a Julia IRT package - all of which are, underneath, doing what you just watched.
:::

::: {.callout-note}
## From the Field
This is not a museum piece. The authors have run this exact machinery on real clinical measures - including, fittingly, a paper co-written with **Ben Wright himself**: Conrad, Wright, McKnight, and colleagues [@ConradWrightMcKnight2004] used Rasch analysis on the Mississippi PTSD Scale and found that five of its seven worst-fitting items were *reverse-scored* - the model exposing a flaw a sum score would have hidden. In another study, McKnight and Babcock-Parziale [@McKnightBabcock2007] paired Rasch with generalizability theory to evaluate a vision-rehabilitation outcome. Rasch is not exotic; it is a working tool for anyone who needs a measure to behave.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself

1.  Wright's data already hands you the lesson: items `1-4`, `2-3`, `1-2-4` were passed by all 35 children and `4-1-3-4-2-1-4` by none. Confirm from `colSums(X)` that these are exactly the items the estimator dropped, and explain in one sentence why no finite difficulty exists for them.
2.  Plot person ability against raw score. What shape is it, and why does every child with the same score (there are several with 10) get the *same* measure?
3.  Translate one of the two `for` loops (ability *or* difficulty updates) into the language you know best, and check it against the R output. You will have written a piece of BIGSTEPS.
:::

## Where We Go Next

Wright's dream - objective, sample-free measurement - is the aspiration behind every scale you will ever build or trust. With reliable, valid, *well-measured* variables in hand, we are almost ready to model relationships among them. One preparation step remains: combining our many measures into fewer, more dependable ones. That is [data reduction](31-datareduction.qmd), the subject of the next part, and it is where we turn before the modeling begins. Ben Wright would tell you: get the measurement right first, and the models take care of themselves.
