# Coding Categorical Predictors {#coding}

In the [ANOVA and the GLM](16-ANOVAandGLM.qmd) chapter we let R turn a factor into numbers for us and moved on. That was a simplification worth revisiting. R *always* codes a categorical predictor into numbers before it fits anything, and **the code you choose decides what every coefficient means.** Change the code and the intercept, the slopes, and the *t*-tests all change their meaning, even though the model's fit (its $R^2$, its $F$, its predicted values) does not change at all.

That is the whole chapter in one idea. The coding scheme is a lens, not a model. Pick the lens that answers your question, and the coefficients read like plain English. Pick a careless lens and you get numbers that are technically correct but not meaningful, what we will call **nonsense coding**.

## Learning Objectives

1.  See that a categorical predictor is just a continuous variable we chopped into bins - and that "how we chop" is a choice.
2.  Interpret **dummy** (reference), **effects** (deviation), **weighted effects**, and **contrast** codes - reading each coefficient straight off the group means.
3.  Recover every coefficient *by hand* from the marginal means, so no `lm()` output is ever a mystery.
4.  Understand why the omnibus fit is invariant to coding, and why bad codes give "nonsense" coefficients.

## A Category Is a Binned Continuum

Let's build a categorical variable from scratch so we can see the seams. Take a continuous "dogness" score, chop it into four bins, and call them `cat`, `fish`, `pig`, `dog`:

```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
library(broom)        # tidy(), glance(), augment() for model output
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(725)
n <- 200
d <- tibble(x = rnorm(n)) |>            # a continuous "dogness" score
  mutate(
    pet   = cut(x, 4, labels = c("cat", "fish", "pig", "dog")),
    happy = x + rnorm(n)                # more dogness -> happier, on average
  )

d |> count(pet)
```

`pet` is now a **factor**: R stores it as integer codes with labels attached. The moment we put it in a model, R replaces those four labels with three columns of numbers (four groups need $4-1=3$ columns, plus the intercept). *Which* three columns is the coding scheme. Here are the group means we will spend the rest of the chapter reconstructing:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch18-pets`, `ch18-unbalanced`. For example, `book_data("ch18-pets")` in R, Julia, or Python, or `!bookdata name = "ch18-pets".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.

The real four-group data is `data/mod3data.csv`.
:::

::: {.panel-tabset}
### R
```{r}
means <- d |>
  summarise(happy = mean(happy), .by = pet) |>
  arrange(pet)

means |> round2()
```
### SPSS
```
* Group means of happy by pet.
MEANS TABLES=happy BY pet
  /CELLS=MEAN STDDEV COUNT.
```
### Julia
```julia
using Statistics, DataFrames
combine(groupby(df, :pet), :happy => mean)
```
### Python
```python
df.groupby("pet")["happy"].mean()
```
:::

## Dummy (Reference) Coding

Dummy coding - R's default, also called *reference* or *indicator* coding - picks one group as the **reference** and compares every other group to it. Let's make `dog` the reference (the 4th level):

::: {.panel-tabset}
### R
```{r}
contrasts(d$pet) <- contr.treatment(4, base = 4)   # reference = level 4 = dog
contrasts(d$pet)

m_dummy <- lm(happy ~ pet, data = d)
tidy2(m_dummy)
```
### SPSS
```
* Dummy / reference coding, reference = dog (4th category).
UNIANOVA happy BY pet
  /CONTRAST(pet)=SIMPLE(4)
  /PRINT=PARAMETER.
```
### Julia
```julia
using GLM, StatsModels
m_dummy = lm(@formula(happy ~ pet), df,
             contrasts = Dict(:pet => DummyCoding(base = "dog")))
```
### Python
```python
import statsmodels.formula.api as smf
m_dummy = smf.ols("happy ~ C(pet, Treatment(reference='dog'))", data=df).fit()
m_dummy.params.round(3)
```
:::

Read the coefficients straight off the means:

- **`(Intercept)`** is the reference-group mean - the mean happiness of `dog`.
- **each slope** is that group's mean *minus* the reference mean: `pet1` is $\bar{y}_{cat} - \bar{y}_{dog}$, `pet2` is $\bar{y}_{fish} - \bar{y}_{dog}$, and so on.

Don't take my word for it - check:

```{r}
mu    <- means |> pull(happy, name = pet)          # named group means
coefs <- tidy(m_dummy) |> pull(estimate, name = term)

c(intercept_is_dog_mean      = all.equal(coefs[["(Intercept)"]], mu[["dog"]]),
  cat_slope_is_cat_minus_dog = all.equal(coefs[["pet1"]], mu[["cat"]] - mu[["dog"]]))
```

Both `TRUE`. Every dummy coefficient is a **simple difference from the reference**. That is exactly why dummy coding is the right lens when you have a control group and want each treatment compared to it.

::: {.callout-warning}
## The interaction trap
Add a second predictor *with an interaction* and a dummy slope is no longer "group mean minus reference mean" overall - it is that difference **only at the reference level of the other predictor** (i.e., where the other predictor's dummy is 0). This is the single most common misreading of regression output. When an interaction is in the model, stop reading coefficients off marginal means and solve the full equation instead. (This bites hardest in the factorial models of [ANOVA & the GLM](16-ANOVAandGLM.qmd).)
:::

## Effects (Deviation) Coding

Dummy coding asks "how does each group differ from *one* group?" **Effects coding** (a.k.a. deviation coding, `contr.sum`) asks a different question: "how does each group differ from the **grand mean**?" The zero point moves from a reference group to the center of everything.

::: {.panel-tabset}
### R
```{r}
contrasts(d$pet) <- contr.sum(4)
contrasts(d$pet)

m_eff <- lm(happy ~ pet, data = d)
tidy2(m_eff)
```
### SPSS
```
* Effects / deviation coding (each group vs the grand mean).
UNIANOVA happy BY pet
  /CONTRAST(pet)=DEVIATION
  /PRINT=PARAMETER.
```
### Julia
```julia
using GLM, StatsModels
m_eff = lm(@formula(happy ~ pet), df,
           contrasts = Dict(:pet => EffectsCoding()))
```
### Python
```python
import statsmodels.formula.api as smf
m_eff = smf.ols("happy ~ C(pet, Sum)", data=df).fit()   # each group vs the grand mean
m_eff.params.round(3)
```
:::

Now the meanings shift:

- **`(Intercept)`** is the **grand mean** - but note *which* grand mean: the mean of the four **group means**, not the mean of the raw data (those differ when groups are unequal in size).
- **each slope** is a group's mean *minus* the grand mean - its "effect."
- **the last group** (`dog`) has no coefficient of its own. Because effects sum to zero, its effect is minus the sum of the others.

Watch the grand-mean subtlety, because it trips everyone:

```{r}
GM_groups <- mean(means$happy)     # mean of the GROUP means
GM_raw    <- mean(d$happy)         # mean of the raw data

tibble(
  intercept            = tidy(m_eff)$estimate[1],
  grand_mean_of_groups = GM_groups,   # this one matches
  grand_mean_of_raw    = GM_raw       # this one does not
) |>
  round2()
```

And recover the invisible fourth effect - `dog` - as minus the sum of the rest:

```{r}
eff <- tidy(m_eff) |> slice(-1) |> pull(estimate)  # pet1, pet2, pet3 = cat, fish, pig
dog_effect <- -sum(eff)

all.equal(mu[["dog"]] - GM_groups, dog_effect)
```

`TRUE`. That is the whole trick of effects coding: the coefficients are **deviations that must sum to zero**, so the omitted group is always recoverable.

::: {.callout-important}
## Dummy vs. effects, in one line
**Dummy codes make the zero *explicit* (it is the reference group); effects codes make the zero *implicit* (it is the grand mean, sitting between the deviations).** That is the mnemonic to carry out of this chapter.
:::

::: {.callout-warning}
## Always check your contrasts
`contr.sum` on a two-level factor assigns `+1` to the *first* level and `-1` to the second, which is often the opposite of what you expect, and it silently flips the sign of your effect. R's default level ordering (alphabetical) can catch you the same way. Print `contrasts(yourfactor)` *before* you interpret anything.
:::

## Weighted Effects Coding

Plain effects coding centers on the mean of the group means, treating a group of 10 and a group of 200 as equals. When your cells are unbalanced and you want the **population** grand mean (weighted by how many people are actually in each group), you need **weighted effects coding**.

Let's make an unbalanced world - many pigs, few dogs - and build the weighted codes by hand, pushing the cell-size ratios onto the comparison group's row:

```{r}
set.seed(9)
cnt  <- c(cat = 20, fish = 30, pig = 40, dog = 10)
mu_w <- c(cat = 0, fish = 1, pig = 2, dog = 3)   # group means, so the demo has structure

dw <- tibble(petw = factor(rep(names(cnt), cnt), levels = names(cnt))) |>
  mutate(happyw = rnorm(n(), mean = mu_w[as.character(petw)], sd = 1))

contrasts(dw$petw) <- contr.sum(4)   # start from unweighted effects codes
M <- contrasts(dw$petw)
N <- dw |> count(petw) |> pull(n)     # cell sizes, in level order

# reweight the comparison (last) group's row by the cell-size ratios:
M[4, ] <- M[4, ] * N[1:3] / N[4]
contrasts(dw$petw) <- M

M |> as_tibble(rownames = "pet") |> round2()

m_weff <- lm(happyw ~ petw, data = dw)
```

The payoff: the intercept is now the **sample** grand mean - the ordinary mean of the raw data, weighted by cell size - not the mean of the group means:

```{r}
tibble(
  intercept   = tidy(m_weff)$estimate[1],
  raw_mean    = mean(dw$happyw),    # the weighted-effects intercept matches THIS
  groups_mean = dw |> summarise(m = mean(happyw), .by = petw) |> pull(m) |> mean()
) |>
  round2()
```

The intercept tracks the raw mean. The weights get pushed onto the comparison-group means, so unbalanced data no longer distorts your grand mean.

## Contrast Coding

Dummy and effects codes answer fixed questions. **Contrast codes** let *you* pose the question - "is the average of these two groups different from the average of those two?" - by writing the comparison directly into the code columns. Done right (orthogonally), each coefficient is one clean, independent comparison.

### Helmert contrasts

R ships one useful family, **Helmert** codes, which compare each level to the *pooled mean of the levels before it*:

::: {.panel-tabset}
### R
```{r}
contrasts(d$pet) <- contr.helmert(4)

m_helm <- lm(happy ~ pet, data = d)
tidy2(m_helm)
```
### SPSS
```
* Helmert contrasts.
UNIANOVA happy BY pet
  /CONTRAST(pet)=HELMERT
  /PRINT=PARAMETER.
```
### Julia
```julia
using GLM, StatsModels
m_helm = lm(@formula(happy ~ pet), df,
            contrasts = Dict(:pet => HelmertCoding()))
```
### Python
```python
import statsmodels.formula.api as smf
# Helmert: each level against the mean of the levels before it.
m_helm = smf.ols("happy ~ C(pet, Helmert)", data=df).fit()
m_helm.params.round(3)
```
:::

The intercept is again the (unweighted) grand mean. The first Helmert slope is *half* the gap between the first two group means - why half? Because the code uses $-1$ and $+1$, a two-unit spread, so the slope is the difference divided by 2:

```{r}
tibble(
  helmert_b1   = tidy(m_helm)$estimate[2],
  half_the_gap = (mu[["fish"]] - mu[["cat"]]) / 2
) |>
  round2()
```

### Custom orthogonal contrasts

The real power is writing your own. Suppose the scientific questions are: (1) are the "usual" pets (`cat`, `fish`) different from the "unusual" ones (`pig`, `dog`)? (2) `cat` vs. `fish`? (3) `pig` vs. `dog`? Encode exactly those, and **name the columns** so the output documents itself:

::: {.panel-tabset}
### R
```{r}
sgc <- matrix(c( 0.5,  0.5, -0.5, -0.5,   # usual (cat,fish) vs unusual (pig,dog)
                 0.5, -0.5,  0.0,  0.0,   # cat vs fish
                 0.0,  0.0,  0.5, -0.5),  # pig vs dog
              nrow = 4,
              dimnames = list(levels(d$pet),
                              c("usual_vs_unusual", "cat_vs_fish", "pig_vs_dog")))

sgc |> as_tibble(rownames = "pet") |> round2()

contrasts(d$pet) <- sgc
m_con <- lm(happy ~ pet, data = d)
tidy2(m_con)
```
### SPSS
```
* Custom orthogonal contrasts:
*   usual (cat,fish) vs unusual (pig,dog); cat vs fish; pig vs dog.
UNIANOVA happy BY pet
  /CONTRAST(pet)=SPECIAL(0.5 0.5 -0.5 -0.5  0.5 -0.5 0 0  0 0 0.5 -0.5)
  /PRINT=PARAMETER.
```
### Julia
```julia
using GLM, StatsModels
m_con = lm(@formula(happy ~ pet), df,
           contrasts = Dict(:pet => HypothesisCoding(
               [0.5 0.5 -0.5 -0.5; 0.5 -0.5 0 0; 0 0 0.5 -0.5])))
```
### Python
```python
import numpy as np
import statsmodels.formula.api as smf
from patsy.contrasts import ContrastMatrix

# rows = cat, fish, pig, dog; columns = the three contrasts
sgc = ContrastMatrix(np.array([[ 0.5,  0.5,  0.0],
                               [ 0.5, -0.5,  0.0],
                               [-0.5,  0.0,  0.5],
                               [-0.5,  0.0, -0.5]]),
                     ["_usual_vs_unusual", "_cat_vs_fish", "_pig_vs_dog"])
m_con = smf.ols("happy ~ C(pet, sgc)", data=df).fit()
m_con.params.round(3)
```
:::

These three contrasts are **orthogonal** (each pair of columns is uncorrelated), so the three comparisons are independent - each coefficient answers its question without contamination from the others. Confirm the first one really is the usual-vs-unusual gap:

```{r}
usual   <- mean(mu[c("cat", "fish")])
unusual <- mean(mu[c("pig", "dog")])

tibble(
  contrast_b1 = tidy(m_con) |>
    filter(term == "petusual_vs_unusual") |> pull(estimate),
  usual_minus_unusual = usual - unusual
) |>
  round2()
```

::: {.callout-important}
## The rule to remember
**Effects and contrast codes must be centered to move between means and coefficients** - so you divide by the right denominator (the spread of the codes) when you compute a $b$ from marginal means, and multiply back when you go the other way. Miss the denominator and your hand-check will be "off by a factor" and you'll think you broke something. You didn't; you forgot to center.
:::

## Nonsense Coding

Now the capstone, and the point of the whole chapter. What if we assign *arbitrary* numbers to the groups, with no reference, no deviations, and no meaningful comparison behind them?

::: {.panel-tabset}
### R
```{r}
nonsense <- matrix(c( 3.0, -7.0,  2.5,  0.1,
                      1.0,  1.0, -2.0,  4.0,
                     -5.0,  2.0,  8.0, -1.0),
                   nrow = 4,
                   dimnames = list(levels(d$pet), c("huh", "wat", "lol")))

contrasts(d$pet) <- nonsense
m_nonsense <- lm(happy ~ pet, data = d)
tidy2(m_nonsense)
```
### SPSS
```
* Nonsense coding: arbitrary contrast coefficients with no meaning.
UNIANOVA happy BY pet
  /CONTRAST(pet)=SPECIAL(3 -7 2.5 0.1  1 1 -2 4  -5 2 8 -1)
  /PRINT=PARAMETER.
```
### Julia
```julia
using GLM, StatsModels
m_nonsense = lm(@formula(happy ~ pet), df,
                contrasts = Dict(:pet => HypothesisCoding(
                    [3 -7 2.5 0.1; 1 1 -2 4; -5 2 8 -1])))
```
### Python
```python
import numpy as np
import statsmodels.formula.api as smf
from patsy.contrasts import ContrastMatrix

# Nonsense coding: arbitrary contrast coefficients with no meaning.
nonsense = ContrastMatrix(np.array([[ 3.0,  1.0, -5.0],
                                    [-7.0,  1.0,  2.0],
                                    [ 2.5, -2.0,  8.0],
                                    [ 0.1,  4.0, -1.0]]),
                          ["_huh", "_wat", "_lol"])
m_nonsense = smf.ols("happy ~ C(pet, nonsense)", data=df).fit()
m_nonsense.params.round(3)
```
:::

Those coefficients are not interpretable, which is exactly why the columns are named `huh`, `wat`, and `lol`. You cannot read a single one off the group means. And yet look at what has **not** changed:

```{r}
# The model fit is byte-for-byte identical to the sensible codings:
c(same_fitted = isTRUE(all.equal(augment(m_nonsense)$.fitted,
                                 augment(m_dummy)$.fitted)),
  same_r2     = all.equal(glance(m_nonsense)$r.squared, glance(m_dummy)$r.squared),
  same_F      = all.equal(glance(m_nonsense)$statistic, glance(m_dummy)$statistic))
```

All `TRUE`. **Every coding scheme in this chapter fits the exact same model** - same predicted values, same $R^2$, same omnibus $F$. Dummy, effects, weighted, contrast, and nonsense codes are five different *descriptions* of one fitted surface. The omnibus test ("does `pet` matter at all?") is blind to coding. Only the **individual coefficients** - the intercept and the $b$'s - carry the fingerprint of the code you chose.

That is the lesson worth holding onto:

::: {.callout-important}
## Coding does not change the model - only the meaning of the coefficients
The data decide the fit. *You* decide, through the coding scheme, which questions the coefficients answer. Choose codes whose contrasts match your scientific question and the output reads like a sentence. Choose thoughtlessly and you get nonsense coding: correct arithmetic, meaningless numbers.
:::

## Summary of All Schemes

| Scheme | R helper | Intercept means | A slope means | Use it when |
|---|---|---|---|---|
| Dummy / reference | `contr.treatment` | reference-group mean | group − reference | you have a control/baseline group |
| Effects / deviation | `contr.sum` | grand mean (of group means) | group − grand mean | groups are balanced; you want deviations |
| Weighted effects | hand-built from `contr.sum` | sample grand mean (weighted) | group − weighted grand mean | groups are **unbalanced** |
| Contrast (Helmert/custom) | `contr.helmert` / your matrix | grand mean | one planned comparison | you have specific, orthogonal questions |
| Nonsense | arbitrary matrix | nothing interpretable | nothing interpretable | never (it's the cautionary tale) |

Across every row the $R^2$, $F$, and fitted values are identical. The columns "Intercept means" and "A slope means" are the only things that move.

## A Real Example: Data From the GLM Course

Simulated pets are friendly, but let's prove the machinery on real data - the Module 3 dataset from the graduate GLM course (`data/mod3data.csv`). It has an outcome `y1` and a genuine four-level grouping factor, `f3`, with unequal cells (real data is always lumpy):

::: {.panel-tabset}
### R
```{r}
d3 <- read_csv("data/mod3data.csv", show_col_types = FALSE) |>
  mutate(grp = factor(f3, labels = c("A", "B", "C", "D")))

group3 <- d3 |>
  summarise(mean = mean(y1), n = n(), .by = grp) |>
  arrange(grp)

group3 |> round2()     # the four real group means, and unequal cell sizes

mu3 <- group3 |> pull(mean, name = grp)
```
### SPSS
```
* Group means and cell sizes of y1 by grp.
MEANS TABLES=y1 BY grp
  /CELLS=MEAN STDDEV COUNT.
```
### Julia
```julia
using Statistics, DataFrames
combine(groupby(df, :grp), :y1 => mean)
```
### Python
```python
import pandas as pd
d3 = pd.read_csv("data/mod3data.csv")
d3["grp"] = pd.Categorical(d3.f3).rename_categories(["A", "B", "C", "D"])
d3.groupby("grp")["y1"].agg(["mean", "size"])   # means and unequal cell sizes
```
:::

Dummy-code it (reference = group A) and read the coefficients straight off those means, exactly as we did for the pets:

::: {.panel-tabset}
### R
```{r}
contrasts(d3$grp) <- contr.treatment(4, base = 1)

m_real <- lm(y1 ~ grp, data = d3)
tidy2(m_real)

cr <- tidy(m_real) |> pull(estimate, name = term)
c(intercept_is_A    = all.equal(cr[["(Intercept)"]], mu3[["A"]]),
  grp2_is_B_minus_A = all.equal(cr[["grp2"]], mu3[["B"]] - mu3[["A"]]))
```
### SPSS
```
* Dummy / reference coding, reference = group A (1st category).
UNIANOVA y1 BY grp
  /CONTRAST(grp)=SIMPLE(1)
  /PRINT=PARAMETER.
```
### Julia
```julia
using GLM, StatsModels
m_real = lm(@formula(y1 ~ grp), df,
            contrasts = Dict(:grp => DummyCoding(base = "A")))
```
### Python
```python
import statsmodels.formula.api as smf
m_real = smf.ols("y1 ~ C(grp, Treatment(reference='A'))", data=d3).fit()
m_real.params.round(3)         # intercept = mean of A; next term = B - A
```
:::

Both `TRUE` - the intercept is group A's mean, each slope a group's mean minus A's. Switch to effects coding and the intercept becomes the grand mean of the four group means, unequal cells and all:

::: {.panel-tabset}
### R
```{r}
contrasts(d3$grp) <- contr.sum(4)

m_real_eff <- lm(y1 ~ grp, data = d3)
all.equal(tidy(m_real_eff)$estimate[1], mean(mu3))
```
### SPSS
```
* Effects / deviation coding (each group vs the grand mean).
UNIANOVA y1 BY grp
  /CONTRAST(grp)=DEVIATION
  /PRINT=PARAMETER.
```
### Julia
```julia
using GLM, StatsModels
m_real_eff = lm(@formula(y1 ~ grp), df,
                contrasts = Dict(:grp => EffectsCoding()))
```
### Python
```python
import statsmodels.formula.api as smf
m_real_eff = smf.ols("y1 ~ C(grp, Sum)", data=d3).fit()
m_real_eff.params[0]           # intercept = unweighted mean of the four means
```
:::

`TRUE`. Real data, real (unbalanced) cells, and every rule from the pet example still holds, because the arithmetic works the same whether the groups are house pets or the conditions of an actual study.

## Challenge

::: {.callout-tip}
## Do One Yourself
1.  Refit the four-group model with `contr.treatment` using `cat` as the reference instead of `dog`. Which coefficients change sign, and why is $R^2$ untouched?
2.  Build a **balanced** version of the pet data (equal cell sizes) and show that weighted and unweighted effects coding then give the *same* intercept. Explain in one sentence why the weights stop mattering.
3.  Invent your own orthogonal contrast set for four groups (check orthogonality with `crossprod(yourmatrix)` - the off-diagonals should be 0). Fit it, and reconstruct one coefficient from the group means, remembering to divide by the code spread.
:::

## Where We Go Next

Coding is where categorical data enter the linear model, and where careful thought pays off in coefficients you can actually explain. It is also the last piece of the *observed-variable* linear model. But the constructs we most want to study, such as depression, curiosity, and intelligence, are never observed directly; we see only their noisy indicators. The [next part](28-sem.qmd) turns to **latent variables** and the factor and structural models built for them, and after that, to what you do when the normal curve no longer describes your data well.
