# Beyond the Normal Curve: Resampling, Robust, and Bayesian Statistics {#beyond}

Every method in this book so far has quietly leaned on three crutches: the **normal curve** (errors are Gaussian), a **point estimate** (one number for the answer), and a **reject / fail-to-reject** verdict (the $p$-value ritual). Those crutches work astonishingly well when the world cooperates. This chapter is about what to do when it doesn't - when your sample is small, your distribution is lumpy, a few wild points hijack the average, or you simply want a more honest account of what you know.

We meet three families of methods: **resampling** (let the data be their own population), **robust** statistics (estimators a few bad points can't bully), and **Bayesian** inference (probability of the *model given the data*, not the data given the model). They are not three unrelated tricks. They share a spine, and naming that spine is the most useful thing this chapter can do for you.

::: {.callout-important}
## The distinction that runs through everything: estimation vs. testing
Two different questions hide inside "doing statistics":

- **Estimation** - *what is the value, and how sure am I?* A number **and its uncertainty**: a slope and its interval, an effect and its spread.
- **Testing** - *is it different from some reference, or which model is better?* A decision or a comparison.

Classical training fuses these into one motion (compute $t$, read $p$, declare significance). The methods in this chapter pull them back apart - and, strikingly, all three do **testing through estimation**: they estimate the whole distribution of the answer first, then let the *test* fall out of it (does the interval exclude zero? how much posterior sits near the null?). Keep asking, of every method below: *is this estimating, or testing - and how does it get from one to the other?*
:::

## Learning Objectives

1.  Use **resampling** (bootstrap, permutation, jackknife) to get standard errors, intervals, and tests without a normal-curve assumption.
2.  Use **robust** estimators (median, trimmed mean, MAD, M-estimators) that a handful of outliers cannot capture.
3.  Read a **Bayesian** analysis - prior, likelihood, posterior - and report it as estimation (credible interval) and as testing (ROPE, model comparison).
4.  Draw the estimation-vs-testing distinction cleanly in each framework.

## Part 1 · Resampling: Let the Data Be Their Own Population

The logic is almost too simple to trust. A statistic - a mean, a slope, an indirect effect - has a **sampling distribution**: the spread of values it would take across repeated samples from the population. Classical theory *derives* that spread from formulas that assume normality and large $n$. **Resampling estimates it directly**, by treating your one sample as a stand-in for the population and drawing new samples from it.

> When data sets are small, the usual methods... may be inappropriate (because of distributional characteristics) or insensitive. Bootstrapping, jackknifing and other methods of resampling may be helpful. - McKnight, McKnight & Figueredo, *Australasian Evaluation Society*, 1997

### The bootstrap (estimation)

**Bootstrapping** resamples your data *with replacement*, `n` observations drawn from `n`, over and over, recomputing the statistic each time. The spread of those recomputed values *is* the estimated sampling distribution. Take a small, skewed sample - exactly where the normal-theory formula is least trustworthy - and bootstrap the mean:

```{r}
set.seed(42)
samp <- rexp(30, rate = 1/50)          # n = 30, right-skewed (non-normal) data

B <- 2000
boot_means <- replicate(B, mean(sample(samp, replace = TRUE)))   # resample & recompute

c(estimate = mean(samp),
  boot_SE  = sd(boot_means),                       # bootstrap standard error
  lower95  = quantile(boot_means, 0.025),          # percentile confidence interval
  upper95  = quantile(boot_means, 0.975))
```

That is the entire idea: `sample(samp, replace = TRUE)` inside `replicate()`, then read the **standard error** off `sd()` and the **confidence interval** off `quantile()`. No formula, no normality, no $z$. The 2.5th and 97.5th percentiles of the bootstrap distribution are a **percentile confidence interval** - and notice it can be *asymmetric*, honestly reflecting the skew that a normal-theory interval would paper over.

```{r}
#| fig-cap: "The bootstrap sampling distribution of the mean, with the percentile 95% interval."
#| fig-height: 4
hist(boot_means, breaks = 40, col = "grey80", border = "white",
     main = "", xlab = "resampled mean")
abline(v = quantile(boot_means, c(.025, .975)), col = "red", lwd = 2, lty = 2)
abline(v = mean(samp), col = "blue", lwd = 2)
```

### The permutation test (testing)

The bootstrap *estimated* an interval. To *test* a hypothesis - say, that two groups have the same mean - we resample differently. Under the null "the group label is irrelevant," we can **shuffle the labels** and see how big a difference arises by chance alone. The observed difference is "significant" if it lands in the tail of that shuffled distribution.

```{r}
set.seed(7)
g1 <- rnorm(15, mean = 0.0)
g2 <- rnorm(15, mean = 0.8)
obs <- mean(g2) - mean(g1)                         # the observed difference

pool   <- c(g1, g2)
labels <- rep(1:2, each = 15)
perm   <- replicate(5000, {                        # shuffle labels 5000 times
  s <- sample(labels)
  mean(pool[s == 2]) - mean(pool[s == 1])
})

c(observed_diff = obs,
  permutation_p = mean(abs(perm) >= abs(obs)))      # two-sided p-value
```

The permutation $p$-value is the fraction of shuffles that beat the real difference. It makes **no distributional assumption at all** - it is the purest possible hypothesis test, built only from the exchangeability of labels under the null.

### The jackknife (a leave-one-out cousin)

Older and simpler than the bootstrap, the **jackknife** recomputes the statistic leaving out one observation at a time. It is handy for a quick standard error and for spotting influential points (if dropping one case moves the estimate a lot, you've found a leverage problem):

```{r}
jk    <- sapply(seq_along(samp), function(i) mean(samp[-i]))       # leave-one-out means
n     <- length(samp)
jk_se <- sqrt((n - 1) * mean((jk - mean(jk))^2))                  # jackknife SE
c(jackknife_mean = mean(jk), jackknife_SE = jk_se)
```

::: {.callout-note}
## Estimation vs. testing, the resampling version
Same engine, two jobs. **Bootstrap → estimation** (an SE and a percentile interval for *whatever* statistic you can compute - a mean, a correlation, a regression slope, even a mediation indirect effect $a\cdot b$). **Permutation → testing** (a $p$-value from label shuffling). And the two meet: a bootstrap **interval that excludes zero** *is* a test. Resampling's honesty is also its discipline - it is **not** a substitute for collecting more data, and it demands a representative, random sample to begin with.
:::

## Part 2 · Robust Statistics: When a Few Points Hijack the Mean

Recall from the [central tendency chapter](03-CTmeasures.qmd) that the mean is a *least-squares* estimator: it sits where squared deviations are smallest. Squaring is exactly why it's fragile - one faraway point contributes an enormous squared deviation and drags the mean toward itself. A single bad value can move the mean anywhere; its **breakdown point** (the fraction of contamination it can tolerate before giving nonsense) is essentially $0\%$.

Watch nineteen clean observations and one fat-fingered data-entry error:

```{r}
set.seed(1)
z <- c(rnorm(19, mean = 100, sd = 10), 250)        # 19 good values + 1 gross outlier

round(c(mean       = mean(z),
        median     = median(z),
        trimmed10  = mean(z, trim = 0.10),         # drop the extreme 10% each tail
        SD         = sd(z),
        MAD        = mad(z)), 2)                    # median absolute deviation
```

The **mean** and **SD** are inflated by the lone outlier; the **median**, **10% trimmed mean**, and **MAD** barely notice it. That resistance is the whole point of robust statistics:

- **Median** - breakdown point $50\%$: half your data can be garbage before it fails.
- **Trimmed mean** - discard a fixed fraction of each tail, then average the rest. A tunable compromise between mean (0% trim) and median (50% trim).
- **MAD** - the robust analogue of the SD, a median of absolute deviations from the median.
- **M-estimators** (Huber, Tukey bisquare) - down-weight, rather than discard, extreme points, via iteratively reweighted least squares. These are the robust-regression workhorses.

For real robust modeling in R, reach for maintained tools - `MASS::rlm()` for robust regression, and Rand Wilcox's **`WRS2`** package for robust group comparisons (trimmed-means tests, robust ANOVA). The idioms look like this:

```r
library(MASS)
rlm(happy ~ dogness, data = d)          # robust regression (Huber M-estimator)

library(WRS2)
yuen(happy ~ group, data = d)           # robust two-group test on trimmed means
t1way(happy ~ group, data = d)          # robust one-way "ANOVA"
```

::: {.callout-note}
## Estimation vs. testing, the robust version
**Estimation:** report a robust *center and spread* - a trimmed mean with a MAD, or robust regression coefficients - instead of mean $\pm$ SD when the tails are heavy. **Testing:** use tests built on those robust estimators (Yuen's trimmed-means test, `t1way`) or - the pragmatic favorite in this book - **bootstrap the robust estimator** and check whether its interval excludes the null. Robustness and resampling are natural allies: diagnose the violation, then resample the resistant estimator.
:::

## Part 3 · Bayesian Inference: Probability of the Model Given the Data

Frequentist statistics computes $P(\text{data} \mid \text{hypothesis})$ - the $p$-value is the probability of data this extreme *if the null were true*. That is a strange quantity to hang careers on: it answers a question nobody asked. Bayesian inference computes what you actually wanted, $P(\text{hypothesis} \mid \text{data})$, by way of **Bayes' rule**:

$$\underbrace{P(\theta \mid \text{data})}_{\text{posterior}} \;\propto\; \underbrace{P(\text{data} \mid \theta)}_{\text{likelihood}} \times \underbrace{P(\theta)}_{\text{prior}}$$

The **prior** is what you believed about the parameter $\theta$ before the data; the **likelihood** is what the data say; the **posterior** is the updated belief. That's the whole machine. As McKnight puts it, Bayesian methods are "merely a vehicle for re-conceptualizing how we think about statistics" - they give you "insights you may already know, but didn't know you knew."

### Why it matters: the arithmetic of the replication crisis

Here is a calculation that should unsettle you, and it needs nothing but Bayes' rule. You run a well-powered study, get $p < .05$, and publish. **What is the probability your hypothesis is actually true?** Not the $p$-value - that's $P(\text{significant} \mid H \text{ false})$. The thing you want is $P(H \text{ true} \mid \text{significant})$, and it depends on how plausible your idea was *before* the data - its prior:

```{r}
power <- 0.99    # P(significant | H true)  -- a very high-powered study
alpha <- 0.01    # P(significant | H false) -- a strict false-positive rate
prior <- 0.02    # P(H true): your idea is ~1 of 50 plausible hypotheses

post <- (power * prior) / (power * prior + alpha * (1 - prior))   # Bayes' rule
round(post, 3)
```

Even with $99\%$ power and a $1\%$ false-positive rate, a significant result leaves you only about **67%** sure - because the idea was a long shot to begin with. Loosen the study to more realistic numbers and it collapses further:

```{r}
# modest power, ordinary screening, a rarer idea
round((0.95 * 0.01) / (0.95 * 0.01 + 0.05 * 0.99), 3)
```

About **16%**. Chase enough low-prior ideas with $p < .05$ and *most* of your "discoveries" will be false, no matter how small the $p$-value - which is the arithmetic engine of the replication crisis we met in [hypothesis testing](08-hypothesistesting.qmd). The $p$-value never answered the question you cared about. Bayes' rule does, but it demands you state your prior out loud. That honesty is the point.

### Building a posterior by hand (grid approximation)

You do not need fancy software to see the machine turn. Take the oldest example in the book - estimating a proportion (a coin's bias, a globe's water fraction, a treatment's success rate) from a handful of trials. Lay down a **grid** of possible values, evaluate prior × likelihood at each, normalize, and you have a posterior:

```{r}
#| fig-cap: "Grid-approximation posterior for a proportion after 6 successes in 9 trials."
#| fig-height: 4
grid  <- seq(0, 1, length.out = 1000)
prior <- dbeta(grid, 2, 2)                 # weakly informative: centered at 0.5, gentle
like  <- dbinom(6, size = 9, prob = grid)  # data: 6 successes in 9 trials
post  <- like * prior
post  <- post / sum(post)                  # normalize to a proper distribution

plot(grid, post, type = "l", lwd = 2, col = "blue",
     xlab = expression(theta ~ "(the proportion)"), ylab = "posterior density")
```

Now *summarize* the posterior - and here the estimation-vs-testing split becomes concrete.

**Estimation.** The answer is the whole posterior; we compress it into a point and an interval. Sampling from the grid makes this a one-liner. A Bayesian interval is a **credible** (or "compatibility") interval, and - unlike a confidence interval - it means the plain-English thing: "given the model and data, there's a 95% probability $\theta$ is in here."

```{r}
draws <- sample(grid, 1e5, replace = TRUE, prob = post)
c(posterior_mean   = mean(draws),
  credible_lower95 = quantile(draws, 0.025),
  credible_upper95 = quantile(draws, 0.975))
```

**Testing.** Rather than a point-null "is $\theta$ *exactly* 0.5?" (a question of measure zero that is never truly true), Bayesians ask a better one: how much posterior belief sits in a **Region of Practical Equivalence (ROPE)** around the null - a band of values "close enough to 0.5 to not care"? If almost none of the posterior is in the ROPE, the effect is credibly *not* null; if most of it is, you have evidence *for* practical equivalence - something a $p$-value can never give you.

```{r}
rope_lo <- 0.45; rope_hi <- 0.55                # "practically a fair coin"
mean(draws > rope_lo & draws < rope_hi)          # posterior probability inside the ROPE
```

### From the toy to the real thing

Grid approximation shows the logic but doesn't scale past a parameter or two. For real models, samplers (MCMC) do the integration. McKnight's teaching stack is McElreath's **`rethinking`** for concepts and **`brms`** for applied work - both compile to Stan under the hood. The *thinking* is identical to the grid above; only the engine changes:

```r
library(rethinking)
m <- quap(alist(
        y    ~ dnorm(mu, sigma),     # likelihood
        mu   <- a + b * x,
        a    ~ dnorm(0, 10),         # priors
        b    ~ dnorm(0, 1),          #   weakly informative on a standardized predictor
        sigma ~ dexp(1)
     ), data = d)
post <- extract.samples(m)
HPDI(post$b, prob = 0.89)            # highest-posterior-density (credible) interval
compare(m1, m2, func = WAIC)         # model comparison (predictive testing)
```

```r
library(brms); library(bayestestR)
fit <- brm(y ~ x, data = d, family = gaussian(),
           prior = prior(normal(0, 1), class = "b"))   # regularizing prior
hdi(fit)                             # estimation: credible intervals
p_direction(fit)                     # probability the effect has a given sign
rope(fit, range = c(-0.05, 0.05))    # testing: mass in the ROPE
loo(fit)                             # model comparison via leave-one-out
```

Two habits from McKnight's applied pipeline are worth stealing: **standardize predictors** so a `normal(0, 1)` prior is genuinely weakly-informative and interpretable, and **simulate from your priors before touching the data** to check they imply sane predictions (an over-tight prior underfits; a too-vague one overfits).

::: {.callout-note}
## Estimation vs. testing, the Bayesian version
**Estimation is primary and everything else is downstream of it.** Fit the model, get a posterior, report a **credible interval** (HPDI). "Testing" then comes in two honest flavors, neither of which is point-null rejection: **practical significance** via the **ROPE** (how much posterior is near the null?), and **model comparison** via predictive scores (**WAIC**, **PSIS-LOO**) or **Bayes factors**. A caution McKnight repeats: *predictive testing is not causal estimation* - a confounded model can out-predict the correct causal one, so "which model forecasts better" and "which parameter is right" are different questions. Do not let a good WAIC talk you out of a good DAG.
:::

## The Distinction, Woven Together

One table, three frameworks, two questions:

| Framework | **Estimation** (value + uncertainty) | **Testing** (difference / comparison) |
|---|---|---|
| Classical (normal theory) | estimate $\pm$ standard error; confidence interval | $t$/$F$ test, $p$-value |
| **Resampling** | bootstrap SE + percentile interval | permutation $p$; bootstrap interval excludes null |
| **Robust** | trimmed mean + MAD; robust regression coefficients | Yuen/`t1way`; bootstrap the robust estimator |
| **Bayesian** | posterior + credible (HPDI) interval | ROPE; model comparison (WAIC/LOO); Bayes factor |

Read down the **estimation** column and the four approaches are more alike than different: all report a best value and a spread. Read down the **testing** column and you see the modern lesson of this whole book - **testing is best done as a consequence of estimation, not as a separate ritual.** Estimate the distribution of your answer honestly, and the test is already in your hands.

## Challenge

::: {.callout-tip}
## Do One Yourself
1.  Bootstrap a statistic that has *no* simple standard-error formula - the **median**, or the **correlation** between two variables. Compare your percentile interval to what you'd get pretending it was normal.
2.  Take the outlier demo (`z`) and sweep the trimmed mean's trim fraction from 0 to 0.5. Plot the trimmed mean against the trim fraction. Where does it stop moving, and what value has it become?
3.  Re-run the grid-approximation posterior with a *strong* prior (`dbeta(grid, 20, 20)`) and a *flat* one (`dbeta(grid, 1, 1)`). How much does the prior move the posterior with only 9 trials, and with a simulated 900 trials? What does that tell you about when priors matter?
:::

## Where We Go Next

You now have three ways past the normal curve, and a single principle - estimate first, test second - that ties them together and reaches back through every chapter you've read. That principle is the quiet thesis of this book: statistics is the disciplined quantification of uncertainty, and the honest currency of uncertainty is a *distribution*, not a verdict. Wherever your data take you - a resampled interval, a robust slope, a posterior - you are doing the same human thing we started with: saying what you know, and how sure you are.

And knowing something is only half the job. The other half is *showing* it - honestly, to someone else who did not run your analysis and will not read your code. That craft, the display of data in [graphics](21-graphics.qmd) and [tables](22-tables.qmd), is where we turn next.
