# Beyond the Normal Curve: Resampling, Robust, and Bayesian Statistics {#beyond}

Every method in this book so far has quietly relied on three assumptions: the **normal curve** (errors are Gaussian), a **point estimate** (one number for the answer), and a **reject / fail-to-reject** verdict (the $p$-value ritual). Those assumptions work well when the data cooperate. This chapter is about what to do when they don't, when your sample is small, your distribution is lumpy, a few extreme points pull the average around, or you simply want a more honest account of what you know.

We meet three families of methods: **resampling** (let the data be their own population), **robust** statistics (estimators that a few bad points can't dominate), and **Bayesian** inference (probability of the *model given the data*, not the data given the model). These are not three unrelated tricks. They share a common logic, and naming that shared logic is the most useful thing this chapter can do for you.

::: {.callout-important}
## The distinction that runs through everything: estimation vs. testing
Two different questions hide inside "doing statistics":

- **Estimation** - *what is the value, and how sure am I?* A number **and its uncertainty**: a slope and its interval, an effect and its spread.
- **Testing** - *is it different from some reference, or which model is better?* A decision or a comparison.

Classical training fuses these into one motion (compute $t$, read $p$, declare significance). The methods in this chapter pull them back apart, and all three do **testing through estimation**: they estimate the whole distribution of the answer first, then let the *test* fall out of it (does the interval exclude zero? how much posterior sits near the null?). As you read each method below, keep asking: *is this estimating, or testing, and how does it get from one to the other?*
:::

## Learning Objectives

1.  Use **resampling** (bootstrap, permutation, jackknife) to get standard errors, intervals, and tests without a normal-curve assumption.
2.  Use **robust** estimators (median, trimmed mean, MAD, M-estimators) that a handful of outliers cannot capture.
3.  Read a **Bayesian** analysis - prior, likelihood, posterior - and report it as estimation (credible interval) and as testing (ROPE, model comparison).
4.  Draw the estimation-vs-testing distinction cleanly in each framework.

## Part 1 · Resampling: Let the Data Be Their Own Population

The logic is simpler than it first appears. A statistic (a mean, a slope, an indirect effect) has a **sampling distribution**: the spread of values it would take across repeated samples from the population. Classical theory *derives* that spread from formulas that assume normality and large $n$. **Resampling estimates it directly**, by treating your one sample as a stand-in for the population and drawing new samples from it.

> When data sets are small, the usual methods... may be inappropriate (because of distributional characteristics) or insensitive. Bootstrapping, jackknifing and other methods of resampling may be helpful. - McKnight, McKnight & Figueredo, *Australasian Evaluation Society*, 1997

### The bootstrap (estimation)

**Bootstrapping** resamples your data *with replacement*, `n` observations drawn from `n`, over and over, recomputing the statistic each time. The spread of those recomputed values *is* the estimated sampling distribution. Take a small, skewed sample - exactly where the normal-theory formula is least trustworthy - and bootstrap the mean:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch25-skewed-sample`, `ch25-twogroups`, `ch25-outlier`. For example, `book_data("ch25-skewed-sample")` in R, Julia, or Python, or `!bookdata name = "ch25-skewed-sample".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(42)
# n = 30, drawn from a normal that leans hard right - skewed enough that the
# normal-theory formula for a confidence interval is on shaky ground
samp <- rsnorm(30, mean = 50, sd = 20, alpha = 8)

B <- 2000
boot_means <- map_dbl(1:B, \(b) mean(sample(samp, replace = TRUE)))  # resample & recompute

tibble(
  estimate = mean(samp),
  boot_SE  = sd(boot_means),                    # bootstrap standard error
  lower95  = quantile(boot_means, 0.025),       # percentile confidence interval
  upper95  = quantile(boot_means, 0.975)
) |>
  round2()
```
### SPSS
```
* SPSS has a genuine BOOTSTRAP subcommand (Bootstrapping add-on module),
* which resamples and reports percentile intervals for most procedures.
INPUT PROGRAM.
COMPUTE #delta = 8 / SQRT(1 + 8**2).   /* alpha = 8 leans hard right */
LOOP #i = 1 TO 30.
  COMPUTE samp = 50 + 20 * (#delta * ABS(RV.NORMAL(0,1))
                            + SQRT(1 - #delta**2) * RV.NORMAL(0,1)).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

BOOTSTRAP /SAMPLING METHOD=SIMPLE /VARIABLES TARGET=samp
  /CRITERIA CILEVEL=95 NSAMPLES=2000 CITYPE=PERCENTILE.
DESCRIPTIVES VARIABLES=samp /STATISTICS=MEAN.
```
### Julia
```julia
using Random, Distributions, Statistics, StatsBase
Random.seed!(42)
# n = 30 from a normal that leans hard right (alpha = 8)
delta = 8 / sqrt(1 + 8^2)
samp = 50 .+ 20 .* (delta .* abs.(rand(Normal(), 30)) .+
                    sqrt(1 - delta^2) .* rand(Normal(), 30))

boot_means = [mean(sample(samp, length(samp), replace = true)) for _ in 1:2000]

mean(samp), std(boot_means),
quantile(boot_means, 0.025), quantile(boot_means, 0.975)
```
### Python
```python
import numpy as np

rng = np.random.default_rng(42)
# n = 30 from a normal that leans hard right (alpha = 8)
delta = 8 / np.sqrt(1 + 8**2)
samp = 50 + 20 * (delta * np.abs(rng.normal(size=30)) +
                  np.sqrt(1 - delta**2) * rng.normal(size=30))

boot_means = np.array([rng.choice(samp, samp.size, replace=True).mean()
                       for _ in range(2000)])

samp.mean(), boot_means.std(ddof=1), *np.quantile(boot_means, [0.025, 0.975])
```
:::

That is the entire idea: `sample(samp, replace = TRUE)` inside `replicate()`, then read the **standard error** off `sd()` and the **confidence interval** off `quantile()`. There is no formula, no normality assumption, and no $z$. The 2.5th and 97.5th percentiles of the bootstrap distribution are a **percentile confidence interval**, and notice that it can be *asymmetric*, honestly reflecting the skew that a normal-theory interval would smooth over.

::: {.panel-tabset}
### R
```{r}
#| label: fig-bootstrap
#| fig-cap: "The bootstrap sampling distribution of the mean, with the percentile 95% interval."
#| fig-height: 4
tibble(resampled_mean = boot_means) |>
  ggplot(aes(x = resampled_mean)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40,
                 fill = "grey80", colour = "white") +
  # the normal the central limit theorem promises for a mean, for comparison
  stat_function(fun = dnorm,
                args = list(mean = mean(boot_means), sd = sd(boot_means)),
                colour = "firebrick", linewidth = 1) +
  geom_vline(xintercept = quantile(boot_means, c(.025, .975)),
             colour = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = mean(samp), colour = "blue", linewidth = 1) +
  labs(x = "resampled mean", y = "density") +
  theme_book()
```
### SPSS
```
* SPSS reports the bootstrap interval as a table rather than drawing the
* resampling distribution. To picture it, save the bootstrap estimates and
* histogram them:
BOOTSTRAP /SAMPLING METHOD=SIMPLE /VARIABLES TARGET=samp
  /CRITERIA NSAMPLES=2000
  /SAVE BOOTSTRAPESTIMATES.

GRAPH /HISTOGRAM = BootstrapMean.
```
### Julia
```julia
using Plots
histogram(boot_means, bins = 40, legend = false,
          xlabel = "resampled mean", ylabel = "count")
vline!(quantile(boot_means, [0.025, 0.975]), color = :red, linestyle = :dash)
vline!([mean(samp)], color = :blue, linewidth = 2)
```
### Python
```python
import matplotlib.pyplot as plt

plt.hist(boot_means, bins=40, color="lightgrey", edgecolor="white")
for q in np.quantile(boot_means, [0.025, 0.975]):
    plt.axvline(q, color="red", linestyle="--", linewidth=2)
plt.axvline(samp.mean(), color="blue", linewidth=2)
plt.xlabel("resampled mean"); plt.ylabel("count"); plt.show()
```
:::

### The permutation test (testing)

The bootstrap *estimated* an interval. To *test* a hypothesis - say, that two groups have the same mean - we resample differently. Under the null "the group label is irrelevant," we can **shuffle the labels** and see how big a difference arises by chance alone. The observed difference is "significant" if it lands in the tail of that shuffled distribution.

::: {.panel-tabset}
### R
```{r}
set.seed(7)
g <- tibble(
  value = c(rnorm(15, mean = 0.0), rnorm(15, mean = 0.8)),
  group = rep(1:2, each = 15)
)

obs <- g |>                                        # the observed difference
  summarise(m = mean(value), .by = group) |>
  arrange(group) |>
  summarise(diff = m[2] - m[1]) |>
  pull(diff)

perm <- map_dbl(1:5000, \(i) {                     # shuffle labels 5000 times
  s <- sample(g$group)
  mean(g$value[s == 2]) - mean(g$value[s == 1])
})

tibble(observed_diff = obs,
       permutation_p = mean(abs(perm) >= abs(obs))) |>   # two-sided p-value
  mutate(permutation_p = fmt_p(permutation_p)) |>
  round2()
```
### SPSS
```
* Base SPSS has no permutation test. The Exact Tests add-on module gives
* exact/Monte-Carlo p-values for many nonparametric procedures, which is
* the same logic applied to rank tests:
NPAR TESTS
  /M-W = value BY group(1 2)
  /MISSING ANALYSIS
  /METHOD=MC CIN(99) SAMPLES(5000).

* A mean-difference permutation test as written in the other tabs has no
* base-SPSS equivalent; it needs the Python or R integration plug-in.
```
### Julia
```julia
using Random, Distributions, Statistics
Random.seed!(7)
value = vcat(rand(Normal(0.0), 15), rand(Normal(0.8), 15))
group = repeat([1, 2], inner = 15)

obs = mean(value[group .== 2]) - mean(value[group .== 1])

perm = [begin
          s = shuffle(group)                        # shuffle labels
          mean(value[s .== 2]) - mean(value[s .== 1])
        end for _ in 1:5000]

obs, mean(abs.(perm) .>= abs(obs))                  # two-sided p-value
```
### Python
```python
import numpy as np

rng = np.random.default_rng(7)
value = np.concatenate([rng.normal(0.0, 1, 15), rng.normal(0.8, 1, 15)])
group = np.repeat([1, 2], 15)

obs = value[group == 2].mean() - value[group == 1].mean()

perm = np.array([(lambda s: value[s == 2].mean() - value[s == 1].mean())
                 (rng.permutation(group)) for _ in range(5000)])

obs, (np.abs(perm) >= abs(obs)).mean()              # two-sided p-value
```
:::

The permutation $p$-value is the fraction of shuffles that beat the real difference. It makes **no distributional assumption at all**; it rests only on the exchangeability of labels under the null.

### The jackknife (a leave-one-out cousin)

Older and simpler than the bootstrap, the **jackknife** recomputes the statistic leaving out one observation at a time. It is handy for a quick standard error and for spotting influential points (if dropping one case moves the estimate a lot, you've found a leverage problem):

::: {.panel-tabset}
### R
```{r}
jk <- map_dbl(seq_along(samp), \(i) mean(samp[-i]))   # leave-one-out means
n  <- length(samp)

tibble(
  jackknife_mean = mean(jk),
  jackknife_SE   = sqrt((n - 1) * mean((jk - mean(jk))^2))
) |>
  round2()
```
### SPSS
```
* Base SPSS has no jackknife. The leave-one-out logic is easy to write by
* hand, though: drop each case in turn and recompute the mean.
* (SPSS's BOOTSTRAP module covers the same ground more directly.)
COMPUTE id = $CASENUM.
EXECUTE.

* For each case i, the mean of everything except i:
COMPUTE whole = 1.
AGGREGATE OUTFILE=* MODE=ADDVARIABLES /BREAK=whole /grand=SUM(samp) /n=N.
COMPUTE loo_mean = (grand - samp) / (n - 1).
EXECUTE.

DESCRIPTIVES VARIABLES=loo_mean /STATISTICS=MEAN STDDEV.
```
### Julia
```julia
jk = [mean(samp[setdiff(1:end, i)]) for i in eachindex(samp)]
n  = length(samp)

mean(jk), sqrt((n - 1) * mean((jk .- mean(jk)).^2))     # jackknife mean and SE
```
### Python
```python
jk = np.array([np.delete(samp, i).mean() for i in range(samp.size)])
n = samp.size

jk.mean(), np.sqrt((n - 1) * ((jk - jk.mean())**2).mean())  # jackknife mean, SE
```
:::

::: {.callout-note}
## Estimation vs. testing, the resampling version
Same engine, two jobs. **Bootstrap → estimation** (an SE and a percentile interval for *whatever* statistic you can compute - a mean, a correlation, a regression slope, even a mediation indirect effect $a\cdot b$, as in the authors' own bootstrapped mixed-effects mediation of coping and physical functioning [@McKnightAfram2010]). **Permutation → testing** (a $p$-value from label shuffling). And the two meet: a bootstrap **interval that excludes zero** *is* a test. Resampling has limits worth respecting: it is **not** a substitute for collecting more data, and it still requires a representative, random sample to begin with.
:::

## Part 2 · Robust Statistics: When a Few Points Hijack the Mean

Recall from the [central tendency chapter](03-CTmeasures.qmd) that the mean is a *least-squares* estimator: it sits where squared deviations are smallest. Squaring is exactly why it's fragile - one faraway point contributes an enormous squared deviation and drags the mean toward itself. A single bad value can move the mean anywhere; its **breakdown point** (the fraction of contamination it can tolerate before giving nonsense) is essentially $0\%$.

Watch nineteen clean observations and one data-entry error:

::: {.panel-tabset}
### R
```{r}
set.seed(1)
z <- c(rnorm(19, mean = 100, sd = 10), 250)        # 19 good values + 1 gross outlier

tibble(
  mean      = mean(z),
  median    = median(z),
  trimmed10 = mean(z, trim = 0.10),   # drop the extreme 10% in each tail
  SD        = sd(z),
  MAD       = mad(z)                  # median absolute deviation
) |>
  round2()
```
### SPSS
```
* Nineteen clean observations and one data-entry error.
INPUT PROGRAM.
LOOP #i = 1 TO 19.
  COMPUTE z = RV.NORMAL(100, 10).
  END CASE.
END LOOP.
COMPUTE z = 250.
END CASE.
END FILE.
END INPUT PROGRAM.
EXECUTE.

* EXAMINE reports the mean, median, the 5% trimmed mean, and the SD.
EXAMINE VARIABLES=z /STATISTICS=DESCRIPTIVES.

* MAD has no base-SPSS keyword; compute it from the median directly:
* COMPUTE absdev = ABS(z - <median>).  then take the median of absdev,
* multiplied by 1.4826 for consistency with the SD under normality.
```
### Julia
```julia
using Statistics, StatsBase
Random.seed!(1)
z = vcat(rand(Normal(100, 10), 19), 250.0)   # 19 good values + 1 gross outlier

mean(z), median(z), mean(trim(z, prop = 0.10)), std(z), mad(z)
```
### Python
```python
from scipy import stats

rng = np.random.default_rng(1)
z = np.append(rng.normal(100, 10, 19), 250)  # 19 good values + 1 gross outlier

(z.mean(), np.median(z), stats.trim_mean(z, 0.10),
 z.std(ddof=1), stats.median_abs_deviation(z, scale="normal"))
```
:::

The **mean** and **SD** are inflated by the lone outlier; the **median**, **10% trimmed mean**, and **MAD** barely notice it. That resistance is the whole point of robust statistics:

- **Median** - breakdown point $50\%$: half your data can be garbage before it fails.
- **Trimmed mean** - discard a fixed fraction of each tail, then average the rest. A tunable compromise between mean (0% trim) and median (50% trim).
- **MAD** - the robust analogue of the SD, a median of absolute deviations from the median.
- **M-estimators** (Huber, Tukey bisquare) - down-weight, rather than discard, extreme points, via iteratively reweighted least squares. These are the robust-regression workhorses.

For real robust modeling in R, reach for maintained tools - `MASS::rlm()` for robust regression, and Rand Wilcox's **`WRS2`** package for robust group comparisons (trimmed-means tests, robust ANOVA). The idioms look like this:

```r
library(MASS)
rlm(happy ~ dogness, data = d)          # robust regression (Huber M-estimator)

library(WRS2)
yuen(happy ~ group, data = d)           # robust two-group test on trimmed means
t1way(happy ~ group, data = d)          # robust one-way "ANOVA"
```

::: {.callout-note}
## Estimation vs. testing, the robust version
**Estimation:** report a robust *center and spread* - a trimmed mean with a MAD, or robust regression coefficients - instead of mean $\pm$ SD when the tails are heavy. **Testing:** use tests built on those robust estimators (Yuen's trimmed-means test, `t1way`) or - the pragmatic favorite in this book - **bootstrap the robust estimator** and check whether its interval excludes the null. Robustness and resampling are natural allies: diagnose the violation, then resample the resistant estimator.
:::

## Part 3 · Bayesian Inference: Probability of the Model Given the Data

Frequentist statistics computes $P(\text{data} \mid \text{hypothesis})$: the $p$-value is the probability of data this extreme *if the null were true*. That is often not the quantity we actually care about. Bayesian inference computes what you usually want, $P(\text{hypothesis} \mid \text{data})$, by way of **Bayes' rule**:

$$\underbrace{P(\theta \mid \text{data})}_{\text{posterior}} \;\propto\; \underbrace{P(\text{data} \mid \theta)}_{\text{likelihood}} \times \underbrace{P(\theta)}_{\text{prior}}$$

The **prior** is what you believed about the parameter $\theta$ before the data; the **likelihood** is what the data say; the **posterior** is the updated belief. That's the whole machine. As McKnight puts it, Bayesian methods are "merely a vehicle for re-conceptualizing how we think about statistics" - they give you "insights you may already know, but didn't know you knew."

### Why it matters: the arithmetic of the replication crisis

Here is a calculation worth sitting with, and it needs nothing but Bayes' rule. You run a well-powered study, get $p < .05$, and publish. **What is the probability your hypothesis is actually true?** It is not the $p$-value; that is $P(\text{significant} \mid H \text{ false})$. The quantity you want is $P(H \text{ true} \mid \text{significant})$, and it depends on how plausible your idea was *before* the data, its prior:

::: {.panel-tabset}
### R
```{r}
tibble(
  power = 0.99,   # P(significant | H true)  -- a very high-powered study
  alpha = 0.01,   # P(significant | H false) -- a strict false-positive rate
  prior = 0.02    # P(H true): your idea is ~1 of 50 plausible hypotheses
) |>
  mutate(posterior =                                        # Bayes' rule
    (power * prior) / (power * prior + alpha * (1 - prior))) |>
  round2()
```
### SPSS
```
* Bayes' rule applied to a screening question: given power, alpha, and a
* prior, what is the posterior probability the hypothesis is true?
DATA LIST FREE /power alpha prior.
BEGIN DATA
0.99 0.01 0.02
END DATA.

COMPUTE posterior = (power * prior) / (power * prior + alpha * (1 - prior)).
EXECUTE.
LIST VARIABLES=power alpha prior posterior.
```
### Julia
```julia
power, alpha, prior = 0.99, 0.01, 0.02   # very high power, strict alpha, rare idea

posterior = (power * prior) / (power * prior + alpha * (1 - prior))
```
### Python
```python
power, alpha, prior = 0.99, 0.01, 0.02   # very high power, strict alpha, rare idea

posterior = (power * prior) / (power * prior + alpha * (1 - prior))
```
:::

Even with $99\%$ power and a $1\%$ false-positive rate, a significant result leaves you only about **67%** sure - because the idea was a long shot to begin with. Loosen the study to more realistic numbers and it collapses further:

::: {.panel-tabset}
### R
```{r}
# modest power, ordinary screening, a rarer idea
tibble(power = 0.95, alpha = 0.05, prior = 0.01) |>
  mutate(posterior =
    (power * prior) / (power * prior + alpha * (1 - prior))) |>
  round2()
```
### SPSS
```
* Modest power, ordinary screening, a rarer idea.
DATA LIST FREE /power alpha prior.
BEGIN DATA
0.95 0.05 0.01
END DATA.

COMPUTE posterior = (power * prior) / (power * prior + alpha * (1 - prior)).
EXECUTE.
LIST VARIABLES=posterior.
```
### Julia
```julia
power, alpha, prior = 0.95, 0.05, 0.01
(power * prior) / (power * prior + alpha * (1 - prior))
```
### Python
```python
power, alpha, prior = 0.95, 0.05, 0.01
(power * prior) / (power * prior + alpha * (1 - prior))
```
:::

About **16%**. Chase enough low-prior ideas with $p < .05$ and *most* of your "discoveries" will be false, no matter how small the $p$-value. This is the arithmetic behind the replication crisis we met in [hypothesis testing](08-hypothesistesting.qmd). The $p$-value does not answer the question you cared about, and Bayes' rule does, but it asks you to state your prior out loud. That honesty is the point.

### Building a posterior by hand (grid approximation)

You do not need fancy software to see the machine turn. Take the oldest example in the book - estimating a proportion (a coin's bias, a globe's water fraction, a treatment's success rate) from a handful of trials. Lay down a **grid** of possible values, evaluate prior × likelihood at each, normalize, and you have a posterior:

::: {.panel-tabset}
### R
```{r}
#| label: fig-posterior
#| fig-cap: "Grid-approximation posterior for a proportion after 6 successes in 9 trials."
#| fig-height: 4
posterior <- tibble(theta = seq(0, 1, length.out = 1000)) |>
  mutate(
    prior = dbeta(theta, 2, 2),                 # weakly informative, centered at 0.5
    like  = dbinom(6, size = 9, prob = theta),  # data: 6 successes in 9 trials
    post  = like * prior,
    post  = post / sum(post)                    # normalize to a proper distribution
  )

ggplot(posterior, aes(x = theta, y = post)) +
  geom_line(colour = "blue", linewidth = 1) +
  labs(x = expression(theta ~ "(the proportion)"), y = "posterior density") +
  theme_book()
```
### SPSS
```
* Grid approximation of a posterior for a proportion after 6 successes
* in 9 trials, with a Beta(2,2) prior.
INPUT PROGRAM.
LOOP #i = 0 TO 1000.
  COMPUTE theta = #i / 1000.
  COMPUTE prior = PDF.BETA(theta, 2, 2).
  COMPUTE like  = PDF.BINOM(6, 9, theta).
  COMPUTE post  = like * prior.
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

* Normalise, then plot.
COMPUTE whole = 1.
AGGREGATE OUTFILE=* MODE=ADDVARIABLES /BREAK=whole /tot=SUM(post).
COMPUTE post = post / tot.
EXECUTE.

GRAPH /LINE(SIMPLE) = VALUE(post) BY theta
  /TITLE = 'Grid-approximation posterior'.
```
### Julia
```julia
using Distributions, Plots
theta = range(0, 1, length = 1000)
prior = pdf.(Beta(2, 2), theta)                   # weakly informative
like  = [pdf(Binomial(9, t), 6) for t in theta]   # data: 6 successes in 9 trials
post  = like .* prior; post ./= sum(post)

plot(theta, post, linewidth = 2, legend = false,
     xlabel = "theta (the proportion)", ylabel = "posterior density")
```
### Python
```python
import numpy as np
import matplotlib.pyplot as plt
from scipy.stats import beta, binom

theta = np.linspace(0, 1, 1000)
prior = beta.pdf(theta, 2, 2)                # weakly informative
like  = binom.pmf(6, 9, theta)               # data: 6 successes in 9 trials
post  = like * prior; post /= post.sum()

plt.plot(theta, post, color="blue", linewidth=2)
plt.xlabel(r"$\theta$ (the proportion)"); plt.ylabel("posterior density")
plt.show()
```
:::

Now *summarize* the posterior - and here the estimation-vs-testing split becomes concrete.

**Estimation.** The answer is the whole posterior; we compress it into a point and an interval. Sampling from the grid makes this a one-liner. A Bayesian interval is a **credible** (or "compatibility") interval, and - unlike a confidence interval - it means the plain-English thing: "given the model and data, there's a 95% probability $\theta$ is in here."

::: {.panel-tabset}
### R
```{r}
draws <- sample(posterior$theta, 1e5, replace = TRUE, prob = posterior$post)

tibble(
  posterior_mean   = mean(draws),
  credible_lower95 = quantile(draws, 0.025),
  credible_upper95 = quantile(draws, 0.975)
) |>
  round2()
```
### SPSS
```
* Sampling from the grid posterior needs a weighted draw, which base SPSS
* does not offer directly. The posterior mean and quantiles can instead be
* read off the grid itself, which is exact rather than simulated:
*   posterior mean = SUM(theta * post)
*   the 2.5th/97.5th percentiles are where the CUMULATIVE post crosses
*   0.025 and 0.975.
CREATE cum_post = CSUM(post).
EXECUTE.

TEMPORARY.
SELECT IF (cum_post >= 0.025 AND cum_post <= 0.975).
DESCRIPTIVES VARIABLES=theta /STATISTICS=MINIMUM MAXIMUM.
```
### Julia
```julia
using StatsBase
draws = sample(collect(theta), Weights(post), 100_000)

mean(draws), quantile(draws, 0.025), quantile(draws, 0.975)
```
### Python
```python
rng = np.random.default_rng()
draws = rng.choice(theta, 100_000, p=post)

draws.mean(), *np.quantile(draws, [0.025, 0.975])
```
:::

**Testing.** Rather than a point-null "is $\theta$ *exactly* 0.5?" (a question of measure zero that is never truly true), Bayesians ask a better one: how much posterior belief sits in a **Region of Practical Equivalence (ROPE)** around the null - a band of values "close enough to 0.5 to not care"? If almost none of the posterior is in the ROPE, the effect is credibly *not* null; if most of it is, you have evidence *for* practical equivalence - something a $p$-value can never give you.

::: {.panel-tabset}
### R
```{r}
rope_lo <- 0.45; rope_hi <- 0.55   # "practically a fair coin"

# posterior probability inside the ROPE
tibble(p_inside_rope = mean(draws > rope_lo & draws < rope_hi)) |> round2()
```
### SPSS
```
* The posterior probability inside a region of practical equivalence
* (here 0.45 to 0.55, "practically a fair coin"). On the grid this is
* simply the posterior mass between those two values.
TEMPORARY.
SELECT IF (theta > 0.45 AND theta < 0.55).
DESCRIPTIVES VARIABLES=post /STATISTICS=SUM.
```
### Julia
```julia
rope_lo, rope_hi = 0.45, 0.55        # "practically a fair coin"
mean(rope_lo .< draws .< rope_hi)     # posterior probability inside the ROPE
```
### Python
```python
rope_lo, rope_hi = 0.45, 0.55        # "practically a fair coin"
((draws > rope_lo) & (draws < rope_hi)).mean()   # probability inside the ROPE
```
:::

### From the toy to the real thing

Grid approximation shows the logic but doesn't scale past a parameter or two. For real models, samplers (MCMC) do the integration. McKnight's teaching stack is McElreath's **`rethinking`** for concepts and **`brms`** for applied work - both compile to Stan under the hood. The *thinking* is identical to the grid above; only the engine changes:

```r
library(rethinking)
m <- quap(alist(
        y    ~ dnorm(mu, sigma),     # likelihood
        mu   <- a + b * x,
        a    ~ dnorm(0, 10),         # priors
        b    ~ dnorm(0, 1),          #   weakly informative on a standardized predictor
        sigma ~ dexp(1)
     ), data = d)
post <- extract.samples(m)
HPDI(post$b, prob = 0.89)            # highest-posterior-density (credible) interval
compare(m1, m2, func = WAIC)         # model comparison (predictive testing)
```

```r
library(brms); library(bayestestR)
fit <- brm(y ~ x, data = d, family = gaussian(),
           prior = prior(normal(0, 1), class = "b"))   # regularizing prior
hdi(fit)                             # estimation: credible intervals
p_direction(fit)                     # probability the effect has a given sign
rope(fit, range = c(-0.05, 0.05))    # testing: mass in the ROPE
loo(fit)                             # model comparison via leave-one-out
```

Two habits from McKnight's applied pipeline are worth stealing: **standardize predictors** so a `normal(0, 1)` prior is genuinely weakly-informative and interpretable, and **simulate from your priors before touching the data** to check they imply sane predictions (an over-tight prior underfits; a too-vague one overfits).

::: {.callout-note}
## Estimation vs. testing, the Bayesian version
**Estimation is primary and everything else is downstream of it.** Fit the model, get a posterior, report a **credible interval** (HPDI). "Testing" then comes in two honest flavors, neither of which is point-null rejection: **practical significance** via the **ROPE** (how much posterior is near the null?), and **model comparison** via predictive scores (**WAIC**, **PSIS-LOO**) or **Bayes factors**. A caution McKnight repeats: *predictive testing is not causal estimation*. A confounded model can out-predict the correct causal one, so "which model forecasts better" and "which parameter is right" are different questions. Do not let a good WAIC override a good DAG.
:::

## The Distinction, Woven Together

One table, three frameworks, two questions:

| Framework | **Estimation** (value + uncertainty) | **Testing** (difference / comparison) |
|---|---|---|
| Classical (normal theory) | estimate $\pm$ standard error; confidence interval | $t$/$F$ test, $p$-value |
| **Resampling** | bootstrap SE + percentile interval | permutation $p$; bootstrap interval excludes null |
| **Robust** | trimmed mean + MAD; robust regression coefficients | Yuen/`t1way`; bootstrap the robust estimator |
| **Bayesian** | posterior + credible (HPDI) interval | ROPE; model comparison (WAIC/LOO); Bayes factor |

Read down the **estimation** column and the four approaches are more alike than different: all report a best value and a spread. Read down the **testing** column and you see the modern lesson of this whole book - **testing is best done as a consequence of estimation, not as a separate ritual.** Estimate the distribution of your answer honestly, and the test is already in your hands.

## Challenge

::: {.callout-tip}
## Do One Yourself
1.  Bootstrap a statistic that has *no* simple standard-error formula - the **median**, or the **correlation** between two variables. Compare your percentile interval to what you'd get pretending it was normal.
2.  Take the outlier demo (`z`) and sweep the trimmed mean's trim fraction from 0 to 0.5. Plot the trimmed mean against the trim fraction. Where does it stop moving, and what value has it become?
3.  Re-run the grid-approximation posterior with a *strong* prior (`dbeta(grid, 20, 20)`) and a *flat* one (`dbeta(grid, 1, 1)`). How much does the prior move the posterior with only 9 trials, and with a simulated 900 trials? What does that tell you about when priors matter?
:::

## Where We Go Next

You now have three ways past the normal curve, and a single principle - estimate first, test second - that ties them together and reaches back through every chapter you've read. That principle is the quiet thesis of this book: statistics is the disciplined quantification of uncertainty, and the honest way to express uncertainty is a *distribution*, not a verdict. Wherever your data take you, whether a resampled interval, a robust slope, or a posterior, you are doing the same thing we started with: saying what you know, and how sure you are.

And knowing something is only half the job. The other half is *showing* it - honestly, to someone else who did not run your analysis and will not read your code. That craft, the display of data in [graphics](21-graphics.qmd) and [tables](22-tables.qmd), is where we turn next.
