# Displaying Data: Graphics {#graphics}

A graph is more than decoration. A graph is an **argument**, a claim about what the numbers mean, made in a language (position, length, color) that the eye reads faster than any sentence. And like any argument, a graph can be honest or it can mislead, sometimes without its maker even noticing. This chapter is about making the honest kind, and recognizing the other kind. It is also about what a graph can *reveal*: [Anscombe's quartet](07-covariance.qmd) showed four datasets with identical statistics but four completely different shapes, a reminder that a picture often catches what a summary number hides.

The examples here come from a real project - [The Sports Page](https://thesportspage.net/), a daily newsletter (a sister project to this book) that takes one strange sports number and tells the truth about what it does and does not mean. Its house rule is the same one that runs through this whole book: *describe what happened, predict what will happen, and never dress up one as the other.* Good graphics are how you keep that promise to your reader.

## Learning Objectives

1.  Match the **geometry** of a graph to the question (trend, comparison, distribution, relationship).
2.  Recognize how an axis can exaggerate or hide a real change - and choose an honest one.
3.  Prefer **small multiples** to overloaded dual-axis charts, and **ranked bars** to pie charts.
4.  Show the **distribution**, not just a summary, and tame overplotting.

## Match the Geometry to the Question

The first decision is not color or font - it is *geometry*. What shape of mark answers your question? A change **over time** wants a line. Here is a true story: over the last four decades, major-league baseball became a strikeout sport. Watch it in one line:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
Every figure here uses real data that ships with the book: `data/mlb_league_by_year.csv`, `data/mlb_hitters_2026.csv`, and `data/nfl_draft_by_team.csv`. Read them directly - no setup needed. See [Getting the Book's Data](setup.qmd) if you want the setup files for your language.
:::

::: {.panel-tabset}
### R
```{r}
#| fig-cap: "Strikeouts per nine innings, MLB, 1985–present. A trend wants a line."
#| fig-height: 3.6
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

mlb <- read_csv("data/mlb_league_by_year.csv", show_col_types = FALSE)

ggplot(mlb, aes(year, K_9)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 1.4) +
  labs(x = NULL, y = "Strikeouts per 9 innings",
       title = "The strikeout revolution") +
  theme_minimal(base_size = 12)
```
### SPSS
```
* A trend over time wants a line.
GET DATA /TYPE=TXT /FILE='data/mlb_league_by_year.csv'
  /DELIMITERS=',' /FIRSTCASE=2.

GRAPH /LINE(SIMPLE) = VALUE(K_9) BY year
  /TITLE = 'The strikeout revolution'.
```
### Julia
```julia
using CSV, DataFrames, Plots
mlb = CSV.read("data/mlb_league_by_year.csv", DataFrame)

plot(mlb.year, mlb.K_9, linewidth = 2, marker = :circle, legend = false,
     ylabel = "Strikeouts per 9 innings", title = "The strikeout revolution")
```
### Python
```python
import pandas as pd
import matplotlib.pyplot as plt

mlb = pd.read_csv("data/mlb_league_by_year.csv")

plt.plot(mlb.year, mlb.K_9, color="steelblue", linewidth=2, marker="o", ms=4)
plt.ylabel("Strikeouts per 9 innings")
plt.title("The strikeout revolution")
plt.show()
```
:::

A line says "these points are connected in sequence" - exactly the claim you want for a time trend. A bar chart here would be wrong: bars say "these are separate categories to compare," which is a different (and misleading) sentence. **The geometry is the argument.** Trend → line; category comparison → bar; one distribution → histogram; two variables → scatter. Pick the mark that makes the claim you actually mean.

## Don't Lie With the Axis

Same data, two axes, two very different impressions. Batting average has drifted down over the same period, from about .257 to about .242, a real but *modest* slide. First, watch a zoomed-in y-axis make that modest slide look like a catastrophe:

::: {.panel-tabset}
### R
```{r}
#| fig-cap: "Dishonest axis: zoomed tight, a 15-point drift looks like a cliff."
#| fig-height: 3.0
#| message: false
ggplot(mlb, aes(year, AVG)) +
  geom_line(color = "firebrick", linewidth = 1) +
  coord_cartesian(ylim = c(0.242, 0.258)) +                 # axis zoomed to the data
  labs(x = NULL, y = "AVG", title = "“Hitting is DEAD”") +
  theme_minimal(base_size = 11)
```
### SPSS
```
* Dishonest axis: zoom the y scale tight and a 15-point drift looks
* like a cliff. In the chart editor you would set the y-axis minimum to
* .242 and maximum to .258; in syntax, GGRAPH carries the scale range.
GGRAPH
  /GRAPHDATASET NAME="g" VARIABLES=year AVG
  /GRAPHSPEC SOURCE=INLINE
  /FRAME INNER=YES.
BEGIN GPL
  SOURCE: s=userSource(id("g"))
  DATA: year=col(source(s), name("year"))
  DATA: AVG=col(source(s), name("AVG"))
  SCALE: linear(dim(2), min(0.242), max(0.258))
  GUIDE: text.title(label("Hitting is DEAD"))
  ELEMENT: line(position(year*AVG))
END GPL.
```
### Julia
```julia
plot(mlb.year, mlb.AVG, linewidth = 2, color = :firebrick, legend = false,
     ylims = (0.242, 0.258),          # axis zoomed to the data
     ylabel = "AVG", title = "\"Hitting is DEAD\"")
```
### Python
```python
plt.plot(mlb.year, mlb.AVG, color="firebrick", linewidth=2)
plt.ylim(0.242, 0.258)                # axis zoomed to the data
plt.ylabel("AVG"); plt.title('"Hitting is DEAD"')
plt.show()
```
:::

Now the exact same numbers, on an axis anchored to a meaningful baseline:

::: {.panel-tabset}
### R
```{r}
#| fig-cap: "Honest axis: from a meaningful zero, the same drift is a gentle slope."
#| fig-height: 3.0
#| message: false
ggplot(mlb, aes(year, AVG)) +
  geom_line(color = "firebrick", linewidth = 1) +
  coord_cartesian(ylim = c(0, 0.3)) +                        # axis from a meaningful zero
  labs(x = NULL, y = "AVG", title = "“Hitting drifted down a bit”") +
  theme_minimal(base_size = 11)
```
### SPSS
```
* Honest axis: from a meaningful zero, the same drift is a gentle slope.
GGRAPH
  /GRAPHDATASET NAME="g" VARIABLES=year AVG
  /GRAPHSPEC SOURCE=INLINE
  /FRAME INNER=YES.
BEGIN GPL
  SOURCE: s=userSource(id("g"))
  DATA: year=col(source(s), name("year"))
  DATA: AVG=col(source(s), name("AVG"))
  SCALE: linear(dim(2), min(0), max(0.3))
  GUIDE: text.title(label("Hitting drifted down a bit"))
  ELEMENT: line(position(year*AVG))
END GPL.
```
### Julia
```julia
plot(mlb.year, mlb.AVG, linewidth = 2, color = :firebrick, legend = false,
     ylims = (0, 0.3),                # axis from a meaningful zero
     ylabel = "AVG", title = "\"Hitting drifted down a bit\"")
```
### Python
```python
plt.plot(mlb.year, mlb.AVG, color="firebrick", linewidth=2)
plt.ylim(0, 0.3)                      # axis from a meaningful zero
plt.ylabel("AVG"); plt.title('"Hitting drifted down a bit"')
plt.show()
```
:::

Same numbers, opposite impressions. The first axis turns a 15-point drift into what looks like a cliff; the second, anchored to a meaningful baseline, shows the truth: hitting softened, it did not collapse. Neither axis is *always* right. A zoomed axis is legitimate when small differences genuinely matter (and you say so), but the choice is never neutral. **Ask what the axis is doing to the size of the effect, and whether that is honest.**

::: {.callout-warning}
## The most common accidental lie
A truncated y-axis is the single most common way a well-meaning person makes a small effect look enormous (or the reverse). Before you publish a chart, look at your axis limits and ask: *would this effect still look this big if I started from zero?* If the honest version looks unremarkable, the honest answer might be that the effect is unremarkable.
:::

## Small Multiples Beat Overloaded Dual-Axis Charts

You have three trends to show at once: strikeouts, walks, and home runs per plate appearance. The temptation is to put them all on one chart with two or three y-axes. It is better not to. Dual-axis charts let you *manufacture* almost any relationship by rescaling one axis, and readers cannot tell which line belongs to which scale. The clearer tool is **small multiples**: the same chart repeated, one panel per series, on comparable axes.

::: {.panel-tabset}
### R
```{r}
#| fig-cap: "Three league trends as small multiples — comparable, honest, readable."
#| fig-height: 3.2
#| message: false
long <- mlb |>
  select(year, Strikeouts = K_PA, Walks = BB_PA, `Home runs` = HR_PA) |>
  pivot_longer(-year, names_to = "kind", values_to = "rate")

ggplot(long, aes(year, rate)) +
  geom_line(color = "steelblue", linewidth = 0.9) +
  facet_wrap(~ kind, scales = "free_y") +               # one panel per series
  labs(x = NULL, y = "per plate appearance") +
  theme_minimal(base_size = 11)
```
### SPSS
```
* Small multiples: three league trends, one panel each.
* SPSS facets with a PANEL clause inside GGRAPH. The data must be stacked
* long first (one row per year per series).
VARSTOCASES /MAKE rate FROM K_PA BB_PA HR_PA /INDEX=kind.

GGRAPH
  /GRAPHDATASET NAME="g" VARIABLES=year rate kind
  /GRAPHSPEC SOURCE=INLINE.
BEGIN GPL
  SOURCE: s=userSource(id("g"))
  DATA: year=col(source(s), name("year"))
  DATA: rate=col(source(s), name("rate"))
  DATA: kind=col(source(s), name("kind"), unit.category())
  GUIDE: axis(dim(1), label("year"))
  GUIDE: axis(dim(2), label("per plate appearance"))
  ELEMENT: line(position(year*rate*kind))
END GPL.
```
### Julia
```julia
using DataFrames, Plots
long = stack(select(mlb, :year, :K_PA, :BB_PA, :HR_PA),
             [:K_PA, :BB_PA, :HR_PA], :year,
             variable_name = :kind, value_name = :rate)

# one panel per series
plot([plot(g.year, g.rate, title = string(first(g.kind)), legend = false)
      for g in groupby(long, :kind)]..., layout = (1, 3))
```
### Python
```python
long = mlb.melt(id_vars="year", value_vars=["K_PA", "BB_PA", "HR_PA"],
                var_name="kind", value_name="rate")

fig, axes = plt.subplots(1, 3, figsize=(10, 3))   # one panel per series
for ax, (kind, g) in zip(axes, long.groupby("kind")):
    ax.plot(g.year, g.rate, color="steelblue")
    ax.set_title(kind)
axes[0].set_ylabel("per plate appearance")
fig.tight_layout(); plt.show()
```
:::

Each panel tells its own honest story (strikeouts soared, walks held steady, homers crept up), and no rescaling trick can fake a relationship between them. When you have many series, **repeat the chart; don't overload it.**

## Rank, Don't Pie

For comparing a quantity across categories, the eye reads **position along a common scale** far more accurately than it reads the *angles* of a pie. So a bar chart beats a pie chart almost always - and a *sorted* bar chart beats an unsorted one, because the ordering is itself information. Which NFL teams draft worst? Thirty-two teams is already far too many for a pie; rank them:

::: {.panel-tabset}
### R
```{r}
#| fig-cap: "NFL first-round bust rate by team — ranked bars, not a 32-slice pie."
#| fig-height: 5.2
#| message: false
nfl <- read_csv("data/nfl_draft_by_team.csv", show_col_types = FALSE)

ggplot(nfl, aes(x = fct_reorder(team, rate_bust_a), y = rate_bust_a)) +
  geom_col(fill = "grey35") +
  coord_flip() +                                        # horizontal: labels stay readable
  labs(x = NULL, y = "Bust rate (no Pro Bowl)", title = "Draft busts by team") +
  theme_minimal(base_size = 10)
```
### SPSS
```
* Rank, don't pie: a sorted horizontal bar chart of bust rate by team.
GET DATA /TYPE=TXT /FILE='data/nfl_draft_by_team.csv'
  /DELIMITERS=',' /FIRSTCASE=2.

SORT CASES BY rate_bust_a (A).

GRAPH /BAR(SIMPLE) = VALUE(rate_bust_a) BY team
  /TITLE = 'Draft busts by team'.
* In the chart editor, transpose to horizontal so the team labels stay
* readable - the equivalent of coord_flip().
```
### Julia
```julia
using CSV, DataFrames, Plots
nfl = CSV.read("data/nfl_draft_by_team.csv", DataFrame)
sort!(nfl, :rate_bust_a)                       # sorted: the ranking IS the message

bar(nfl.team, nfl.rate_bust_a, orientation = :horizontal, legend = false,
    xlabel = "Bust rate (no Pro Bowl)", title = "Draft busts by team")
```
### Python
```python
nfl = pd.read_csv("data/nfl_draft_by_team.csv")
nfl = nfl.sort_values("rate_bust_a")           # sorted: the ranking IS the message

plt.figure(figsize=(6, 8))
plt.barh(nfl.team, nfl.rate_bust_a, color="0.35")   # horizontal: labels readable
plt.xlabel("Bust rate (no Pro Bowl)"); plt.title("Draft busts by team")
plt.tight_layout(); plt.show()
```
:::

Sorted, horizontal, one bar per team: you can read off the best and worst drafters quickly, and every team is comparable on the same axis. A pie chart of these same 32 numbers would be very hard to read; you could not tell 22% from 25% by eye, and the ranking would disappear. **Pie charts work for maybe two or three slices; past that, rank a bar chart.**

## Show the Distribution, Not Just the Mean

A single summary number - a mean, a bar - throws away the shape of the data, and the shape is usually the story. Do not draw a bar of averages when you could draw the whole distribution. Here are this season's qualified hitters; a histogram shows the *spread* of performance a mean would erase:

::: {.panel-tabset}
### R
```{r}
#| fig-cap: "The distribution of OPS across qualified hitters — the shape a single mean hides."
#| fig-height: 3.3
#| message: false
h <- read_csv("data/mlb_hitters_2026.csv", show_col_types = FALSE) |>
  filter(PA >= 50)

ggplot(h, aes(OPS)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = median(h$OPS), color = "firebrick", linewidth = 1) +
  labs(x = "OPS", y = "hitters", title = "Most hitters are ordinary; a few are stars") +
  theme_minimal(base_size = 12)
```
### SPSS
```
* Show the distribution, not just the mean.
GET DATA /TYPE=TXT /FILE='data/mlb_hitters_2026.csv'
  /DELIMITERS=',' /FIRSTCASE=2.
SELECT IF (PA >= 50).
EXECUTE.

FREQUENCIES VARIABLES=OPS /STATISTICS=MEDIAN /HISTOGRAM.
```
### Julia
```julia
using CSV, DataFrames, Statistics, Plots
h = filter(:PA => >=(50), CSV.read("data/mlb_hitters_2026.csv", DataFrame))

histogram(h.OPS, bins = 30, legend = false, xlabel = "OPS", ylabel = "hitters",
          title = "Most hitters are ordinary; a few are stars")
vline!([median(h.OPS)], color = :firebrick, linewidth = 2)
```
### Python
```python
h = pd.read_csv("data/mlb_hitters_2026.csv").query("PA >= 50")

plt.hist(h.OPS, bins=30, color="steelblue", edgecolor="white")
plt.axvline(h.OPS.median(), color="firebrick", linewidth=2)
plt.xlabel("OPS"); plt.ylabel("hitters")
plt.title("Most hitters are ordinary; a few are stars")
plt.show()
```
:::

And when you plot two variables against each other with hundreds of points, they pile up and hide their own density. The fix is **transparency** (or jitter): let the dark regions show where the data actually concentrate.

::: {.panel-tabset}
### R
```{r}
#| fig-cap: "On-base vs. slugging, 400+ hitters. Transparency reveals the crowd."
#| fig-height: 3.4
#| message: false
ggplot(h, aes(OBP, SLG)) +
  geom_point(alpha = 0.35, color = "steelblue") +         # alpha tames overplotting
  labs(x = "On-base percentage", y = "Slugging percentage") +
  theme_minimal(base_size = 12)
```
### SPSS
```
* Transparency tames overplotting. SPSS sets marker transparency in the
* chart editor, or through GGRAPH's transparency() aesthetic.
GGRAPH
  /GRAPHDATASET NAME="g" VARIABLES=OBP SLG
  /GRAPHSPEC SOURCE=INLINE.
BEGIN GPL
  SOURCE: s=userSource(id("g"))
  DATA: OBP=col(source(s), name("OBP"))
  DATA: SLG=col(source(s), name("SLG"))
  GUIDE: axis(dim(1), label("On-base percentage"))
  GUIDE: axis(dim(2), label("Slugging percentage"))
  ELEMENT: point(position(OBP*SLG), transparency.exterior(transparency."0.65"))
END GPL.
```
### Julia
```julia
scatter(h.OBP, h.SLG, alpha = 0.35, legend = false,   # alpha tames overplotting
        xlabel = "On-base percentage", ylabel = "Slugging percentage")
```
### Python
```python
plt.scatter(h.OBP, h.SLG, alpha=0.35, color="steelblue")  # tames overplotting
plt.xlabel("On-base percentage"); plt.ylabel("Slugging percentage")
plt.show()
```
:::

The cloud's shape, with its positive drift, its dense middle, and a few outliers in the upper right, is information a table of two means could never carry. That is much of the reason graphics exist.

::: {.callout-important}
## The honest-graph checklist
Before a chart leaves your hands, run it past four questions:

1.  **Geometry** - does the mark (line/bar/histogram/scatter) make the claim I actually mean?
2.  **Axis** - would the effect look this big if I started from a meaningful zero? If not, am I saying so?
3.  **Load** - am I overloading one chart (dual axes, too many series, a 32-slice pie) where small multiples or a ranked bar would be clearer?
4.  **Honesty of summary** - am I hiding a distribution behind a mean, or implying a relationship the data (small $n$, no mechanism) haven't earned?

A graph that survives all four is an argument you can stand behind.
:::

## Challenge

::: {.callout-tip}
## Do One Yourself
1.  Redraw the strikeout-revolution line with a truncated y-axis that makes the rise look even more extreme, then with a zero-based axis that makes it look tame. Which is honest, and why does "honest" depend on the question?
2.  Take the hitters data and draw a bar chart of *mean* OPS by team, then a boxplot (or jittered points) of OPS by team. What does the second show that the first hides?
3.  Find a chart in the wild (a news site, a company slide) with a truncated axis or a pie chart with too many slices. Redraw it honestly and write one sentence on what changed.
:::

## Where We Go Next

Graphics are half of how we show data; the other half is the **table**, and tables have their own quiet craft, their own ways of telling the truth or hiding it. That is the [next chapter](22-tables.qmd). Then, with tables and graphics both in hand, we turn to the payoff of this whole book: watching these tools used on real published work, with every judgment call laid bare.
