# Case Study: When the Total Score Hides the Story {#depression}

Our first real study is one of our own - a collaboration (McKnight, Stuewig, Kashdan) that started from a point you have already met in this book: **the mean throws away information.** In the [central tendency chapter](03-CTmeasures.qmd) that was a general idea. Here it becomes a research program, with real patients and a real depression scale. It also includes something you rarely see in a finished paper: a claim we made, tested, and then had to take back. We want you to follow the decisions along the way, not just the results.

The instrument is the **Beck Depression Inventory (BDI-II)**: 21 symptoms (sadness, guilt, sleep, appetite, …), each scored 0 (absent) to 3 (severe). Clinicians add the 21 items into one number, call a 15 "mild depression," and move on. This study asks what that number leaves out.

## How Much Does a Sum Score Hide?

Start with pure arithmetic - no data and no model, just counting. How many different 21-item profiles produce a total of exactly 15? The answer comes from the **inclusion–exclusion principle**, and it is worth computing yourself because the number is larger than most people expect:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch27-bdi`. For example, `book_data("ch27-bdi")` in R, Julia, or Python, or `!bookdata name = "ch27-bdi".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

n_patterns <- function(k, a, n) {                 # k items, each 0..a, summing to n
  imax <- floor(n / (a + 1))
  map_dbl(0:imax, \(i)
    (-1)^i * choose(k, i) * choose(n - i * (a + 1) + k - 1, k - 1)) |>
    sum()
}

format(n_patterns(21, 3, 15), big.mark = ",")     # ways to score 15 on the BDI-II
```
### SPSS
```
* How many 21-item profiles (each item 0..3) sum to exactly 15?
* Inclusion-exclusion, evaluated as a running total.
DATA LIST FREE /dummy.
BEGIN DATA
1
END DATA.

COMPUTE k = 21.
COMPUTE a = 3.
COMPUTE n = 15.
COMPUTE total = 0.
LOOP #i = 0 TO TRUNC(n / (a + 1)).
  COMPUTE total = total + ((-1) ** #i)
    * (EXP(LNGAMMA(k+1) - LNGAMMA(#i+1) - LNGAMMA(k-#i+1)))
    * (EXP(LNGAMMA(n - #i*(a+1) + k) - LNGAMMA(k) - LNGAMMA(n - #i*(a+1) + 1))).
END LOOP.
EXECUTE.
LIST VARIABLES=total.
```
### Julia
```julia
function n_patterns(k, a, n)          # k items, each 0..a, summing to n
    sum((-1)^i * binomial(k, i) * binomial(n - i*(a+1) + k - 1, k - 1)
        for i in 0:fld(n, a + 1))
end

n_patterns(21, 3, 15)                 # ways to score 15 on the BDI-II
```
### Python
```python
from math import comb

def n_patterns(k, a, n):              # k items, each 0..a, summing to n
    return sum((-1)**i * comb(k, i) * comb(n - i*(a+1) + k - 1, k - 1)
               for i in range(n // (a + 1) + 1))

f"{n_patterns(21, 3, 15):,}"          # ways to score 15 on the BDI-II
```
:::

**1.65 billion** distinct symptom profiles all land on "15 = mild depression." Across all possible totals there are $4^{21} \approx 4.4$ trillion profiles, peaking at 338 billion different ways to score a 31. The clinical convention - "minimal / mild / moderate / severe" - treats all of them as the same. That is a large amount of discarded information, and the question is whether any of it mattered.

## Same Score, Different People

::: {.callout-note}
## A note on the data
The clinical and student profiles behind this project are the authors' own research collections and are **not public**. So this chapter illustrates the *mechanics* on a **simulated** BDI-II-like sample - 21 items scored 0–3, generated from a single latent severity dimension so the symptoms covary the way real ones do - while every *finding* it quotes comes from the real six-sample analysis. Watch the machinery here; trust the numbers to the published study.
:::

The billion is abstract; two people make it concrete. First we build the simulated sample, then pull two who **both scored exactly 15**:

::: {.panel-tabset}
### R
```{r}
#| message: false
set.seed(2026)
N <- 1200; k <- 21
symptoms <- c("sadness","pessimism","past_failure","loss_pleasure","guilt","punishment",
              "self_dislike","self_critical","suicidal","crying","agitation","loss_interest",
              "indecision","worthlessness","loss_energy","sleep","irritability","appetite",
              "concentration","tiredness","loss_sex_interest")
theta <- rnorm(N, mean = -0.5)                # latent severity (one per person)
diff  <- sort(rnorm(k, mean = 2.6, sd = 1.3)) # item "difficulty": most symptoms are rare

bdi <- map(seq_len(k), \(j) {
  p <- plogis(outer(theta, diff[j] + c(-1.1, 0, 1.1), `-`))   # P(score >= 1, 2, 3)
  rowSums(runif(N) < p)                                       # a graded 0..3 response
}) |>
  set_names(symptoms) |>
  as_tibble()

totals <- rowSums(bdi)

pair <- which(totals == 15)[c(1, 2)]          # two people, identical total of 15

bdi |>
  slice(pair) |>
  mutate(person = c("person A", "person B"), .before = 1) |>
  pivot_longer(-person, names_to = "symptom", values_to = "score") |>
  pivot_wider(names_from = person, values_from = score)
```
### SPSS
```{pspp}
* Simulate 1200 people answering 21 graded (0..3) BDI-II-like items from a
* single latent severity dimension, then pull two who both scored 15.
* (Item names s1 to s21; theta is the person's latent severity.)
INPUT PROGRAM.
LOOP #p = 1 TO 1200.
  COMPUTE theta = RV.NORMAL(-0.5, 1).
  VECTOR s(21).
  LOOP #j = 1 TO 21.
    COMPUTE #d = 2.6 + 1.3 * IDF.NORMAL(#j / 22, 0, 1).
    COMPUTE s(#j) = (RV.UNIFORM(0,1) < (1/(1+EXP(-(theta - (#d - 1.1))))))
                  + (RV.UNIFORM(0,1) < (1/(1+EXP(-(theta - #d)))))
                  + (RV.UNIFORM(0,1) < (1/(1+EXP(-(theta - (#d + 1.1)))))).
  END LOOP.
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
COMPUTE total = SUM(s1 TO s21).
EXECUTE.

TEMPORARY.
SELECT IF (total = 15).
LIST VARIABLES=s1 TO s21 /CASES=2.
```
### Julia
```{julia}
using CSV, DataFrames
df = CSV.read("data/sim/ch27-bdi.csv", DataFrame)

using Random, Distributions, DataFrames
Random.seed!(2026)
N, k = 1200, 21
symptoms = ["sadness","pessimism","past_failure","loss_pleasure","guilt",
            "punishment","self_dislike","self_critical","suicidal","crying",
            "agitation","loss_interest","indecision","worthlessness",
            "loss_energy","sleep","irritability","appetite","concentration",
            "tiredness","loss_sex_interest"]

theta = rand(Normal(-0.5, 1), N)
diff  = sort(rand(Normal(2.6, 1.3), k))

bdi = DataFrame([Symbol(symptoms[j]) =>
        [sum(rand() < 1/(1+exp(-(t - (diff[j] + s)))) for s in (-1.1, 0, 1.1))
         for t in theta] for j in 1:k])

totals = sum(eachcol(bdi))
pair = findall(==(15), totals)[1:2]
bdi[pair, :]
```
### Python
```{python}
import pandas as pd
df = pd.read_csv("data/sim/ch27-bdi.csv")

import numpy as np, pandas as pd

rng = np.random.default_rng(2026)
N, k = 1200, 21
symptoms = ["sadness","pessimism","past_failure","loss_pleasure","guilt",
            "punishment","self_dislike","self_critical","suicidal","crying",
            "agitation","loss_interest","indecision","worthlessness",
            "loss_energy","sleep","irritability","appetite","concentration",
            "tiredness","loss_sex_interest"]

theta = rng.normal(-0.5, 1, N)
diff  = np.sort(rng.normal(2.6, 1.3, k))

def item(j):                                   # P(score >= 1, 2, 3)
    p = 1 / (1 + np.exp(-(theta[:, None] - (diff[j] + np.array([-1.1, 0, 1.1])))))
    return (rng.random((N, 3)) < p).sum(axis=1)

bdi = pd.DataFrame({s: item(j) for j, s in enumerate(symptoms)})
totals = bdi.sum(axis=1)

pair = totals[totals == 15].index[:2]           # two people, identical total
bdi.loc[pair].T
```
:::

One person's distress concentrates in a few symptoms; the other's spreads thin across many. Same "15," but different clinical pictures, and a sum score cannot tell them apart. The **profile** can.

## The Manifold Collapse

If 4.4 trillion profiles are *possible*, how many actually *occur*? Almost none - because depression symptoms do not vary independently; they hang together. Count the distinct profiles:

::: {.panel-tabset}
### R
```{r}
key <- bdi |> unite("profile", everything(), sep = "-") |> pull(profile)

tibble(
  people            = nrow(bdi),
  distinct_profiles = n_distinct(key),
  pct_unique        = 100 * mean(!(duplicated(key) | duplicated(key, fromLast = TRUE)))
) |>
  round2()
```
### SPSS
```{pspp}
* The book's simulated BDI-II-like responses: 1200 people, 21 items scored
* 0-3. Pull two people who both scored exactly 15.
INSERT FILE='data/sim/ch27-bdi.sps'.

COMPUTE total = SUM(sadness TO loss_sex_interest).
EXECUTE.

TEMPORARY.
SELECT IF (total = 15).
LIST VARIABLES=sadness pessimism past_failure loss_pleasure guilt /CASES=2.
```
### Julia
```{julia}
using CSV, DataFrames
bdi = CSV.read("data/sim/ch27-bdi.csv", DataFrame)

totals = sum(eachcol(bdi))
pair = findall(==(15), totals)[1:2]     # two people, identical total of 15
bdi[pair, :]
```
### Python
```{python}
import pandas as pd
bdi = pd.read_csv("data/sim/ch27-bdi.csv")

totals = bdi.sum(axis=1)
pair = totals[totals == 15].index[:2]   # two people, identical total of 15
bdi.loc[pair].T
```
:::

Roughly **three-quarters have a profile no one else shares**, and in the real six-sample analysis it was **79.7%**. Even our simple simulation reproduces the effect: of the trillions of mathematically possible patterns, people occupy only a small fraction, what the study called the **manifold collapse**. The space of possible profiles is huge, but real people use very little of it. That raises a genuine question: are the profiles that *never* appear merely rare, or are some of them *impossible in principle*, incoherent given how depression actually works?

## Antitypes: Profiles That Shouldn't Exist

A profile can be numerically fine and clinically nonsensical - maximum *worthlessness* with zero *sadness*, say. Borrowing a term from Configural Frequency Analysis, we call these **antitypes**: possible on paper, absent in reality because they violate the internal structure of the construct. To catch them we need a notion of which symptoms are "easy" (commonly endorsed) and which are "hard" (rare), and then flag people who endorse hard symptoms while skipping easy ones. Here is the idea in miniature - a light **ordering-violation** count you can run:

::: {.panel-tabset}
### R
```{r}
# higher mean = more commonly endorsed = "easier"
difficulty <- bdi |> summarise(across(everything(), mean)) |> unlist()

# count "harder symptom scored above an easier one" pairs, person by person
ovi <- pmap_dbl(bdi, \(...) {
  r <- c(...)
  sum(outer(seq_along(r), seq_along(r),
            \(i, j) difficulty[i] < difficulty[j] & r[i] > r[j]))
})

tibble(
  median_violations = median(ovi),
  max_violations    = max(ovi),
  cor_with_total    = cor(ovi, totals)    # NOT just severity
) |>
  round2()
```
### SPSS
```{pspp}
* How many distinct 21-item profiles actually occur among 1200 people?
INSERT FILE='data/sim/ch27-bdi.sps'.

STRING profile (A63).
COMPUTE profile = CONCAT(STRING(sadness,F1), STRING(pessimism,F1),
                         STRING(past_failure,F1), STRING(loss_pleasure,F1)).
EXECUTE.

AGGREGATE OUTFILE=* MODE=ADDVARIABLES /BREAK=profile /n_sharing=N.
FREQUENCIES VARIABLES=n_sharing.
```
### Julia
```{julia}
using CSV, DataFrames, StatsBase
bdi = CSV.read("data/sim/ch27-bdi.csv", DataFrame)

key = [join(r, "-") for r in eachrow(bdi)]
counts = countmap(key)

nrow(bdi), length(counts), 100 * mean(counts[k] == 1 for k in key)
```
### Python
```{python}
import pandas as pd
bdi = pd.read_csv("data/sim/ch27-bdi.csv")

key = bdi.astype(str).agg("-".join, axis=1)

len(bdi), key.nunique(), 100 * (key.map(key.value_counts()) == 1).mean()
```
:::

Incoherence is its own dimension - only weakly tied to how *high* the total is. The real study replaces this rough count with a proper **Graded Response Model** (an IRT model, cousin of the [Rasch model](18-rasch.qmd)) that estimates each item's difficulty and each person's severity, then scores every profile's log-likelihood. The two measures - our Ordering Violation Index (OVI) and the GRM profile log-likelihood - agree strongly, and that agreement is stable across samples:

```r
# The real analysis (needs the mirt package — shown, not run on the book's server)
library(mirt)
fit  <- mirt(bdi, 1, itemtype = "graded")           # fit the Graded Response Model
theta <- fscores(fit)                               # each person's estimated severity
ll    <- personfit(fit)$logLik                      # how probable each profile is
cor(ovi, ll)                                        # OVI and GRM-likelihood track each other
```

Across all six independent samples, the OVI–likelihood correlation held between **−0.73 and −0.83**, and the model flagged **13 strong antitypes** in a pooled $n = 1{,}479$. Incoherent profiles are not a fluke of one dataset; they are a stable, measurable property.

Finding structure that a sum score hides is a recurring theme in the authors' work. In a related vein, Kashdan, McKnight, and colleagues [@KashdanMcKnightRichey2009] used **latent class analysis** on the National Comorbidity Survey to uncover a neglected, risk-prone *subtype* of social-anxiety disorder - a whole subgroup of people invisible to anyone who only looked at total symptom counts.

## Does It Actually Matter? (The Decision to Chase It)

A finding earns a place in a paper when it *predicts* something. In 207 patients tracked across four therapy sessions, we asked whether the **coherence of a patient's intake profile** forecast how much they improved, over and above their initial severity. It did: more incoherent intake profiles improved *less* ($p = .017$), controlling for baseline depression. And treatment did not merely lower the total; it **reorganized** the profile (cognitive and emotional symptoms improved ~20%; somatic symptoms barely moved). The *shape* of depression changed, not just its size. The information the sum score discarded was clinically meaningful.

## Replication - and a Claim We Had to Retract

Two habits separate research from storytelling: **replicate**, and **try hard to disprove your own finding**. We did both here.

**Replication.** We re-ran the entire pipeline on a completely different scale - the **CES-D**, 20 different items - in a completely different population (a multigenerational community panel). No shared items, no shared people. Same result: 78.7% unique profiles (BDI: 79.7%), OVI–likelihood $r = -0.77$. The manifold collapse is a property of *depression*, not of Beck's particular questionnaire.

**The retraction.** Encouraged, we noticed the two scales reproduced nearly the *same* configural geometry (difficulty spread, profile uniqueness, incoherence) at a structural concordance of about 0.95, despite sharing nothing. We were ready to call it convergent validity: two windows onto the same construct.

Then we ran the control that could undermine the claim. We took an **anxiety** scale of the same 0–3 format (the STICSA), a *different construct*, and measured its concordance with the depression scales.

::: {.callout-important}
## When a good control undoes a good story
The anxiety scale concorded *just as highly* as the two depression scales did with each other. If a different construct fits the pattern equally well, the pattern is **not** about the construct; it is about the shared **response format** (21-ish items, 0–3, summed). Our claim of convergent validity did not hold up. The concordance is real, but it is a format-general regularity, not evidence that the scales read the same thing. We cut the claim.

This is one of the most important moves in statistics: a **discriminant test**, deliberately giving your hypothesis a chance to fail by showing something it *shouldn't* explain that it explains anyway. We wanted the convergent-validity story. The data said no. In a textbook you usually see only the claims that survived; here you have seen one that did not. That is a normal and healthy part of the work.
:::

## What This Case Teaches

- **The mean/sum discards structure** ([Ch. 3](03-CTmeasures.qmd)): a total score is a lossy compression, and sometimes the lost bits are the clinically important ones.
- **IRT recovers the structure** ([Ch. 18](18-rasch.qmd)): a Graded Response Model turns "which profiles are coherent?" into a computable question.
- **Prediction is the test** ([Ch. 20](20-beyond.qmd)): the finding earned its keep by forecasting treatment response, not by fitting the past.
- **Discriminant tests protect you from yourself** ([Ch. 8](08-hypothesistesting.qmd)): the discipline that retracted our convergent-validity claim is the same discipline that makes any claim trustworthy.

The next case study builds a measure from scratch, study by study - and shows what it looks like to preregister a prediction *before* the data arrive.
