# Case Study: When the Total Score Hides the Story {#depression}

Our first real study is one of our own - a collaboration (McKnight, Stuewig, Kashdan) that started from a complaint you have already heard in this book: **the mean throws away information.** Back in the [central tendency chapter](03-CTmeasures.qmd) that was a slogan. Here it becomes a research program, with real patients, a real depression scale, a surprising finding, and - the part you rarely see in a polished paper - a claim we made, tested, and had to take back. Watch the decisions, not just the results.

The instrument is the **Beck Depression Inventory (BDI-II)**: 21 symptoms (sadness, guilt, sleep, appetite, …), each scored 0 (absent) to 3 (severe). Clinicians add the 21 items into one number, call a 15 "mild depression," and move on. This study asks what that number destroys.

## How Much Does a Sum Score Hide?

Start with pure arithmetic - no data, no model, just counting. How many different 21-item profiles produce a total of exactly 15? The answer comes from the **inclusion–exclusion principle**, and it is worth computing yourself because the number is hard to believe:

```{r}
n_patterns <- function(k, a, n) {                 # k items, each 0..a, summing to n
  imax <- floor(n / (a + 1))
  sum(sapply(0:imax, function(i)
    (-1)^i * choose(k, i) * choose(n - i * (a + 1) + k - 1, k - 1)))
}
format(n_patterns(21, 3, 15), big.mark = ",")     # ways to score 15 on the BDI-II
```

**1.65 billion** distinct symptom profiles all land on "15 = mild depression." Across all possible totals there are $4^{21} \approx 4.4$ trillion profiles, peaking at 338 billion different ways to score a 31. The clinical convention - "minimal / mild / moderate / severe" - treats all of them as the same. That is a staggering amount of thrown-away information, and the question is whether any of it mattered.

## Same Score, Different People

The billion is abstract; two real students make it concrete. Here are two people from our student sample who **both scored exactly 15** - and could not look more different:

```{r}
bdi <- read.csv("data/bdi_students.csv")           # 1,246 real BDI-II profiles, item level
totals <- rowSums(bdi)
pair <- which(totals == 15)[c(1, 2)]
t(bdi[pair, ])                                      # two profiles, identical total
```

One student's distress concentrates in self-punishment, self-dislike, and concentration; the other's is spread thinly across crying, lost pleasure, appetite, and fatigue. Same "15," different human beings, different clinical pictures. A sum score cannot tell them apart. The **profile** can.

## The Manifold Collapse

If 4.4 trillion profiles are *possible*, how many actually *occur*? Almost none - because depression symptoms do not vary independently; they hang together. Count the distinct profiles in real data:

```{r}
key <- apply(bdi, 1, paste, collapse = "-")
c(people            = nrow(bdi),
  distinct_profiles = length(unique(key)),
  pct_unique        = round(100 * mean(!(duplicated(key) | duplicated(key, fromLast = TRUE))), 1))
```

About **76% of these students have a profile no one else in the sample shares** (in the full six-sample analysis it was 79.7%). Of trillions of mathematically possible patterns, real people occupy a vanishingly thin sliver - what we called the **manifold collapse**. The space is huge; life uses almost none of it. Which raises the sharp question: are the profiles that *never* appear merely rare, or are some of them *impossible in principle* - incoherent given how depression actually works?

## Antitypes: Profiles That Shouldn't Exist

A profile can be numerically fine and clinically nonsensical - maximum *worthlessness* with zero *sadness*, say. Borrowing a term from Configural Frequency Analysis, we call these **antitypes**: possible on paper, absent in reality because they violate the internal structure of the construct. To catch them we need a notion of which symptoms are "easy" (commonly endorsed) and which are "hard" (rare), and then flag people who endorse hard symptoms while skipping easy ones. Here is the idea in miniature - a light **ordering-violation** count you can run:

```{r}
difficulty <- colMeans(bdi)                         # higher = more commonly endorsed = "easier"
ovi <- apply(bdi, 1, function(r)                     # count "harder scored above easier" pairs
  sum(outer(seq_along(r), seq_along(r),
            function(i, j) difficulty[i] < difficulty[j] & r[i] > r[j])))
c(median_violations = median(ovi), max_violations = max(ovi),
  cor_with_total    = round(cor(ovi, totals), 2))    # NOT just severity
```

Incoherence is its own dimension - only weakly tied to how *high* the total is. The real study replaces this rough count with a proper **Graded Response Model** (an IRT model, cousin of the [Rasch model](18-rasch.qmd)) that estimates each item's difficulty and each person's severity, then scores every profile's log-likelihood. The two measures - our Ordering Violation Index (OVI) and the GRM profile log-likelihood - agree strongly, and that agreement is remarkably stable:

```r
# The real analysis (needs the mirt package — shown, not run on the book's server)
library(mirt)
fit  <- mirt(bdi, 1, itemtype = "graded")           # fit the Graded Response Model
theta <- fscores(fit)                               # each person's estimated severity
ll    <- personfit(fit)$logLik                      # how probable each profile is
cor(ovi, ll)                                        # OVI and GRM-likelihood track each other
```

Across all six independent samples, the OVI–likelihood correlation held between **−0.73 and −0.83**, and the model flagged **13 strong antitypes** in a pooled $n = 1{,}479$. Incoherent profiles are not a fluke of one dataset; they are a stable, measurable property.

## Does It Actually Matter? (The Decision to Chase It)

A finding is only worth a paper if it *predicts* something. In 207 patients tracked across four therapy sessions, we asked whether the **coherence of a patient's intake profile** forecast how much they improved - over and above their initial severity. It did: more incoherent intake profiles improved *less* ($p = .017$), controlling for baseline depression. And treatment did not merely lower the total - it **reorganized** the profile (cognitive and emotional symptoms improved ~20%; somatic symptoms barely moved). The *shape* of depression changed, not just its size. The information the sum score discarded was clinically live.

## Replication - and a Claim We Had to Retract

Two habits separate research from storytelling: **replicate**, and **try to kill your own finding**. We did both, and only one felt good.

**Replication.** We re-ran the entire pipeline on a completely different scale - the **CES-D**, 20 different items - in a completely different population (a multigenerational community panel). No shared items, no shared people. Same result: 78.7% unique profiles (BDI: 79.7%), OVI–likelihood $r = -0.77$. The manifold collapse is a property of *depression*, not of Beck's particular questionnaire.

**The retraction.** Emboldened, we noticed the two scales reproduced nearly the *same* configural geometry - difficulty spread, profile uniqueness, incoherence - at a structural concordance of about 0.95, despite sharing nothing. We were ready to call it convergent validity: two windows onto the same construct.

Then we ran the control that could destroy the claim. We took an **anxiety** scale of the same 0–3 format (the STICSA) - a *different construct* - and measured its concordance with the depression scales.

::: {.callout-important}
## Watch a good control kill a good story
The anxiety scale concorded *just as highly* as the two depression scales did with each other. If a different construct fits the pattern equally well, the pattern is **not** about the construct - it is about the shared **response format** (21-ish items, 0–3, summed). Our "convergent validity" evaporated. The concordance is real, but it is a format-general regularity, not evidence that the scales read the same thing. We cut the claim.

This is the least glamorous and most important move in statistics: a **discriminant test** - deliberately giving your hypothesis a chance to fail by showing something it *shouldn't* explain that it explains anyway. We wanted the convergent-validity story. The data said no. In a textbook you usually see only the surviving claims; here you saw one die. That is the job.
:::

## What This Case Teaches

- **The mean/sum discards structure** ([Ch. 3](03-CTmeasures.qmd)): a total score is a lossy compression, and sometimes the lost bits are the clinically important ones.
- **IRT recovers the structure** ([Ch. 18](18-rasch.qmd)): a Graded Response Model turns "which profiles are coherent?" into a computable question.
- **Prediction is the test** ([Ch. 20](20-beyond.qmd)): the finding earned its keep by forecasting treatment response, not by fitting the past.
- **Discriminant tests protect you from yourself** ([Ch. 8](08-hypothesistesting.qmd)): the discipline that retracted our convergent-validity claim is the same discipline that makes any claim trustworthy.

The next case study builds a measure from scratch, study by study - and shows what it looks like to preregister a prediction *before* the data arrive.
