# Case Study: Building a Measure, Study by Study {#curiosity}

The [last case study](24-casestudy-depression.qmd) took a finished scale apart. This one does the opposite - it *builds* one from nothing, and shows why that is far harder than writing a few questions and adding them up. The project (McKnight, Kashdan, and colleagues) set out to measure **political curiosity**: the disposition to seek out, and sit with, political information and views - including those you find uncomfortable. There is no ready-made instrument for that. You have to construct one, and then earn the right to trust it.

::: {.callout-note}
## A note on the data
As in the previous chapter, the study's real datasets are the authors' own research collections and are **not public**. Every runnable illustration here uses **simulated** data built to expose the mechanics; the study's actual findings are described in the text.
:::

The build follows a discipline worth memorizing, because it is how *every* good measure in psychology is made: **explore the structure, confirm it on fresh data, check it means the same thing for everyone, and predict something before you look.**

## Study 1 - Explore: How Many Things Are We Even Measuring?

You write a big pool of candidate items and collect responses. Now the first question: do these items measure *one* thing, or several? This is a job for **Exploratory Factor Analysis (EFA)** - a method that looks at how items correlate and infers the small number of hidden dimensions ("factors") that could produce those correlations.

The workhorse decision - *how many factors?* - starts with the **eigenvalues** of the item correlation matrix. Each eigenvalue is the amount of variance one potential factor accounts for; the classic rules keep factors with an eigenvalue above 1 (they explain more than a single item's worth) and look for the "elbow" in a scree plot. Watch it recover a structure we planted on purpose - twelve items secretly built from three factors:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch28-curiosity`. For example, `book_data("ch28-curiosity")` in R, Julia, or Python, or `!bookdata name = "ch28-curiosity".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.
:::

::: {.panel-tabset}
### R
```{r}
#| label: fig-scree
#| fig-cap: "A scree plot. Three eigenvalues stand above 1, then the scree falls to rubble — three factors."
#| fig-height: 3.4
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(2026)
N <- 500; load <- 0.75
factors <- list(F1 = rnorm(N), F2 = rnorm(N), F3 = rnorm(N))   # three hidden factors
mk <- function(f) load * f + rnorm(N, sd = sqrt(1 - load^2))

# four indicators per factor, twelve items in all
items <- factors[rep(c("F1", "F2", "F3"), each = 4)] |>
  map(mk) |>
  set_names(paste0("item", 1:12)) |>
  as_tibble()

ev <- eigen(cor(items))$values    # eigenvalues of the correlation matrix

tibble(factor_number = seq_along(ev), eigenvalue = ev) |>
  ggplot(aes(x = factor_number, y = eigenvalue)) +
  # Kaiser's "keep it if > 1" line
  geom_hline(yintercept = 1, linetype = "dashed", colour = "firebrick") +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue", size = 2.5) +
  labs(x = "Factor number", y = "Eigenvalue") +
  theme_book()
```
### SPSS
```
* Twelve items built from three hidden factors, then the scree plot.
* (Assumes item1 to item12 are already the active dataset.)
FACTOR
  /VARIABLES item1 TO item12
  /EXTRACTION PC
  /PRINT INITIAL EXTRACTION
  /PLOT EIGEN.
```
### Julia
```julia
using Random, Distributions, Statistics, LinearAlgebra, DataFrames, Plots
Random.seed!(2026)
N, load = 500, 0.75
F = [rand(Normal(), N) for _ in 1:3]                # three hidden factors
mk(f) = load .* f .+ rand(Normal(0, sqrt(1 - load^2)), N)

items = DataFrame([Symbol("item", i) => mk(F[cld(i, 4)]) for i in 1:12])
ev = eigvals(cor(Matrix(items))) |> reverse         # eigenvalues, largest first

plot(1:12, ev, marker = :circle, legend = false,
     xlabel = "Factor number", ylabel = "Eigenvalue")
hline!([1], linestyle = :dash, color = :firebrick)  # Kaiser's "keep it if > 1"
```
### Python
```python
import numpy as np, pandas as pd
import matplotlib.pyplot as plt

rng = np.random.default_rng(2026)
N, load = 500, 0.75
F = [rng.normal(size=N) for _ in range(3)]          # three hidden factors
mk = lambda f: load * f + rng.normal(0, np.sqrt(1 - load**2), N)

items = pd.DataFrame({f"item{i+1}": mk(F[i // 4]) for i in range(12)})
ev = np.sort(np.linalg.eigvals(items.corr()).real)[::-1]

plt.plot(range(1, 13), ev, marker="o", color="steelblue")
plt.axhline(1, linestyle="--", color="firebrick")   # Kaiser's "keep it if > 1"
plt.xlabel("Factor number"); plt.ylabel("Eigenvalue")
plt.show()
```
:::

::: {.panel-tabset}
### R
```{r}
# how many factors the data suggest
tibble(factors_above_1 = sum(ev > 1))
```
### SPSS
```
* How many factors the data suggest: count the eigenvalues above 1.
* SPSS reports this directly in the FACTOR output above - the number of
* components retained under the default Kaiser (eigenvalue > 1) rule.
```
### Julia
```julia
count(>(1), ev)                 # how many factors the data suggest
```
### Python
```python
(ev > 1).sum()                  # how many factors the data suggest
```
:::

Three, exactly as planted. The real study did this on its item pool and found a coherent multi-dimensional structure of political curiosity. In practice you would rotate and interpret the factors with a dedicated tool:

```r
# The real EFA (needs the psych package — shown, not run here)
library(psych)
fa.parallel(items)              # parallel analysis: a better "how many factors?" than Kaiser alone
efa <- fa(items, nfactors = 3, rotate = "oblimin")   # extract & rotate to interpret
efa$loadings                    # which items load on which factor
```

But finding a structure in your data is the *easy* part. Any rich enough dataset will hand you a tidy factor solution if you go looking. Remember that fitting the past is not the same as predicting the future. The real test comes next.

## Study 2 - Confirm: On Data You Have Not Seen

The cardinal rule that separates measurement from wishful thinking is straightforward: **you do not confirm a structure on the same data that suggested it.** EFA is exploratory, and it will always find *something*. To believe the structure is real, you fit it as a fixed hypothesis to a **new, independent sample** using **Confirmatory Factor Analysis (CFA)**, and ask whether the pre-specified model actually fits.

```r
# The real CFA on Study 2's fresh sample (needs lavaan — shown, not run here)
library(lavaan)
model <- '
  curiosity  =~ item1 + item2 + item3 + item4
  tolerance  =~ item5 + item6 + item7 + item8
  engagement =~ item9 + item10 + item11 + item12
'
fit <- cfa(model, data = study2_data)
fitMeasures(fit, c("cfi", "tli", "rmsea", "srmr"))
```

CFA does not ask "what structure is here?" - it asks "does *this* structure hold?" and answers with **fit indices** you judge against conventional standards (Hu & Bentler): **CFI and TLI ≥ .95**, **RMSEA ≤ .06**, **SRMR ≤ .08**. The political-curiosity model met those standards on the independent Study 2 sample - which is what earns the measure the right to be used. Explore once, confirm elsewhere; a structure that survives a fresh sample is a structure worth believing.

## Does the Ruler Mean the Same Thing for Everyone?

Before you compare groups on a measure - do liberals and conservatives differ in political curiosity? - you must first check that the measure means the *same thing* in both groups. If "curiosity" items behave differently for conservatives than for liberals, then a group difference in scores might be an artifact of the **ruler**, not a real difference in the trait. This is **measurement invariance**, tested with multi-group CFA that progressively constrains the model to be equal across groups:

```r
# Measurement invariance across political groups (lavaan — shown, not run here)
configural <- cfa(model, data = dat, group = "conservative")            # same shape?
metric     <- cfa(model, data = dat, group = "conservative",
                  group.equal = "loadings")                             # same loadings?
scalar     <- cfa(model, data = dat, group = "conservative",
                  group.equal = c("loadings", "intercepts"))            # same intercepts?
anova(configural, metric, scalar)   # does constraining hurt fit? if not, invariance holds
```

The study ran exactly this before comparing political groups - because a group comparison on a non-invariant scale is comparing inches to centimeters and calling the difference meaningful. It is the same lesson the [depression study](24-casestudy-depression.qmd) learned the hard way: *make sure two measurements are on the same footing before you read anything into their difference.*

## Preregister: Call Your Shot Before the Data Arrive

With the measure built and validated, the team turned to a prediction: an "optimal distinctiveness" intervention would raise political curiosity. And here they did one of the most credibility-earning things you can do in empirical science - they **preregistered**: wrote down the hypothesis, the design, the sample size, and the exact analysis *before collecting a single response*, and time-stamped it publicly (Studies 3 and 4 carry preregistrations in the project's OSF archive).

::: {.callout-important}
## Why preregistration is the same idea as the accountability pledge
Recall the [Sports Page rule](23-sportspage.qmd): fitting the past is easy, so the only honest test is predicting the future and admitting when you were wrong. Preregistration puts that rule on the record. Without it, a researcher who peeks at the data can, consciously or not, tune the hypothesis and the analysis until something is "significant," fitting the past and calling it a discovery. Preregistration ties your hands *before* you see the outcome, so a confirmed prediction actually means something. It turns a study from "we found a pattern" (description dressed as discovery) into "we predicted it, and it happened" (a real forecast that could have failed). A preregistered study that *fails* is worth more than an unregistered one that "succeeds."
:::

The confirmatory, preregistered tests - including a **mediation** analysis asking *why* the intervention worked (through what psychological mechanism) - are the payoff of the whole arc. But note the order: the measure had to be **built and trusted** (Studies 1–2) before any intervention result meant anything. You cannot detect a change in a thing you cannot yet measure.

## What This Case Teaches

- **Measures are built, not assumed** ([reliability](10-reliability.qmd), [validity](11-validity.qmd)): a sum of items is a hypothesis about structure, and the hypothesis has to be earned.
- **Explore, then confirm - on fresh data.** EFA finds a structure; CFA on an independent sample tests whether it was real. Confirming on your exploration sample is the past cheating again.
- **Check invariance before you compare groups.** A difference on a non-invariant scale can be an artifact of the instrument.
- **Preregistration turns a pattern into a prediction** ([Ch. 23](23-sportspage.qmd)): call your shot before the data, and a hit finally counts.

The final case study takes on the messiest reality of all - the data you *didn't* get - and asks whether the holes in a dataset can quietly change the answer.
