# Case Study: The Data You Didn't Get {#missing}

The last case study takes on the part of every real dataset that textbooks often set aside: the **holes**. People drop out of a two-year study. Respondents skip the touchy item. A sensor fails. In the purpose-versus-happiness project - a three-wave longitudinal study tracking hundreds of people over two years - attrition was not a footnote; it was a genuine decision point, because *how you handle the missing data can change the answer.*

We have a slight advantage here: one of us wrote the book on it - literally [@McKnight2007], *Missing Data: A Gentle Introduction* (Guilford). The one-sentence version of that book is the thesis of this chapter: **missing data is not a nuisance to be swept aside; it is a modeling problem, and the first question is always *why* the data are missing.**

::: {.callout-note}
## A note on the data
As with the other case studies, the study's real data are the authors' private collection. The mechanics below run on **simulated** data with a missingness mechanism we control on purpose, so you can watch a hole bend an estimate; the real study's findings are described in the text.
:::

## First Ask *Why* It's Missing

Recall the three mechanisms from the [data-cleaning chapter](05-datacleaning.qmd) - they matter more than any particular technique:

- **MCAR** (missing completely at random): the holes are unrelated to anything. Rare, and the only case where throwing away incomplete rows is merely *wasteful*, not *biased*.
- **MAR** (missing at random): missingness depends on things you *observed*. Fixable - if you use the observed information.
- **MNAR** (missing not at random): missingness depends on the *unobserved value itself* (people with the highest depression skip the depression item). The hardest case; no method fully saves you, and honesty about it is the best you can do.

The danger is that the most common default - **complete-case analysis** (drop anyone with a missing value) - is only safe under MCAR, and MCAR is the least common mechanism in real life. Watch what MAR does to a simple estimate.

## Watch a Hole Bend an Estimate

Simulate an outcome `y` that correlates with an auxiliary variable `z`, then make `y` **missing more often when `z` is high** - a textbook MAR mechanism (the missingness depends on `z`, which we *observe*):

```{r}
set.seed(2026)
N <- 2000
z <- rnorm(N)                              # an auxiliary variable, fully observed
y <- 0.6 * z + rnorm(N, sd = 0.8)          # outcome; its true mean is ~ 0
p_miss <- plogis(1.8 * z - 0.2)            # P(missing) rises with z  -> MAR
y_obs  <- ifelse(runif(N) < p_miss, NA, y) # the data we actually get to see

c(pct_missing         = round(100 * mean(is.na(y_obs)), 1),
  true_mean           = round(mean(y), 3),                 # unobtainable in real life
  complete_case_mean  = round(mean(y_obs, na.rm = TRUE), 3))
```

The complete-case mean is badly **biased downward**. That makes sense: we systematically dropped the high-`z` people, and high `z` means high `y`, so the survivors are not representative. The hole did more than cost us sample size; it *moved the answer*.

Now recover it. The core idea of imputation is to **use the observed information** (`z`) to fill the holes in `y` - here with a stochastic regression imputation (predict the missing `y` from `z`, and add noise so we don't fake certainty):

```{r}
cc  <- !is.na(y_obs)
fit <- lm(y_obs[cc] ~ z[cc])                               # learn y-from-z on complete cases
imp <- y_obs
imp[!cc] <- coef(fit)[1] + coef(fit)[2] * z[!cc] +         # predicted value ...
            rnorm(sum(!cc), 0, sigma(fit))                 # ... plus honest noise
round(mean(imp), 3)                                        # recovered — back near the true 0
```

Filling the holes with what `z` implies pulls the estimate back to the truth. That is the entire logic of principled missing-data handling: the information to fix the bias was *in the data we kept*, and complete-case analysis simply threw it away.

## The Twist: It Depends on What You're Estimating

There is a subtlety here worth slowing down for. The *marginal mean* of `y` was wrecked, but watch the *regression slope* of `y` on `z`:

```{r}
c(slope_full          = round(coef(lm(y ~ z))[2], 3),      # the truth
  slope_complete_case = round(coef(lm(y_obs ~ z))[2], 3))  # barely moved
```

The slope is essentially fine, even under 47% MAR missingness - because the missingness depends on `z`, and `z` is already *in* the model, so conditioning on it absorbs the mechanism. **Same data, same holes: the mean is biased, the slope is not.** Whether the missing data changes your answer depends on *which* answer you want and *how* the holes relate to your model. "Does missing data matter here?" is a real question with a real, case-by-case answer - not a reflex to reach for either complete-case or imputation.

## Multiple Imputation: The Honest Tool

Our single imputation recovered the mean, but it glossed over one thing: it treated the filled-in values as if they were real, understating our uncertainty. The fix is **multiple imputation** - impute the holes several times (each with fresh noise), analyze each completed dataset, and **pool** the results with Rubin's rules, so the between-imputation variability is folded into the standard errors. In R that is the `mice` package:

```r
# The real tool (needs the mice package — shown, not run on the book's server)
library(mice)
imp  <- mice(data, m = 20, method = "pmm", printFlag = FALSE)  # 20 imputations
fit  <- with(imp, lm(happiness ~ purpose + wave))              # analyze each
pool(fit)                                                       # pool via Rubin's rules
```

The purpose-versus-happiness study did exactly this - fitting its growth models under **complete-case and under multiple imputation side by side** - and reported both, so a reader could see whether the two-year conclusions survived the missing data. That comparison *is* the analysis; reporting only the complete-case result would have meant leaving the question unanswered.

## What This Case Teaches

- **Missing data is a modeling problem** [@McKnight2007], not a cleaning chore. The first question is the *mechanism* ([Ch. 5](05-datacleaning.qmd)): MCAR, MAR, or MNAR.
- **Complete-case analysis is a decision, not a default** - safe only under MCAR, and quietly biased under the far more common MAR.
- **Imputation works by using what you kept.** Auxiliary variables that predict both the outcome and its missingness are gold.
- **Whether it changes the answer is estimand-specific.** Check; don't assume. And when you handle it, use *multiple* imputation so your uncertainty tells the truth.

That closes our tour of four real studies, with the judgment calls left in view: a total score caught hiding information, a measure built and confirmed and preregistered, and a dataset's holes that bent one estimate and left another alone. The tools themselves were rarely the hard part. The harder work is knowing which tool to use, when, and why, and being honest about what you cannot know.
