# Latent Variables: Confirmatory Factor Analysis and SEM {#sem}

Every model so far has related variables we can *see* - a test score, a reaction time, a GPA. But the things social scientists most want to study cannot be seen at all. Nobody has ever directly observed *depression*, *curiosity*, or *intelligence*. We observe only their shadows: answers to questionnaire items, behaviors, symptoms. A construct is **latent** - inferred from many imperfect indicators, never measured directly.

**Structural Equation Modeling (SEM)** is the framework built for exactly this: it models latent variables and the relationships among them. Its measurement half - **Confirmatory Factor Analysis (CFA)** - is how you establish that a set of indicators really does reflect one underlying construct. This is home turf for the authors, whose published work leans heavily on factor-analytic and covariance-structure methods; this chapter shows the machinery underneath.

## Learning Objectives

1.  State the **reflective measurement model**: a latent factor as the common cause of its indicators.
2.  Fit and read a factor analysis, and understand loadings and uniquenesses.
3.  See that fitting a factor model means matching a **model-implied covariance matrix** to the observed one.
4.  Understand how SEM disattenuates relationships by modeling measurement error - and where it connects to [reliability](10-reliability.qmd).

## The Reflective Measurement Model

Why do the items on a good depression scale correlate with one another? The reflective model's answer is causal: they share a **common cause** - the latent factor. Depression makes you sadder, *and* less energetic, *and* more pessimistic, so those items rise and fall together. The factor is the reason for the correlations; each indicator is a noisy reflection of it:

$$x_i = \lambda_i \, \xi + \varepsilon_i$$

where $\xi$ is the latent factor, $\lambda_i$ is the **loading** (how strongly item $i$ reflects the factor), and $\varepsilon_i$ is that item's unique part (its reliability gap plus whatever else is specific to it). This is just the regression idea from [earlier chapters](14-bivariateregression.qmd) - but the predictor is *unobserved*.

## Factor Analysis, Executable

R's base `factanal()` fits a factor model by maximum likelihood - no extra packages. Build eight items secretly generated by **two** correlated factors (four indicators each), and watch the analysis recover the structure blind:

```{r}
set.seed(2026)
N <- 400; L <- 0.75
F1 <- rnorm(N)
F2 <- 0.5 * F1 + rnorm(N, sd = sqrt(1 - 0.25))          # two factors, correlated ~ .5
mk <- function(f) L * f + rnorm(N, sd = sqrt(1 - L^2))  # an indicator = loading*factor + noise
X <- data.frame(a1 = mk(F1), a2 = mk(F1), a3 = mk(F1), a4 = mk(F1),
                b1 = mk(F2), b2 = mk(F2), b3 = mk(F2), b4 = mk(F2))

fa <- factanal(X, factors = 2, rotation = "promax")     # ML factor analysis (base R)
round(unclass(fa$loadings), 2)
```

The pattern is unmistakable: `a1`–`a4` load on one factor (~0.75), `b1`–`b4` on the other, and the cross-loadings are essentially zero. The analysis reconstructed the hidden blueprint from nothing but the correlations. And the two factors are themselves related, just as we built them:

```{r}
round(fa$correlation[1:2, 1:2], 2)   # estimated correlation between the two factors
```

## What "Fit" Actually Means: The Model-Implied Covariance

Here is the engine of all of SEM, and it is simpler than its reputation. A factor model does not just describe the data - it **predicts the entire covariance matrix**. If each item is $\lambda_i \xi + \varepsilon_i$, then the model *implies* that the correlation matrix must equal $\Lambda\Lambda^{\top} + \Psi$ (loadings times loadings, plus the unique variances on the diagonal). **Fitting the model means finding loadings that make this implied matrix match the observed one.** Watch it work on a single-factor block:

```{r}
fa1     <- factanal(X[, 1:4], factors = 1)
lambda  <- as.numeric(fa1$loadings)          # the four loadings
psi     <- fa1$uniquenesses                  # each item's unique variance
implied <- lambda %*% t(lambda) + diag(psi)  # the covariance the model PREDICTS

max(abs(cor(X[, 1:4]) - implied))            # how far off is the prediction?
```

The largest discrepancy between the observed correlations and the ones the model *predicts* is about 0.02 - a near-perfect reproduction. That gap, summarized across the whole matrix, is exactly what **fit indices** measure. A model that reproduces the observed covariances well fits well; one that cannot is rejected. This is why CFA is *confirmatory*: you specify the structure, and the data get to say whether that structure could have produced the correlations you actually see.

::: {.callout-important}
## EFA explores, CFA confirms - and fit has thresholds
[Exploratory factor analysis](25-casestudy-curiosity.qmd) asks "how many factors, and what loads where?" - it discovers structure. **CFA** fixes the structure in advance (item 1 loads on factor A, not B) and tests it, ideally on a *fresh* sample. You judge the model-implied-vs-observed gap against conventional standards (Hu & Bentler): **CFI/TLI ≥ .95**, **RMSEA ≤ .06**, **SRMR ≤ .08**. Good fit is the model earning the right to be believed.
:::

## From Measurement to Structure

CFA is the measurement half. The "structural" in SEM is the other half: once you have clean latent variables, you can regress them *on each other* - does latent curiosity predict latent well-being? - with the measurement model and the structural model estimated **together**. For that you reach for a dedicated engine like `lavaan`:

```r
# A full SEM: measurement model + structural paths (needs lavaan — shown, not run here)
library(lavaan)
model <- '
  # measurement model (CFA)
  curiosity  =~ a1 + a2 + a3 + a4
  wellbeing  =~ b1 + b2 + b3 + b4
  # structural model (a regression among latent variables)
  wellbeing ~ curiosity
'
fit <- sem(model, data = X)
summary(fit, fit.measures = TRUE, standardized = TRUE)
```

The path `wellbeing ~ curiosity` is an ordinary regression - but between two variables neither of which was ever observed, each purified of its measurement error first.

## Why Bother? SEM Fixes Attenuation

Recall from the [reliability chapter](10-reliability.qmd) that unreliable measures **attenuate** correlations - a real relationship looks weaker than it is because the sum scores are noisy. Regressing one sum score on another inherits that bias; the slope is pulled toward zero by measurement error. SEM's quiet superpower is that it **separates the true score from the error before estimating the relationship**. The regression among latent variables is *disattenuated* - it estimates the relationship between the constructs themselves, not between their noisy stand-ins. That is the entire reason to pay SEM's price in complexity: it gives you the relationship you actually wanted, the one between the things, not between the rulers.

::: {.callout-note}
## From the Field
The authors' factor-analytic and covariance-structure work runs throughout their research - the [political-curiosity measure](25-casestudy-curiosity.qmd) was built with exactly this EFA→CFA→invariance pipeline. And Voelkle and McKnight [@VoelkleMcKnight2012] showed by Monte-Carlo simulation that classic repeated-measures (M)ANOVA is itself a **latent curve model** in disguise - the same "everything is one model" lesson from the [GLM chapter](16-ANOVAandGLM.qmd), pushed all the way into structural models. The [methods showcase](27-methods-showcase.qmd) lists more.
:::

## What This Chapter Teaches

- **Constructs are latent**; we model them through multiple imperfect indicators, not a single score.
- **A factor is a common cause** of its indicators - which is why they correlate, and why the loadings mean something.
- **Fitting = matching a model-implied covariance** ($\Lambda\Lambda^{\top}+\Psi$) to the observed one. Fit indices measure the gap.
- **SEM disattenuates**: by modeling measurement error, it estimates relationships among *constructs*, correcting the bias that sinks sum-score regression.

Factor models find *dimensions* along which people vary. The [next chapter](29-latentclass.qmd) asks a different question - not "where does each person fall on a continuum?" but "which *kind* of person is this?" - and meets the latent-class family of models.
