# Latent Classes: Which Kind of Person Is This? {#latentclass}

Factor analysis, in the [last chapter](28-sem.qmd), asked a *dimensional* question: **where** does each person fall on a continuum of depression, or curiosity? Latent **class** analysis (LCA) asks a categorical one: **which kind** of person is this? Instead of a score along an axis, it looks for hidden *types* - subgroups that respond in qualitatively different ways. A dimension sorts people high to low; a class says they come in distinct *kinds*.

This matters because some of the most important findings in psychology are subgroups that a single score would blur together. A total symptom count cannot tell you that some anxious people are also **risk-prone approachers** rather than shy avoiders - but a latent-class model can, and the authors' own work found exactly that (more below).

## Learning Objectives

1.  Contrast a **dimensional** latent variable (a factor) with a **categorical** one (a class).
2.  State the latent-class model as a **mixture** of response patterns.
3.  Fit one from scratch with the **EM algorithm**, and read the classes off the item probabilities.
4.  Recognize that classes can differ in *pattern*, not just *severity* - which is the whole point.

## The Model: A Mixture of Kinds

Suppose there are $K$ hidden classes. Each class $k$ has (a) a **size** $\pi_k$ (its share of the population) and (b) its own set of **item-endorsement probabilities** $\rho_{jk}$ - the chance a member of class $k$ says "yes" to item $j$. A person belongs to one class, but we do not see which; we see only their answers. The probability of a whole response pattern is a **mixture** - a weighted blend across the classes:

$$P(\mathbf{x}_i) = \sum_{k=1}^{K} \pi_k \prod_{j=1}^{J} \rho_{jk}^{x_{ij}} (1-\rho_{jk})^{1-x_{ij}}$$

The classes are the profiles $\rho_{\cdot k}$; membership is hidden. Our job is to recover both from the responses alone.

## Fitting It From Scratch with EM

The difficulty is a chicken-and-egg: if we knew who was in each class we could estimate the class profiles, and if we knew the profiles we could assign people to classes - but we know neither. The **EM algorithm** breaks the loop by alternating, exactly as [Rasch estimation](18-rasch.qmd) did:

- **E-step** (Expectation): given current profiles, compute each person's *posterior* probability of belonging to each class.
- **M-step** (Maximization): given those soft memberships, re-estimate each class's size and item probabilities as membership-weighted averages.

Repeat until nothing moves. First, simulate two classes that differ in *pattern* - items 7–8 are reversed, so the classes are not simply "more" vs. "less" symptomatic:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch22-lca`. For example, `book_data("ch22-lca")` in R, Julia, or Python, or `!bookdata name = "ch22-lca".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(2026)
N <- 800; J <- 8
prev <- c(0.35, 0.65)
rho  <- rbind(rep(0.80, J), rep(0.20, J))
rho[1, c(7, 8)] <- 0.30; rho[2, c(7, 8)] <- 0.70   # reverse two items: pattern, not severity

cls <- sample(1:2, N, replace = TRUE, prob = prev)
# one response vector per person; the estimator below wants them as a matrix
X <- do.call(rbind, map(cls, \(k) rbinom(J, 1, rho[k, ])))
```
### SPSS
```
* Two latent classes that differ in PATTERN: items 7-8 are reversed, so
* the classes are not simply "more" vs "less" symptomatic.
INPUT PROGRAM.
LOOP #i = 1 TO 800.
  COMPUTE cls = 1 + (RV.UNIFORM(0,1) < 0.65).
  VECTOR x(8).
  LOOP #j = 1 TO 8.
    DO IF (cls = 1).
      COMPUTE #p = 0.80.
      IF (#j >= 7) #p = 0.30.
    ELSE.
      COMPUTE #p = 0.20.
      IF (#j >= 7) #p = 0.70.
    END IF.
    COMPUTE x(#j) = RV.UNIFORM(0,1) < #p.
  END LOOP.
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.
```
### Julia
```julia
using Random, Distributions
Random.seed!(2026)
N, J = 800, 8
prev = [0.35, 0.65]
rho  = [fill(0.80, J), fill(0.20, J)]
rho[1][7:8] .= 0.30; rho[2][7:8] .= 0.70   # reverse two items

cls = rand(Categorical(prev), N)
X = [rand() < rho[cls[i]][j] ? 1 : 0 for i in 1:N, j in 1:J]
```
### Python
```python
import numpy as np

rng = np.random.default_rng(2026)
N, J = 800, 8
prev = [0.35, 0.65]
rho = np.vstack([np.full(J, 0.80), np.full(J, 0.20)])
rho[0, 6:8] = 0.30; rho[1, 6:8] = 0.70     # reverse two items

cls = rng.choice([0, 1], N, p=prev)
X = (rng.random((N, J)) < rho[cls]).astype(int)
```
:::

Now the entire estimator - about a dozen lines:

::: {.panel-tabset}
### R
```{r}
lca_em <- function(X, K = 2, tol = 1e-6, maxit = 500) {
  N <- nrow(X); J <- ncol(X)
  set.seed(1); pi <- rep(1/K, K); p <- matrix(runif(K * J, .2, .8), K, J)
  ll_old <- -Inf
  # EM is inherently sequential - each sweep starts from the previous estimates -
  # so this stays a loop. The per-class E-step inside it does not, and maps.
  for (it in 1:maxit) {
    # E-step: log P(x_i, class k) for every person and class, then normalize to a posterior
    loglik <- map(1:K, \(k)
      log(pi[k]) + X %*% log(p[k, ]) + (1 - X) %*% log(1 - p[k, ])) |>
      reduce(cbind)
    m    <- apply(loglik, 1, max)
    post <- exp(loglik - m); post <- post / rowSums(post)
    # M-step: re-estimate sizes and item probabilities as posterior-weighted averages
    pi <- colMeans(post)
    p  <- t(post) %*% X / colSums(post)
    ll <- sum(m + log(rowSums(exp(loglik - m))))
    if (abs(ll - ll_old) < tol) break; ll_old <- ll
  }
  list(prevalence = pi, item_prob = p, iterations = it)
}

fit <- lca_em(X, K = 2)
cat("converged in", fit$iterations, "iterations\n")

# recovered class sizes (true: .35 / .65)
tibble(class = seq_along(fit$prevalence), prevalence = fit$prevalence) |> round2()
```
### SPSS
```
* Base SPSS has no latent class analysis. The nearest built-in relative is
* TwoStep Cluster (CLUSTER / TWOSTEP), which groups cases but does not
* estimate class-conditional item probabilities or class membership
* posteriors the way an LCA does.
*
* A genuine LCA in the SPSS world needs the Latent GOLD plug-in, Mplus, or
* R's poLCA. The from-scratch EM estimator is shown in the R, Julia, and
* Python tabs.
TWOSTEP CLUSTER
  /CATEGORICAL VARIABLES=x1 x2 x3 x4 x5 x6 x7 x8
  /CRITERIA NUMCLUSTERS(2).
```
### Julia
```julia
using Statistics
function lca_em(X, K = 2; tol = 1e-6, maxit = 500)
    N, J = size(X)
    pi_ = fill(1/K, K)
    p   = rand(K, J) .* 0.6 .+ 0.2
    ll_old = -Inf
    # EM is sequential: each sweep starts from the previous estimates
    for it in 1:maxit
        loglik = [log(pi_[k]) + sum(X[i,j]*log(p[k,j]) + (1-X[i,j])*log(1-p[k,j])
                                    for j in 1:J) for i in 1:N, k in 1:K]
        m    = maximum(loglik, dims = 2)
        post = exp.(loglik .- m); post ./= sum(post, dims = 2)
        pi_  = vec(mean(post, dims = 1))
        p    = (post' * X) ./ sum(post, dims = 1)'
        ll   = sum(m .+ log.(sum(exp.(loglik .- m), dims = 2)))
        abs(ll - ll_old) < tol && return (prevalence = pi_, item_prob = p, iterations = it)
        ll_old = ll
    end
end

fit = lca_em(X, 2)
fit.prevalence                     # recovered class sizes (true: .35 / .65)
```
### Python
```python
import numpy as np

def lca_em(X, K=2, tol=1e-6, maxit=500):
    N, J = X.shape
    rng = np.random.default_rng(1)
    pi_ = np.full(K, 1 / K)
    p = rng.uniform(0.2, 0.8, (K, J))
    ll_old = -np.inf
    # EM is sequential: each sweep starts from the previous estimates
    for it in range(1, maxit + 1):
        loglik = (np.log(pi_) + X @ np.log(p).T + (1 - X) @ np.log(1 - p).T)
        m = loglik.max(axis=1, keepdims=True)
        post = np.exp(loglik - m); post /= post.sum(axis=1, keepdims=True)
        pi_ = post.mean(axis=0)
        p = (post.T @ X) / post.sum(axis=0)[:, None]
        ll = float((m + np.log(np.exp(loglik - m).sum(axis=1, keepdims=True))).sum())
        if abs(ll - ll_old) < tol:
            return dict(prevalence=pi_, item_prob=p, iterations=it)
        ll_old = ll

fit = lca_em(X, 2)
fit["prevalence"]                  # recovered class sizes (true: .35 / .65)
```
:::

The class sizes come back essentially exact. And the item-probability profiles - the *definitions* of the two kinds - recover the reversal we planted:

::: {.panel-tabset}
### R
```{r}
# rows = classes, columns = items 1..8
fit$item_prob |>
  as_tibble(.name_repair = \(x) paste0("item", seq_along(x))) |>
  mutate(class = row_number(), .before = 1) |>
  round2()
```
### SPSS
```
* The item-probability profiles - the DEFINITIONS of the two kinds - are
* what an LCA reports and TwoStep Cluster does not. See the note in the
* previous SPSS tab; use Latent GOLD, Mplus, or R's poLCA for these.
*
* The closest built-in summary is the endorsement rate of each item within
* each cluster:
MEANS TABLES=x1 x2 x3 x4 x5 x6 x7 x8 BY TSC_Predicted
  /CELLS=MEAN.
```
### Julia
```julia
fit.item_prob                      # rows = classes, columns = items 1..8
```
### Python
```python
fit["item_prob"].round(2)          # rows = classes, columns = items 1..8
```
:::

Read the two rows: one class endorses items 1–6 heavily but items 7–8 rarely; the other is its mirror image. **These classes are not "high" and "low" depression - they are two different *shapes* of response.** A sum score, which only counts yeses, would blend them together; the mixture model pulls them apart. That is what LCA is *for*.

## How Many Classes?

We told the estimator $K = 2$, but in real life the number of classes is itself unknown - and it is the crux of the analysis. You fit $K = 1, 2, 3, \dots$ and compare, usually with an information criterion (**BIC** or **AIC**) that rewards fit and penalizes complexity, plus interpretability (a class you cannot describe is probably an artifact). In practice you would use a maintained tool that also handles this:

```r
# The real tool (needs the poLCA package — shown, not run on the book's server)
library(poLCA)
f  <- cbind(i1, i2, i3, i4, i5, i6, i7, i8) ~ 1
m2 <- poLCA(f, data, nclass = 2)      # fit 2, 3, 4 classes ...
m3 <- poLCA(f, data, nclass = 3)      # ... and compare
c(m2$bic, m3$bic)                     # lower BIC = better; also judge interpretability
```

::: {.callout-note}
## From the Field
Kashdan, McKnight, and colleagues [@KashdanMcKnightRichey2009] applied latent-class analysis to the National Comorbidity Survey and found a **neglected subtype** of social-anxiety disorder: people who are anxious *and* risk-prone/approach-oriented, not the shy-avoidant stereotype. That subgroup is invisible to a symptom count - you can only see it once you stop asking "how much?" and start asking "what kind?" It is the same lesson as the [antitypes case study](24-casestudy-depression.qmd): structure that a total score throws away can be the most clinically important thing in the data.
:::

## What This Chapter Teaches

- **Dimensions vs. types.** Factor analysis places people on continua; latent-class analysis sorts them into kinds. Ask which question your science is really asking.
- **A latent class model is a mixture** of response-pattern probabilities - classes you cannot see, recovered from responses you can.
- **EM solves the chicken-and-egg** by alternating soft assignment (E) and re-estimation (M), the same iterate-until-stable trick behind Rasch.
- **Classes can differ in pattern, not just level** - and those pattern-defined subgroups are exactly what a sum score is built to hide.

We have now met latent variables as *dimensions* (factors) and as *categories* (classes). The [next chapter](30-irt.qmd) returns to the dimensional view but drills deeper into it - extending the [Rasch model](18-rasch.qmd) into the wider world of item response theory and generalizability, where the authors have published extensively.
