# More Item Response Theory, and Generalizability {#irt}

The [Rasch chapter](18-rasch.qmd) built one item response model from scratch and promised there was a wider world. Here it is. Item Response Theory (IRT) is a whole family of models for the same task - measuring people from their item responses - and **generalizability theory (G-theory)** is a broad extension of [reliability](10-reliability.qmd) that asks not only "how much error?" but "error *from what*?" Both are areas where the authors have published extensively, often together in a single study.

## Learning Objectives

1.  See how the **2-parameter (2PL)** model generalizes Rasch by letting items differ in *discrimination*.
2.  Read an **item characteristic curve** and know what its slope and location mean.
3.  Decompose measurement error into **facets** with generalizability theory.
4.  Compute a **G-coefficient** and use a **D-study** to design a measure - the "how many items?" question.

## Beyond Rasch: Items That Discriminate Differently

Rasch makes a strong, simple assumption: every item is equally good at separating people; items differ *only* in difficulty. The **2PL model** relaxes it, giving each item its own **discrimination** $a_i$ (how sharply it distinguishes high from low) alongside its difficulty $b_i$:

$$P(x_{ij}=1) = \frac{1}{1 + e^{-a_i(\theta_j - b_i)}}$$

The clearest way to feel the difference is to *look* at the items. An item's **characteristic curve** plots the probability of a correct/endorsed response against ability $\theta$. Difficulty slides the curve left-right; discrimination changes its steepness:

```{r}
#| fig-cap: "Item characteristic curves. Location = difficulty; steepness = discrimination."
#| fig-height: 3.8
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
library(broom)        # tidy(), glance(), augment() for model output
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

icc <- function(theta, a, b) 1 / (1 + exp(-a * (theta - b)))   # 2PL curve

items <- tibble(
  item = c("easy, moderate a", "medium, high a (steep)", "hard, low a (flat)"),
  a    = c(1.0, 2.5, 0.5),      # discrimination (steepness)
  b    = c(-1,  0.0, 1.0)       # difficulty (left-right location)
)

crossing(theta = seq(-4, 4, length.out = 300), items) |>
  mutate(p = icc(theta, a, b)) |>
  ggplot(aes(x = theta, y = p, colour = item)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(
    name   = NULL,
    values = c("easy, moderate a"       = "steelblue",
               "medium, high a (steep)" = "firebrick",
               "hard, low a (flat)"     = "darkgreen")
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = expression(theta ~ "(ability)"), y = "P(correct)") +
  theme_minimal() +
  theme(legend.position = "top")
```

The steep red item is *discriminating*: over a narrow band of ability it flips from "probably wrong" to "probably right," so it sorts people sharply there. The flat green item barely distinguishes anyone - its responses are nearly a coin flip across a wide ability range. **Rasch is the special case where every curve has the same slope**; the 2PL, and its polytomous cousin the **Graded Response Model** (which we met dissecting [depression profiles](24-casestudy-depression.qmd)), let the data say otherwise. In practice you fit these with a maintained engine:

```r
library(mirt)                              # shown, not run on the book's server
fit  <- mirt(data, 1, itemtype = "2PL")    # or "graded" for 0..k items (GRM)
coef(fit, simplify = TRUE)                 # each item's discrimination (a) and difficulty (b)
```

## Generalizability: Reliability, Grown Up

Classical reliability ([Ch. 10](10-reliability.qmd)) splits observed variance into "true" and "error." But error from where? It could come from the particular item you used, the rater who scored you, or the day you happened to show up. **Generalizability theory** treats each of these as a **facet** and estimates how much variance each contributes, turning "reliability" from a single number into an analysis of *sources*.

The machinery is the [variance-partitioning](04-Dmeasures.qmd) you already know, run with `aov`. Simulate people crossed with items, then recover how much of the variance is real person differences versus item and residual noise:

::: {.panel-tabset}
### R
```{r}
set.seed(2026)
np <- 120; ni <- 12
person <- rnorm(np, sd = 1.0)   # real differences between people
item   <- rnorm(ni, sd = 0.5)   # items differ in easiness

d <- expand_grid(p = 1:np, i = 1:ni) |>
  mutate(
    score = person[p] + item[i] + rnorm(np * ni, sd = 0.7),
    p     = factor(p),
    i     = factor(i)
  )

ms  <- aov(score ~ p + i, data = d) |> tidy() |> pull(meansq)   # mean squares
v_e <- ms[3]                    # residual variance
v_p <- (ms[1] - v_e) / ni       # person variance (the signal)
v_i <- (ms[2] - v_e) / np       # item variance

tibble(person = v_p, item = v_i, residual = v_e) |> round2()
```
### SPSS
```
* Data assumed present as the active dataset with columns score, p (person), i (item).
* VARCOMP returns the person, item, and residual variance components directly,
* no hand EMS math from mean squares.
VARCOMP score BY p i
  /RANDOM = p i
  /METHOD = REML.
```
### Julia
```julia
using MixedModels

# df is a DataFrame with columns score, p (person), i (item).
# The random-effect variances ARE the G-theory person/item components;
# the residual variance is the leftover noise.
fit(MixedModel, @formula(score ~ 1 + (1|p) + (1|i)), df)
```
### Python
```python
import numpy as np
import statsmodels.formula.api as smf

# df has columns score, p (person), i (item).
# Persons and items are crossed random effects; their variances ARE the
# G-theory components, and the model's scale is the residual variance.
vc = {"p": "0 + C(p)", "i": "0 + C(i)"}
smf.mixedlm("score ~ 1", df, groups=np.ones(len(df)), vc_formula=vc).fit().summary()
```
:::

Person variance is the **signal** - the real differences you want to measure. Item and residual variance are **noise**, now itemized by source. The **generalizability coefficient** is the same signal-share idea as reliability, but honest about the fact that you averaged over a *sample* of items:

```{r}
# G-coefficient for a test of ni items
tibble(n_items = ni, G = v_p / (v_p + v_e / ni)) |> round2()
```

### The D-study: Designing the Measure

This is where G-theory pays for itself. Because you know the variance components, you can ask a **design** question - a *decision (D) study*: how many items would I need for a target reliability? It is the [power analysis](09-statisticalpower.qmd) of measurement.

```{r}
tibble(n_items = c(4, 8, 12, 20, 40)) |>
  mutate(G = v_p / (v_p + v_e / n_items)) |>   # generalizability at each test length
  round2()
```

Reliability climbs with test length exactly as the [Spearman-Brown](10-reliability.qmd) logic predicts - and now you can pick the length that buys the reliability you need and stop, instead of guessing. Add a second facet (raters, occasions) and G-theory tells you whether your money is better spent on more items or more raters. That is measurement design as engineering.

::: {.callout-note}
## From the Field
This is the authors' bread and butter. McKnight and Babcock-Parziale [@McKnightBabcock2007] combined **Rasch and generalizability theory in a single study** of a vision-rehabilitation outcome - exactly the two halves of this chapter, together. The Rasch analysis of the Mississippi PTSD Scale with Ben Wright [@ConradWrightMcKnight2004] showed reverse-scored items misbehaving; and Stroud, McKnight, and Jensen [@StroudMcKnightJensen2004] used IRT to *shorten* a disability scale without losing information - a D-study decision in action. The [methods showcase](27-methods-showcase.qmd) has the full list.
:::

## What This Chapter Teaches

- **IRT is a family.** Rasch fixes item discrimination; the 2PL frees it; the GRM extends it to graded items. Read the **item characteristic curve** - location is difficulty, slope is discrimination.
- **G-theory generalizes reliability** by decomposing error into named facets (items, raters, occasions) with the same variance-partitioning you use for ANOVA.
- **The D-study is power analysis for measurement**: knowing the variance components, you design the test - how many items, how many raters - to hit a target reliability.

We have now seen latent variables from every angle the authors work in - factors, classes, and item-response continua. That completes our tour of *modeling* the things we measure. But building a model raises a question models cannot answer on their own: which variables *belong* in it, and what a coefficient means causally. The [next part](33-causal.qmd) takes that up directly, with directed acyclic graphs and the logic of causal inference.
