# Causal Inference with DAGs {#causal}

Regression can tell you that two variables are associated. It cannot, on its own, tell you whether one *causes* the other, or which variables you should control for to find out. Those are causal questions, and causal questions need a causal model. The clearest way to state one is the **directed acyclic graph (DAG)**: a diagram of what you assume causes what. This is the framework Richard McElreath builds his *Statistical Rethinking* around, and the one we teach in our own Bayesian course. It pairs naturally with everything you have learned about [regression](15-MRC.qmd) and [structural models](28-sem.qmd), because it answers the question those tools cannot: *which variables belong in the model?*

## Learning Objectives

1.  Read a DAG as a set of causal assumptions, and see why it decides what you control for.
2.  Recognize the four elemental confounds: the fork, the pipe, the collider, and the descendant.
3.  Use the back-door criterion to choose an adjustment set.
4.  Know that a DAG makes *testable* predictions.

## A DAG Is Your Causal Theory, Drawn

A DAG is variables (nodes) joined by arrows. An arrow $X \rightarrow Y$ means "$X$ is a direct cause of $Y$." *Acyclic* means no loops: nothing causes itself, directly or through a chain. The graph has no numbers; it is the causal *theory*, and it does something a regression cannot: it tells you which variables to include and which to leave out. Two analysts with identical data but different DAGs will, correctly, control for different things.

Take McElreath's running example. Does a state's marriage rate ($M$) *cause* its divorce rate ($D$)? Both are correlated with the median age at marriage ($A$). One plausible DAG is $A \rightarrow M$, $A \rightarrow D$, and $M \rightarrow D$. Is the $M$–$D$ association a real causal effect, or just a shadow cast by $A$? A regression of $D$ on both $M$ and $A$ can separate them, but *only because the DAG told you to include $A$*. Without the causal picture, you would not know which variable to condition on, and the data alone will never tell you.

## The Four Elemental Confounds

Every DAG, however complicated, is built from four elementary pieces. Knowing how each behaves is most of causal inference (McElreath). Three of them we can watch directly.

**The Fork** ($X \leftarrow Z \rightarrow Y$). $Z$ is a common cause, a **confounder**. $X$ and $Y$ become correlated even with no arrow between them. **Control $Z$** to close the path:

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `ch24-fork`, `ch24-pipe`, `ch24-collider`. For example, `book_data("ch24-fork")` in R, Julia, or Python, or `!bookdata name = "ch24-fork".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.
:::

::: {.panel-tabset}
### R
```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
library(broom)        # tidy(), glance(), augment() for model output
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(4); n <- 2000
fork <- tibble(Z = rnorm(n)) |>              # Z causes both; no X->Y arrow
  mutate(X = Z + rnorm(n), Y = Z + rnorm(n))

fork |>
  summarise(
    naive   = cor(X, Y),                     # spurious association
    control = cor(resid(lm(X ~ Z)), resid(lm(Y ~ Z)))   # control Z: it vanishes
  ) |>
  round2()
```
### SPSS
```
* The Fork: Z causes both X and Y, so X and Y correlate with no arrow
* between them. Controlling Z makes the association vanish.
INPUT PROGRAM.
LOOP #i = 1 TO 2000.
  COMPUTE Z = RV.NORMAL(0, 1).
  COMPUTE X = Z + RV.NORMAL(0, 1).
  COMPUTE Y = Z + RV.NORMAL(0, 1).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

* Naive association.
CORRELATIONS /VARIABLES=X Y.
* Partial correlation controlling Z - the same thing as correlating residuals.
PARTIAL CORR /VARIABLES=X Y BY Z.
```
### Julia
```julia
using Random, Distributions, Statistics, GLM, DataFrames
Random.seed!(4)
n = 2000
fork = DataFrame(Z = rand(Normal(), n))
fork.X = fork.Z .+ rand(Normal(), n)
fork.Y = fork.Z .+ rand(Normal(), n)

cor(fork.X, fork.Y)                                  # spurious association
cor(residuals(lm(@formula(X ~ Z), fork)),
    residuals(lm(@formula(Y ~ Z), fork)))            # control Z: it vanishes
```
### Python
```python
import numpy as np, pandas as pd
import statsmodels.formula.api as smf

rng = np.random.default_rng(4)
n = 2000
fork = pd.DataFrame({"Z": rng.normal(size=n)})
fork["X"] = fork.Z + rng.normal(size=n)
fork["Y"] = fork.Z + rng.normal(size=n)

fork.X.corr(fork.Y)                                  # spurious association
np.corrcoef(smf.ols("X ~ Z", fork).fit().resid,
            smf.ols("Y ~ Z", fork).fit().resid)[0, 1]   # control Z: it vanishes
```
:::

**The Pipe** ($X \rightarrow M \rightarrow Y$). $M$ is a **mediator**; $X$ affects $Y$ *through* $M$. Control $M$ and you block the very effect you were trying to measure (this is *post-treatment bias*):

::: {.panel-tabset}
### R
```{r}
set.seed(5)
pipe <- tibble(X = rnorm(n)) |>              # the effect runs X -> M -> Y
  mutate(M = X + rnorm(n), Y = M + rnorm(n))

pipe |>
  summarise(
    naive   = cor(X, Y),                     # the real, total effect
    control = cor(resid(lm(X ~ M)), resid(lm(Y ~ M)))   # control M: effect blocked
  ) |>
  round2()
```
### SPSS
```
* The Pipe: the effect runs X -> M -> Y. Controlling M blocks the very
* effect you were trying to measure (post-treatment bias).
INPUT PROGRAM.
LOOP #i = 1 TO 2000.
  COMPUTE X = RV.NORMAL(0, 1).
  COMPUTE M = X + RV.NORMAL(0, 1).
  COMPUTE Y = M + RV.NORMAL(0, 1).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

CORRELATIONS /VARIABLES=X Y.
PARTIAL CORR /VARIABLES=X Y BY M.
```
### Julia
```julia
Random.seed!(5)
pipe = DataFrame(X = rand(Normal(), n))
pipe.M = pipe.X .+ rand(Normal(), n)
pipe.Y = pipe.M .+ rand(Normal(), n)

cor(pipe.X, pipe.Y)                                  # the real, total effect
cor(residuals(lm(@formula(X ~ M), pipe)),
    residuals(lm(@formula(Y ~ M), pipe)))            # control M: effect blocked
```
### Python
```python
rng = np.random.default_rng(5)
pipe = pd.DataFrame({"X": rng.normal(size=n)})
pipe["M"] = pipe.X + rng.normal(size=n)
pipe["Y"] = pipe.M + rng.normal(size=n)

pipe.X.corr(pipe.Y)                                  # the real, total effect
np.corrcoef(smf.ols("X ~ M", pipe).fit().resid,
            smf.ols("Y ~ M", pipe).fit().resid)[0, 1]   # control M: blocked
```
:::

**The Collider** ($X \rightarrow Z \leftarrow Y$). $Z$ is a common *effect*. Here $X$ and $Y$ are genuinely unrelated, but conditioning on $Z$ **creates** an association that was never there:

::: {.panel-tabset}
### R
```{r}
set.seed(6)
collider <- tibble(X = rnorm(n), Y = rnorm(n)) |>   # X and Y both cause Z
  mutate(Z = X + Y + rnorm(n))

collider |>
  summarise(
    naive   = cor(X, Y),                            # correctly ~ 0
    control = cor(resid(lm(X ~ Z)), resid(lm(Y ~ Z)))   # control Z: spurious!
  ) |>
  round2()
```
### SPSS
```
* The Collider: X and Y both cause Z and are genuinely unrelated.
* Conditioning on Z CREATES an association that was never there.
INPUT PROGRAM.
LOOP #i = 1 TO 2000.
  COMPUTE X = RV.NORMAL(0, 1).
  COMPUTE Y = RV.NORMAL(0, 1).
  COMPUTE Z = X + Y + RV.NORMAL(0, 1).
  END CASE.
END LOOP.
END FILE.
END INPUT PROGRAM.
EXECUTE.

CORRELATIONS /VARIABLES=X Y.
PARTIAL CORR /VARIABLES=X Y BY Z.
```
### Julia
```julia
Random.seed!(6)
collider = DataFrame(X = rand(Normal(), n), Y = rand(Normal(), n))
collider.Z = collider.X .+ collider.Y .+ rand(Normal(), n)

cor(collider.X, collider.Y)                          # correctly ~ 0
cor(residuals(lm(@formula(X ~ Z), collider)),
    residuals(lm(@formula(Y ~ Z), collider)))        # control Z: spurious!
```
### Python
```python
rng = np.random.default_rng(6)
collider = pd.DataFrame({"X": rng.normal(size=n), "Y": rng.normal(size=n)})
collider["Z"] = collider.X + collider.Y + rng.normal(size=n)

collider.X.corr(collider.Y)                          # correctly ~ 0
np.corrcoef(smf.ols("X ~ Z", collider).fit().resid,
            smf.ols("Y ~ Z", collider).fit().resid)[0, 1]   # control Z: spurious!
```
:::

**The Descendant.** Conditioning on a variable *caused by* another partly conditions on that other variable. A descendant of a collider re-opens the collider a little; a descendant of a confounder closes the fork a little. The rule of thumb: a descendant behaves like a weak version of its parent.

::: {.callout-important}
## The one table to remember

Whether controlling for a variable *helps or hurts* depends entirely on its role in the DAG, not on anything in the data:

- **Confounder (fork):** control it. Not controlling it leaves a spurious association.
- **Mediator (pipe):** do *not* control it if you want the total effect. Controlling it blocks the effect.
- **Collider:** *never* control it. Controlling it manufactures a spurious association.
- **Descendant:** treat it as a weaker version of whatever it descends from.

No fit statistic can tell these apart. The difference is entirely in the direction of the arrows, which the data cannot reveal. That is why you need the DAG.
:::

## The Back-Door Criterion

Given a DAG, *which* variables must you control to estimate the causal effect of $X$ on $Y$? The **back-door criterion** answers it: block every "back-door path" - every non-causal path from $X$ to $Y$ that begins with an arrow *into* $X$ - by controlling confounders, while being careful not to open any colliders. For anything but the simplest graph this is easy to get wrong by hand, so the `dagitty` package (which the course uses) computes the adjustment set for you:

```r
# The real tool (needs dagitty — shown, not run on the book's server)
library(dagitty)
dag <- dagitty("dag{ X -> Y ; X <- Z -> Y }")     # a fork: Z confounds X and Y
adjustmentSets(dag, exposure = "X", outcome = "Y")   # -> { Z }: control Z, nothing else
```

The output is the set of variables to put in your regression. Sometimes it is empty (no adjustment needed); sometimes there is more than one valid set; sometimes it warns you that the effect is *not identifiable* from the data you have. All of that is causal information you cannot get from the numbers alone.

## A DAG Makes Testable Predictions

A DAG is an assumption, but not an untestable one. Every DAG implies a set of **conditional independencies** - statements like "$A$ and $B$ should be independent once we account for $C$" - that the data can confirm or refute:

```r
impliedConditionalIndependencies(dag)   # the testable implications of your DAG
```

If the data violate an independence your DAG predicts, the DAG is wrong and needs revising. This is how a causal model earns trust: not by fitting the data (any flexible model can do that), but by surviving the specific predictions it makes. It is the same discipline the [Sports Page chapter](23-sportspage.qmd) preached, applied to causal structure.

## DAGs and SEM: Two Notations, One Idea

Coming from psychology, we will let you in on something: a DAG is the *same causal object* as a [structural equation model](28-sem.qmd). The arrows are identical; only the notation differs. One of us cheerfully admits to being more fluent in standard SEM path-diagram notation than in "the rather sparse DAG notation" - but the causal logic underneath is one and the same. An SEM is a DAG with its arrows *estimated*; a DAG is an SEM *before* you attach the numbers. So the workflow is always the same two steps: **draw the DAG** to decide the structure (which paths exist, which variables to control), then **fit the model** to put numbers on it. And carry the caution from the mediation work: fitting never *verifies* the causal story. The perennial temptation, in the authors' own words, is to "assume causality but rely on correlation." The DAG is what keeps you honest about the difference.

## What This Chapter Teaches

- A **DAG** states your causal assumptions and, from them, tells you which variables to control - something no regression can decide on its own.
- The **four elemental confounds**: control a fork, leave a pipe alone (for the total effect), never control a collider, and treat a descendant as a weak version of its parent.
- The **back-door criterion** (via `dagitty::adjustmentSets`) picks the variables to adjust for.
- A DAG makes **testable conditional-independence predictions**; a good causal model survives them.
- A DAG and an SEM are the **same causal object** in different notation: draw the graph, then fit the model.

## Where We Go Next

DAGs are one half of the *Statistical Rethinking* worldview the authors teach; the other half is Bayesian. Having drawn our causal assumptions, we turn to a different way of quantifying uncertainty about them. The [next part](20-beyond.qmd) steps beyond the normal-curve, point-estimate, reject-or-accept world into resampling, robust methods, and Bayesian inference - the framework in which McElreath's DAGs are usually put to work.
