# The Same Analysis in R, SPSS, Julia, and Python {#three-languages .unnumbered}

This book runs on **R**, but the ideas are not R's property. A correlation is a correlation, a regression is a regression, whatever software you type it into. This appendix shows the core procedures from the book side by side in four environments: **R**, **SPSS**, **Julia**, and **Python**. Click the tabs to switch languages. The R code is *live* - it runs here and produces the output you see - and the SPSS, Julia, and Python versions produce the same result, written the way each language likes to write it.

We use one small dataset throughout: 80 people in two groups, with a predictor `x` and an outcome `y`.

```{r}
#| message: false
library(tidyverse)    # dplyr, ggplot2, purrr, tibble, readr, stringr, forcats
library(broom)        # tidy(), glance(), augment() for model output
source("_common.R")   # book-wide helpers: round2(), fmt_p(), tidy2()

set.seed(1)
d <- tibble(
  group = factor(rep(c("A", "B"), each = 40)),
  x     = rnorm(80, mean = 50, sd = 10)
) |>
  mutate(y = 100 + 0.8 * x + rnorm(80, sd = 12))   # y depends on x

d |> head(3) |> round2()
```

In SPSS, imagine this same table loaded as the active dataset with variables `group`, `x`, and `y`. In Julia, imagine it as a `DataFrame` called `d`. In Python, imagine it as a pandas `DataFrame`, also called `d`.

## Descriptive Statistics

::: {.callout-note}
## Working in SPSS, Julia, or Python?
The code tabs below assume this chapter's data is already loaded. Grab the one-file setup for your language from [Getting the Book's Data](setup.qmd), run it once, then load what you need by name - this chapter uses `appendix-d`, `appendix-items`. For example, `book_data("appendix-d")` in R, Julia, or Python, or `!bookdata name = "appendix-d".` in SPSS. Every language reads the same shipped files, so your numbers will match the ones printed here exactly.
:::

::: {.panel-tabset}
### R
```{r}
d |>
  select(x, y) |>
  pivot_longer(everything(), names_to = "variable") |>
  summarise(mean = mean(value), sd = sd(value),
            min = min(value), max = max(value), .by = variable) |>
  round2()
```
### SPSS
```
DESCRIPTIVES VARIABLES=x y
  /STATISTICS=MEAN STDDEV MIN MAX.
```
### Julia
```julia
using Statistics, DataFrames
describe(d[:, [:x, :y]], :mean, :std, :min, :max)
```
### Python
```python
d[["x", "y"]].describe().loc[["mean", "std", "min", "max"]]
```
:::

## Correlation

::: {.panel-tabset}
### R
```{r}
d |> summarise(r = cor(x, y)) |> round2()
```
### SPSS
```
CORRELATIONS
  /VARIABLES=x y.
```
### Julia
```julia
using Statistics
cor(d.x, d.y)
```
### Python
```python
d.x.corr(d.y)
```
:::

## Comparing Two Groups (t-test)

::: {.panel-tabset}
### R
```{r}
t.test(y ~ group, data = d) |>
  tidy() |>
  mutate(p.value = fmt_p(p.value)) |>
  round2()
```
### SPSS
```
T-TEST GROUPS=group('A' 'B')
  /VARIABLES=y.
```
### Julia
```julia
using HypothesisTests
UnequalVarianceTTest(d.y[d.group .== "A"], d.y[d.group .== "B"])
```
### Python
```python
from scipy import stats
stats.ttest_ind(d.y[d.group == "A"], d.y[d.group == "B"], equal_var=False)
```
:::

## Linear Regression

::: {.panel-tabset}
### R
```{r}
tidy2(lm(y ~ x, data = d))
```
### SPSS
```
REGRESSION
  /DEPENDENT y
  /METHOD=ENTER x.
```
### Julia
```julia
using GLM, DataFrames
coeftable(lm(@formula(y ~ x), d))
```
### Python
```python
import statsmodels.formula.api as smf
smf.ols("y ~ x", data=d).fit().summary()
```
:::

## One-Way ANOVA

::: {.panel-tabset}
### R
```{r}
aov(y ~ group, data = d) |>
  tidy() |>
  mutate(p.value = fmt_p(p.value)) |>
  round2()
```
### SPSS
```
ONEWAY y BY group
  /STATISTICS DESCRIPTIVES.
```
### Julia
```julia
using GLM
m = lm(@formula(y ~ group), d)
ftest(m.model)          # overall F test for the group effect
```
### Python
```python
import statsmodels.formula.api as smf
from statsmodels.stats.anova import anova_lm
anova_lm(smf.ols("y ~ group", data=d).fit())   # overall F test for the group effect
```
:::

## Principal Components

::: {.panel-tabset}
### R
```{r}
# a small item set for the reduction example
set.seed(2)
it <- map(1:4, \(i) rnorm(80) + d$x / 10) |>
  set_names(paste0("item", 1:4)) |>
  as_tibble()

tibble(component  = paste0("PC", 1:4),
       eigenvalue = prcomp(it, scale. = TRUE)$sdev^2) |>
  round2()
```
### SPSS
```
FACTOR
  /VARIABLES x1 x2 x3 x4
  /EXTRACTION PC
  /ROTATION VARIMAX.
```
### Julia
```julia
using MultivariateStats
M = fit(PCA, Matrix(it)'; maxoutdim = 4)
principalvars(M)        # variance along each component
```
### Python
```python
from sklearn.decomposition import PCA
from sklearn.preprocessing import StandardScaler
z = StandardScaler().fit_transform(it)
PCA().fit(z).explained_variance_    # variance along each component
```
:::

## A Note on the Four Environments

- **R** is free, open source, and the language this book runs in. Everything you see executes.
- **SPSS** is menu-driven, but every menu click also writes *syntax* - the code shown above. Pasting that syntax into a Syntax window and running it is the reproducible way to use SPSS, and it reads much like the R.
- **Julia** is free and open source, fast, and increasingly used for heavy Bayesian and simulation work. Its statistics live in packages (`Statistics`, `DataFrames`, `GLM`, `HypothesisTests`, `MultivariateStats`), loaded with `using`.
- **Python** is free and open source, and probably the language your students already know. Data lives in a `pandas` DataFrame; the models come from `statsmodels`, whose formula interface (`smf.ols("y ~ x", data=d)`) reads almost exactly like R's. `scipy.stats` supplies the classical tests and `scikit-learn` the reduction methods.

The point is not that one is best. It is that a procedure you understand in one language you can carry to any of them, because the statistics were never about the software.
