# The Same Analysis in R, SPSS, and Julia {#three-languages .unnumbered}

This book runs on **R**, but the ideas are not R's property. A correlation is a correlation, a regression is a regression, whatever software you type it into. This appendix shows the core procedures from the book side by side in three environments: **R**, **SPSS**, and **Julia**. Click the tabs to switch languages. The R code is *live* - it runs here and produces the output you see - and the SPSS and Julia versions produce the same result, written the way each language likes to write it.

We use one small dataset throughout: 80 people in two groups, with a predictor `x` and an outcome `y`.

```{r}
set.seed(1)
d <- data.frame(
  group = factor(rep(c("A", "B"), each = 40)),
  x     = rnorm(80, mean = 50, sd = 10)
)
d$y <- 100 + 0.8 * d$x + rnorm(80, sd = 12)   # y depends on x
head(d, 3)
```

In SPSS, imagine this same table loaded as the active dataset with variables `group`, `x`, and `y`. In Julia, imagine it as a `DataFrame` called `d`.

## Descriptive Statistics

::: {.panel-tabset}
### R
```{r}
sapply(d[c("x", "y")], function(v) c(mean = mean(v), sd = sd(v)))
```
### SPSS
```
DESCRIPTIVES VARIABLES=x y
  /STATISTICS=MEAN STDDEV MIN MAX.
```
### Julia
```julia
using Statistics, DataFrames
describe(d[:, [:x, :y]], :mean, :std, :min, :max)
```
:::

## Correlation

::: {.panel-tabset}
### R
```{r}
cor(d$x, d$y)
```
### SPSS
```
CORRELATIONS
  /VARIABLES=x y.
```
### Julia
```julia
using Statistics
cor(d.x, d.y)
```
:::

## Comparing Two Groups (t-test)

::: {.panel-tabset}
### R
```{r}
t.test(y ~ group, data = d)
```
### SPSS
```
T-TEST GROUPS=group('A' 'B')
  /VARIABLES=y.
```
### Julia
```julia
using HypothesisTests
UnequalVarianceTTest(d.y[d.group .== "A"], d.y[d.group .== "B"])
```
:::

## Linear Regression

::: {.panel-tabset}
### R
```{r}
summary(lm(y ~ x, data = d))$coefficients
```
### SPSS
```
REGRESSION
  /DEPENDENT y
  /METHOD=ENTER x.
```
### Julia
```julia
using GLM, DataFrames
coeftable(lm(@formula(y ~ x), d))
```
:::

## One-Way ANOVA

::: {.panel-tabset}
### R
```{r}
summary(aov(y ~ group, data = d))
```
### SPSS
```
ONEWAY y BY group
  /STATISTICS DESCRIPTIVES.
```
### Julia
```julia
using GLM
m = lm(@formula(y ~ group), d)
ftest(m.model)          # overall F test for the group effect
```
:::

## Principal Components

::: {.panel-tabset}
### R
```{r}
# a small item matrix for the reduction example
set.seed(2); it <- sapply(1:4, function(i) rnorm(80) + d$x/10)
round(prcomp(it, scale. = TRUE)$sdev^2, 2)   # eigenvalues
```
### SPSS
```
FACTOR
  /VARIABLES x1 x2 x3 x4
  /EXTRACTION PC
  /ROTATION VARIMAX.
```
### Julia
```julia
using MultivariateStats
M = fit(PCA, Matrix(it)'; maxoutdim = 4)
principalvars(M)        # variance along each component
```
:::

## A Note on the Three Environments

- **R** is free, open source, and the language this book runs in. Everything you see executes.
- **SPSS** is menu-driven, but every menu click also writes *syntax* - the code shown above. Pasting that syntax into a Syntax window and running it is the reproducible way to use SPSS, and it reads much like the R.
- **Julia** is free and open source, fast, and increasingly used for heavy Bayesian and simulation work. Its statistics live in packages (`Statistics`, `DataFrames`, `GLM`, `HypothesisTests`, `MultivariateStats`), loaded with `using`.

The point is not that one is best. It is that a procedure you understand in one language you can carry to any of them, because the statistics were never about the software.
